---
title: "Phase 2 - Character Frame Review"
subtitle: "Review notebook for the modular Phase 2 Character build"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 2
    code-fold: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)

options(
  dplyr.summarise.inform = FALSE,
  knitr.kable.NA = ""
)

library(here)
library(dplyr)
library(readr)
library(arrow)
library(ggplot2)
library(scales)
library(knitr)

output_dir <- here("exploration/intelligence_framework/phase_2_character_calibration", "outputs")

cluster_calibration <- readr::read_csv(file.path(output_dir, "character_phase2_cluster_count_calibration.csv"), show_col_types = FALSE)
cluster_sizes <- readr::read_csv(file.path(output_dir, "character_phase2_cluster_count_sizes.csv"), show_col_types = FALSE)
cluster_centroids <- readr::read_csv(file.path(output_dir, "character_phase2_cluster_centroids.csv"), show_col_types = FALSE)
cluster_representatives <- readr::read_csv(file.path(output_dir, "character_phase2_cluster_representatives.csv"), show_col_types = FALSE)
cluster_kpi_averages <- readr::read_csv(file.path(output_dir, "character_phase2_cluster_kpi_averages.csv"), show_col_types = FALSE)
cluster_metric_extremes <- readr::read_csv(file.path(output_dir, "character_phase2_cluster_metric_extremes.csv"), show_col_types = FALSE)
gmm_summary <- readr::read_csv(file.path(output_dir, "character_phase2_gmm_summary.csv"), show_col_types = FALSE)
gmm_hybrids <- readr::read_csv(file.path(output_dir, "character_phase2_gmm_hybrids.csv"), show_col_types = FALSE)
similarity_top10 <- readr::read_csv(file.path(output_dir, "character_phase2_similarity_top10.csv"), show_col_types = FALSE)
size_dominance_summary <- readr::read_csv(file.path(output_dir, "character_phase2_size_dominance_summary.csv"), show_col_types = FALSE)
size_dominance_scatter <- readr::read_csv(file.path(output_dir, "character_phase2_size_dominance_scatter.csv"), show_col_types = FALSE)
connectedness_shift_summary <- readr::read_csv(file.path(output_dir, "character_phase2_connectedness_cluster_shift_summary.csv"), show_col_types = FALSE)
connectedness_shift_outliers <- readr::read_csv(file.path(output_dir, "character_phase2_connectedness_cluster_shift_outliers.csv"), show_col_types = FALSE)
character_scores <- arrow::read_parquet(file.path(output_dir, "character_scores.parquet"))

selected_k <- 7
selected_cluster_sizes <- cluster_sizes |>
  filter(method == "kmeans", k == selected_k) |>
  arrange(cluster)

model_summary <- tibble::tibble(
  modeling_universe = nrow(character_scores),
  selected_k = selected_k,
  kmeans_silhouette_at_k7 = cluster_calibration |>
    filter(k == selected_k) |>
    pull(kmeans_avg_silhouette),
  size_correlation = size_dominance_summary$raw_correlation_log_pop,
  density_correlation = size_dominance_summary$raw_correlation_density,
  connectedness_shift_share = connectedness_shift_summary$share_changed
)
```

## Purpose

This notebook is the review layer for Phase 2, not the build pipeline.

The canonical build now lives in:

- `R/run_phase2_character.R`
- `R/phase2_*.R`

This notebook reads the saved artifacts in `outputs/` so we can:

- review the chosen `k = 7` Character calibration
- inspect the published cluster structure and provisional names
- sanity-check soft membership and peer outputs
- review the Character-specific hypothesis tests
- identify any final naming or interpretation adjustments before broader promotion

## Model Snapshot

```{r model-summary}
model_summary |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable()
```

## Cluster Count Calibration

```{r cluster-calibration-plot, fig.height=4.5}
cluster_calibration |>
  ggplot(aes(x = k, y = kmeans_avg_silhouette)) +
  geom_line(color = "#355070", linewidth = 0.8) +
  geom_point(aes(color = k == selected_k), size = 2.8) +
  scale_color_manual(values = c("FALSE" = "#6c757d", "TRUE" = "#d62828"), guide = "none") +
  scale_x_continuous(breaks = cluster_calibration$k) +
  labs(
    title = "Character cluster-count calibration",
    x = "Cluster count (k)",
    y = "Average silhouette"
  ) +
  theme_minimal(base_size = 12)
```

```{r cluster-calibration-table}
cluster_calibration |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable()
```

## Published Cluster Sizes

```{r selected-cluster-sizes}
selected_cluster_sizes |>
  rename(character_kmeans_cluster = cluster, metros_in_cluster = cluster_size) |>
  left_join(
    cluster_centroids |>
      distinct(character_kmeans_cluster, character_cluster_name),
    by = "character_kmeans_cluster"
  ) |>
  select(character_kmeans_cluster, character_cluster_name, metros_in_cluster) |>
  kable()
```

## Published Cluster Profiles

```{r centroids}
cluster_centroids |>
  transmute(
    character_cluster_name,
    metros_in_cluster,
    character_score = round(character_score, 3),
    character_percentile = round(character_percentile, 1),
    demographics_score = round(subject_score_demographics, 3),
    social_fabric_score = round(subject_score_social_fabric, 3)
  ) |>
  kable()
```

```{r representatives}
cluster_representatives |>
  group_by(character_cluster_name) |>
  slice_head(n = 6) |>
  ungroup() |>
  mutate(
    distance_to_center = round(distance_to_center, 2),
    character_percentile = round(character_percentile, 1)
  ) |>
  select(character_cluster_name, cbsa_name, distance_to_center, character_percentile) |>
  kable()
```

## Cluster KPI Table

This is the raw side-by-side view of the average KPI profile for each published cluster.

```{r cluster-kpi-table}
cluster_kpi_averages |>
  transmute(
    character_cluster_name,
    metros,
    diversity_index = round(imputed_diversity_index, 3),
    pct_black_nh = round(imputed_pct_black_nh, 3),
    pct_asian_nh = round(imputed_pct_asian_nh, 3),
    pct_hispanic = round(imputed_pct_hispanic, 3),
    pct_age_over_64 = round(imputed_pct_age_over_64, 3),
    pct_ba_plus = round(imputed_pct_ba_plus, 3),
    pct_foreign_born = round(imputed_pct_foreign_born, 3),
    pop_weighted_density_sqmi = round(imputed_pop_weighted_density_sqmi, 0),
    friending_bias = round(imputed_friending_bias, 3),
    civic_engagement_volunteering_rate = round(imputed_civic_engagement_volunteering_rate, 3),
    civic_organizations_per_1000 = round(imputed_civic_organizations_per_1000, 3),
    nonprofits_per_100k = round(imputed_nonprofits_per_100k, 1),
    irs_net_migration_rate = round(imputed_irs_net_migration_rate, 4),
    pct_moved_diff_st = round(imputed_pct_moved_diff_st, 3),
    pct_moved_abroad = round(imputed_pct_moved_abroad, 3),
    social_associations_per_10k = round(imputed_social_associations_per_10k, 2),
    pct_struct_multifam = round(imputed_pct_struct_multifam, 3)
  ) |>
  kable()
```

## Metric Signatures

```{r metric-extremes}
cluster_metric_extremes |>
  left_join(
    cluster_centroids |>
      distinct(character_kmeans_cluster, character_cluster_name),
    by = "character_kmeans_cluster"
  ) |>
  mutate(avg_metric_score = round(avg_metric_score, 3)) |>
  arrange(character_cluster_name, extreme_direction, desc(avg_metric_score)) |>
  select(character_cluster_name, metric_id, avg_metric_score, extreme_direction) |>
  kable()
```

## Soft Membership Review

```{r gmm-summary}
gmm_summary |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable()
```

```{r gmm-hybrids}
gmm_hybrids |>
  arrange(membership_gap) |>
  slice_head(n = 20) |>
  mutate(across(contains("probability"), \(x) round(x, 3))) |>
  kable()
```

## Similarity Sanity Check

```{r peers}
similarity_top10 |>
  filter(cbsa_code %in% c(16980, 19660, 31540)) |>
  mutate(cosine_similarity = round(cosine_similarity, 3)) |>
  kable()
```

## Hypothesis Review

### Size Dominance Check

```{r size-dominance-summary}
size_dominance_summary |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable()
```

```{r size-dominance-plot, fig.height=5}
size_dominance_scatter |>
  ggplot(aes(
    x = log_pop_total,
    y = character_score,
    color = character_cluster_name
  )) +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "Character score is related to scale, but not dominated by raw metro size",
    x = "Log population",
    y = "Character frame score",
    color = "Cluster"
  ) +
  theme_minimal(base_size = 12)
```

### Connectedness Cluster Shift Check

```{r connectedness-summary}
connectedness_shift_summary |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable()
```

```{r connectedness-outliers}
connectedness_shift_outliers |>
  mutate(changed_under_connectedness = if_else(changed_under_connectedness, "yes", "no")) |>
  kable()
```

## Final Read

- The canonical Phase 2 build now lives in `R/run_phase2_character.R`.
- The published Character clustering set keeps all `17` approved KPIs.
- The current working Character typology uses `k = 7`.
- The regularized GMM fit now produces usable soft-membership review outputs.
- The main remaining judgment call is interpretive: whether any of the provisional cluster names or writeups should be tuned before broader promotion.
