---
title: "Variable Selection - Livability"
subtitle: "Subject, topic, and KPI diagnostic for Livability frame inputs"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 3
    code-fold: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)

options(
  dplyr.summarise.inform = FALSE,
  knitr.kable.NA = ""
)

library(here)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(glue)
library(purrr)
library(stringr)
library(knitr)

source(here("exploration/intelligence_framework/R/utils.R"))
con <- db_connect()
```

```{r catalog, include=FALSE}
metric_labels <- c(
  rent_to_income = "Rent-to-income ratio",
  value_to_income = "Home value-to-income ratio",
  rpp_real_pc_income = "RPP-adjusted per-capita income",
  pct_rent_burden_30plus = "Rent burden 30%+",
  pct_rent_burden_50plus = "Rent burden 50%+",
  pct_cost_burdened = "CHAS cost burdened share",
  pct_severely_cost_burdened = "CHAS severely burdened share",
  pct_renter_severely_cost_burdened = "CHAS renter severely burdened share",
  pov_rate = "Poverty rate",
  permits_per_1000_housing_units = "Permits per 1k housing units",
  permits_per_1000_population = "Permits per 1k population",
  permits_share_multifam_units = "Multifamily permit share",
  permits_share_units_5_plus = "5+ unit permit share",
  permits_avg_units_per_bldg = "Average units per building",
  pct_struct_multifam = "Existing multifamily share",
  pct_struct_sf_det = "Single-family detached share",
  pct_struct_small_mf = "Small multifamily share",
  pct_struct_mid_mf = "Mid multifamily share",
  pct_struct_large_mf = "Large multifamily share",
  pct_struct_mobile = "Mobile-home share",
  life_expectancy = "Life expectancy",
  premature_death_rate = "Premature death rate",
  child_mortality_rate = "Child mortality rate",
  infant_mortality_rate = "Infant mortality rate",
  drug_overdose_death_rate = "Drug overdose death rate",
  pct_uninsured_adults = "Uninsured adult share",
  preventable_hospital_stay_rate = "Preventable hospital stay rate",
  primary_care_ratio = "Primary care provider ratio",
  mental_health_provider_ratio = "Mental health provider ratio",
  physical_inactivity = "Physical inactivity",
  adult_obesity = "Adult obesity",
  poor_mental_health_days = "Poor mental health days",
  food_insecurity_rate = "Food insecurity rate",
  child_care_cost_burden_rate = "Child care cost burden rate",
  homicide_rate = "Homicide rate",
  firearm_fatality_rate = "Firearm fatality rate",
  motor_vehicle_crash_rate = "Motor vehicle crash death rate",
  mean_travel_time = "Mean travel time",
  pct_commute_transit = "Transit commute share",
  pct_commute_walk = "Walk commute share",
  pct_commute_wfh = "Work-from-home share",
  pct_hh_0_vehicles = "Zero-vehicle household share",
  vacancy_rate = "Vacancy rate",
  pct_broadband_subscription = "Broadband subscription share",
  pct_no_internet_access = "No-internet share",
  walkability_index = "Walkability index",
  transit_service_density = "Transit service density",
  jobs_access_45min_transit = "Jobs access within 45 min transit",
  jobs_access_45min_auto = "Jobs access within 45 min auto",
  pct_population_low_access_1_10 = "Low-access population share",
  pct_population_low_income_low_access_1_10 = "Low-income low-access population share",
  pop_weighted_density_sqmi = "Population-weighted density",
  gross_density_sqmi = "Gross density",
  hs_graduation_rate = "HS graduation rate",
  math_score_index = "Math score index",
  reading_score_index = "Reading score index",
  air_pollution_pm25 = "CHR PM2.5 proxy",
  aqi_median = "Median AQI",
  aqi_p90 = "AQI P90",
  unhealthy_days = "Unhealthy AQI days",
  ejs_pm25 = "EJScreen PM2.5",
  ejs_ozone = "EJScreen ozone",
  ejs_diesel_pm = "EJScreen diesel PM",
  ejs_traffic_proximity = "EJScreen traffic proximity",
  ejs_superfund_proximity = "EJScreen superfund proximity",
  ejs_rmp_proximity = "EJScreen RMP proximity",
  ejs_wastewater_discharge = "EJScreen wastewater discharge",
  ejs_drinking_water_noncompliance = "EJScreen drinking water noncompliance",
  adverse_climate_events = "CHR adverse climate-events proxy",
  fema_risk_score = "FEMA composite risk score",
  fema_eal_score = "FEMA expected annual loss score"
)

metric_catalog <- tibble::tribble(
  ~metric, ~subject, ~topic, ~source_family, ~role, ~note,
  "rent_to_income", "Affordability", "Price pressure", "ACS affordability", "direct", "Recurring ACS affordability level.",
  "value_to_income", "Affordability", "Price pressure", "ACS affordability", "direct", "Recurring ACS affordability level.",
  "rpp_real_pc_income", "Affordability", "Price pressure", "BEA RPP", "direct", "Cross-metro purchasing-power adjustment with partial CBSA coverage.",
  "pct_rent_burden_30plus", "Affordability", "Housing burden", "ACS rent burden", "direct", "Preferred recurring burden KPI in the Phase 1 mapping audit.",
  "pct_rent_burden_50plus", "Affordability", "Housing burden", "ACS rent burden", "direct", "More extreme recurring burden variant.",
  "pct_cost_burdened", "Affordability", "Housing burden", "HUD CHAS", "baseline", "CHAS burden field; contextual and not recurring across years.",
  "pct_severely_cost_burdened", "Affordability", "Housing burden", "HUD CHAS", "baseline", "CHAS burden field; contextual and not recurring across years.",
  "pct_renter_severely_cost_burdened", "Affordability", "Housing burden", "HUD CHAS", "baseline", "CHAS renter-only burden field; contextual and not recurring across years.",
  "pov_rate", "Affordability", "Poverty context", "ACS income", "direct", "Affordability context rather than a pure housing KPI.",
  "permits_per_1000_housing_units", "Affordability", "Housing supply", "BPS permits", "direct", "Recurring permit intensity signal.",
  "permits_per_1000_population", "Affordability", "Housing supply", "BPS permits", "direct", "Alternative permit intensity denominator.",
  "permits_share_multifam_units", "Affordability", "Housing supply", "BPS permits", "direct", "Density-of-supply signal.",
  "permits_share_units_5_plus", "Affordability", "Housing supply", "BPS permits", "direct", "Higher-density supply composition signal.",
  "permits_avg_units_per_bldg", "Affordability", "Housing supply", "BPS permits", "direct", "Average building scale of new supply.",
  "pct_struct_multifam", "Affordability", "Housing structure mix", "ACS housing stock", "direct", "Existing stock composition, distinct from permits.",
  "pct_struct_sf_det", "Affordability", "Housing structure mix", "ACS housing stock", "direct", "Existing stock composition, distinct from permits.",
  "pct_struct_small_mf", "Affordability", "Housing structure mix", "ACS housing stock", "direct", "Existing stock composition, distinct from permits.",
  "pct_struct_mid_mf", "Affordability", "Housing structure mix", "ACS housing stock", "direct", "Existing stock composition, distinct from permits.",
  "pct_struct_large_mf", "Affordability", "Housing structure mix", "ACS housing stock", "direct", "Existing stock composition, distinct from permits.",
  "pct_struct_mobile", "Affordability", "Housing structure mix", "ACS housing stock", "direct", "Existing stock composition, distinct from permits.",
  "life_expectancy", "Health & Safety", "Health outcomes", "CHR health", "direct", "Headline CHR health outcome.",
  "premature_death_rate", "Health & Safety", "Health outcomes", "CHR health", "direct", "Headline CHR mortality outcome.",
  "child_mortality_rate", "Health & Safety", "Health outcomes", "CHR health", "direct", "Child mortality coverage is decent but below the headline measures.",
  "infant_mortality_rate", "Health & Safety", "Health outcomes", "CHR health", "direct", "Important but materially sparse at CBSA scale.",
  "drug_overdose_death_rate", "Health & Safety", "Health behavior and access", "CHR health", "direct", "Important contemporary health-risk signal.",
  "pct_uninsured_adults", "Health & Safety", "Health behavior and access", "CHR health", "direct", "Insurance access lens.",
  "preventable_hospital_stay_rate", "Health & Safety", "Health behavior and access", "CHR health", "direct", "Healthcare-system access / quality lens.",
  "primary_care_ratio", "Health & Safety", "Health behavior and access", "CHR health", "direct", "Primary care access ratio.",
  "mental_health_provider_ratio", "Health & Safety", "Health behavior and access", "CHR health", "direct", "Mental-health access ratio.",
  "physical_inactivity", "Health & Safety", "Health behavior and access", "CHR health", "direct", "Behavioral health lens.",
  "adult_obesity", "Health & Safety", "Health behavior and access", "CHR health", "direct", "Behavioral health lens.",
  "poor_mental_health_days", "Health & Safety", "Health behavior and access", "CHR health", "direct", "Mental-health outcome lens.",
  "food_insecurity_rate", "Health & Safety", "Health behavior and access", "CHR health", "direct", "Household stress / nutrition lens.",
  "child_care_cost_burden_rate", "Health & Safety", "Health behavior and access", "CHR health", "direct", "Household stress / care infrastructure lens.",
  "homicide_rate", "Health & Safety", "Violence and injury", "CHR health", "direct", "Violence outcome lens, but coverage is limited.",
  "firearm_fatality_rate", "Health & Safety", "Violence and injury", "CHR health", "direct", "Broader violence / self-harm / accident exposure lens.",
  "motor_vehicle_crash_rate", "Health & Safety", "Violence and injury", "CHR health", "direct", "Transport safety lens.",
  "mean_travel_time", "Access & Infrastructure", "Commute and mode access", "ACS transport", "direct", "Recurring commute burden signal.",
  "pct_commute_transit", "Access & Infrastructure", "Commute and mode access", "ACS transport", "direct", "Transit use lens.",
  "pct_commute_walk", "Access & Infrastructure", "Commute and mode access", "ACS transport", "direct", "Walk mode share lens.",
  "pct_commute_wfh", "Access & Infrastructure", "Commute and mode access", "ACS transport", "direct", "Hybrid / remote access lens.",
  "pct_hh_0_vehicles", "Access & Infrastructure", "Vehicle access", "ACS transport", "direct", "Vehicle dependence / transit need lens.",
  "vacancy_rate", "Access & Infrastructure", "Housing slack", "ACS housing stock", "direct", "Housing-market slack lens that also matters for access.",
  "pct_broadband_subscription", "Access & Infrastructure", "Digital access", "ACS broadband", "direct", "Digital infrastructure lens.",
  "pct_no_internet_access", "Access & Infrastructure", "Digital access", "ACS broadband", "direct", "Digital exclusion counterpart.",
  "walkability_index", "Access & Infrastructure", "Walkability and transit baseline", "EPA SLD", "baseline", "Single-vintage baseline promoted because the metric map names the topic but not the KPIs.",
  "transit_service_density", "Access & Infrastructure", "Walkability and transit baseline", "EPA SLD", "baseline", "Single-vintage baseline promoted because the metric map names the topic but not the KPIs.",
  "jobs_access_45min_transit", "Access & Infrastructure", "Walkability and transit baseline", "EPA SLD", "baseline", "Single-vintage baseline promoted because the metric map names the topic but not the KPIs.",
  "jobs_access_45min_auto", "Access & Infrastructure", "Walkability and transit baseline", "EPA SLD", "baseline", "Auto-access complement to the transit accessibility story.",
  "pct_population_low_access_1_10", "Access & Infrastructure", "Food access baseline", "USDA food access", "baseline", "Single-vintage 2019 baseline promoted because the metric map names the topic but not the KPIs.",
  "pct_population_low_income_low_access_1_10", "Access & Infrastructure", "Food access baseline", "USDA food access", "baseline", "Single-vintage 2019 baseline promoted because the metric map names the topic but not the KPIs.",
  "pop_weighted_density_sqmi", "Access & Infrastructure", "Built-form proxy", "ACS built form", "proxy", "Practical proxy for access intensity; primary home is Character.",
  "gross_density_sqmi", "Access & Infrastructure", "Built-form proxy", "ACS built form", "proxy", "Practical proxy for access intensity; primary home is Character.",
  "hs_graduation_rate", "Education access", "School performance", "CHR education", "direct", "Only high-coverage school KPI in the live set.",
  "math_score_index", "Education access", "School performance", "CHR education", "direct", "Live but somewhat sparser than HS graduation.",
  "reading_score_index", "Education access", "School performance", "CHR education", "direct", "Live but materially sparser than HS graduation.",
  "air_pollution_pm25", "Physical Environment", "Air pollution", "CHR environment proxy", "proxy", "Lagged CHR air-pollution proxy retained for comparison with direct AQI/EJScreen fields.",
  "aqi_median", "Physical Environment", "Air pollution", "EPA AQI", "direct", "Direct AQI exposure signal.",
  "aqi_p90", "Physical Environment", "Air pollution", "EPA AQI", "direct", "Tail-risk AQI signal.",
  "unhealthy_days", "Physical Environment", "Air pollution", "EPA AQI", "direct", "Air-quality bad-days count.",
  "ejs_pm25", "Physical Environment", "Air pollution", "EPA EJScreen", "direct", "Environmental burden lens with partial coverage.",
  "ejs_ozone", "Physical Environment", "Air pollution", "EPA EJScreen", "direct", "Environmental burden lens with partial coverage.",
  "ejs_diesel_pm", "Physical Environment", "Air pollution", "EPA EJScreen", "direct", "Environmental burden lens with partial coverage.",
  "ejs_traffic_proximity", "Physical Environment", "Hazard exposure", "EPA EJScreen", "direct", "Traffic-related burden baseline.",
  "ejs_superfund_proximity", "Physical Environment", "Hazard exposure", "EPA EJScreen", "direct", "Toxic-site burden baseline.",
  "ejs_rmp_proximity", "Physical Environment", "Hazard exposure", "EPA EJScreen", "direct", "Risk-management-plan site proximity baseline.",
  "ejs_wastewater_discharge", "Physical Environment", "Hazard exposure", "EPA EJScreen", "direct", "Water-quality burden baseline.",
  "ejs_drinking_water_noncompliance", "Physical Environment", "Hazard exposure", "EPA EJScreen", "direct", "Drinking-water burden baseline.",
  "adverse_climate_events", "Physical Environment", "Climate hazard risk", "CHR environment proxy", "proxy", "Lagged CHR climate proxy retained for comparison with FEMA.",
  "fema_risk_score", "Physical Environment", "Climate hazard risk", "FEMA NRI", "direct", "Composite climate and hazard risk surface.",
  "fema_eal_score", "Physical Environment", "Climate hazard risk", "FEMA NRI", "direct", "Expected annual loss complement to FEMA risk score."
)

topic_inventory_notes <- tibble::tribble(
  ~subject, ~topic, ~status, ~note,
  "Access & Infrastructure", "Walkability and transit baseline", "live baseline", "Metric map named the topic but did not pin concrete KPIs. This notebook uses `walkability_index`, `transit_service_density`, and `jobs_access_45min_transit` from `gold.transport_built_form_sld` as the first-pass live stand-ins.",
  "Access & Infrastructure", "Food access baseline", "live baseline", "Metric map named the topic but did not pin concrete KPIs. This notebook uses USDA baseline shares from `gold.food_access_wide` as the first-pass live stand-ins.",
  "Access & Infrastructure", "Built-form proxy", "proxy", "Metric map points to density derivation via `dim_geo`, but live Gold already exposes recurring built-form density in `gold.transport_built_form_wide`. Keep as a proxy topic and explicitly test overlap against housing structure and vacancy.",
  "Education access", "K-12 learning quality", "deferred", "Stanford SEDA remains deferred in the metric map and is not part of the live Gold frame.",
  "Physical Environment", "Air pollution", "mixed direct + proxy", "Use EPA AQI and EJScreen as the direct source family. Keep CHR `air_pollution_pm25` only as a proxy comparison field.",
  "Physical Environment", "Climate hazard risk", "mixed direct + proxy", "Use FEMA NRI as the direct source family. Keep CHR `adverse_climate_events` only as a proxy comparison field.",
  "Health & Safety", "Health & Safety weighting note", "modeling note", "Use one consolidated Health & Safety subject in modeling so the Safety topic does not get an automatic extra subject vote."
)
```

## Goal

Build a complete Phase 1 Livability selection pass that is explicit at three levels:

1. **Subject**: does the subject belong in the Livability frame at all?
2. **Topic**: does the topic survive as a real dimension, collapse to one lead KPI, stay only as a baseline / sensitivity topic, or drop?
3. **KPI**: which metrics are core, redundant, optional sensitivity tests, or drops?

This notebook uses the same CBSA universe as the Character pass:

- `geo_level = 'cbsa'`
- `pop_total >= 100000`
- latest available row by **source family**, not a single blanket year for every table

## Decision rules

These rules are used throughout the notebook and should be read as screens, not rigid auto-decisions:

- **Coverage**: below `90%` coverage is a caution flag, not an automatic drop.
- **Variance**: below `CV = 0.2` is weak differentiation at CBSA scale.
- **Correlation**: above `|r| = 0.75` means we should not carry every KPI at full weight inside the same topic.
- **PCA**: diagnostic only. It tells us whether a topic is really multi-dimensional after the basic screens.

This pass also uses a few explicit judgment calls:

- recurring series beat single-vintage baselines when the concepts are very similar
- single-vintage baseline topics can still stay in the Livability frame if they are conceptually strong and not too redundant
- Physical Environment stays in scope even with weaker coverage because it opens a genuinely different line of thinking
- Safety stays visible as its own topic family, but it now rolls up under one broader `Health & Safety` subject

## Phase 1 inventory

Livability is being treated as five subjects in this pass:

- `Affordability`
- `Health & Safety`
- `Access & Infrastructure`
- `Education access`
- `Physical Environment`

The goal is completeness, so this notebook explicitly distinguishes:

- recurring direct KPIs
- live but single-vintage baseline KPIs
- proxy KPIs
- deferred topics that still need to be named in the audit

```{r data}
df <- dbGetQuery(con, "
with spine as (
    select
        geo_id,
        geo_name,
        pop_total,
        year as spine_year
    from gold.population_demographics
    where geo_level = 'cbsa'
      and year = 2024
      and pop_total >= 100000
),
affordability_acs_latest as (
    select *
    from (
        select
            geo_id,
            year as affordability_acs_year,
            rent_to_income,
            value_to_income,
            pct_rent_burden_30plus,
            pct_rent_burden_50plus,
            row_number() over (
                partition by geo_id
                order by year desc
            ) as rn
        from gold.affordability_wide
        where geo_level = 'cbsa'
    )
    where rn = 1
),
affordability_rpp_latest as (
    select *
    from (
        select
            geo_id,
            year as affordability_rpp_year,
            rpp_real_pc_income,
            row_number() over (
                partition by geo_id
                order by case when rpp_real_pc_income is not null then year end desc nulls last
            ) as rn
        from gold.affordability_wide
        where geo_level = 'cbsa'
    )
    where rn = 1
),
affordability_chas_latest as (
    select *
    from (
        select
            geo_id,
            year as affordability_chas_year,
            pct_cost_burdened,
            pct_severely_cost_burdened,
            pct_renter_severely_cost_burdened,
            row_number() over (
                partition by geo_id
                order by case
                    when pct_cost_burdened is not null
                      or pct_severely_cost_burdened is not null
                      or pct_renter_severely_cost_burdened is not null
                    then year
                end desc nulls last
            ) as rn
        from gold.affordability_wide
        where geo_level = 'cbsa'
    )
    where rn = 1
),
income_latest as (
    select *
    from (
        select
            geo_id,
            year as income_year,
            pov_rate,
            row_number() over (
                partition by geo_id
                order by year desc
            ) as rn
        from gold.economics_income_wide
        where geo_level = 'cbsa'
    )
    where rn = 1
),
housing_latest as (
    select *
    from (
        select
            geo_id,
            year as housing_year,
            vacancy_rate,
            permits_per_1000_housing_units,
            permits_per_1000_population,
            permits_share_multifam_units,
            permits_share_units_5_plus,
            permits_avg_units_per_bldg,
            pct_struct_multifam,
            pct_struct_sf_det,
            pct_struct_small_mf,
            pct_struct_mid_mf,
            pct_struct_large_mf,
            pct_struct_mobile,
            row_number() over (
                partition by geo_id
                order by year desc
            ) as rn
        from gold.housing_core_wide
        where geo_level = 'cbsa'
    )
    where rn = 1
),
transport_latest as (
    select *
    from (
        select
            geo_id,
            year as transport_year,
            mean_travel_time,
            pct_commute_transit,
            pct_commute_walk,
            pct_commute_wfh,
            pct_hh_0_vehicles,
            pop_weighted_density_sqmi,
            gross_density_sqmi,
            row_number() over (
                partition by geo_id
                order by year desc
            ) as rn
        from gold.transport_built_form_wide
        where geo_level = 'cbsa'
    )
    where rn = 1
),
social_infra_latest as (
    select *
    from (
        select
            geo_id,
            year as social_infra_year,
            pct_broadband_subscription,
            pct_no_internet_access,
            row_number() over (
                partition by geo_id
                order by year desc
            ) as rn
        from gold.social_infra_wide
        where geo_level = 'cbsa'
    )
    where rn = 1
),
health_latest as (
    select *
    from (
        select
            geo_id,
            year as health_year,
            life_expectancy,
            premature_death_rate,
            child_mortality_rate,
            infant_mortality_rate,
            drug_overdose_death_rate,
            pct_uninsured_adults,
            preventable_hospital_stay_rate,
            primary_care_ratio,
            mental_health_provider_ratio,
            physical_inactivity,
            adult_obesity,
            poor_mental_health_days,
            food_insecurity_rate,
            child_care_cost_burden_rate,
            homicide_rate,
            firearm_fatality_rate,
            motor_vehicle_crash_rate,
            hs_graduation_rate,
            math_score_index,
            reading_score_index,
            air_pollution_pm25,
            adverse_climate_events,
            row_number() over (
                partition by geo_id
                order by year desc
            ) as rn
        from gold.health_wide
        where geo_level = 'cbsa'
    )
    where rn = 1
),
environment_aqi_latest as (
    select *
    from (
        select
            geo_id,
            year as environment_aqi_year,
            aqi_median,
            aqi_p90,
            unhealthy_days,
            row_number() over (
                partition by geo_id
                order by case
                    when aqi_median is not null
                      or aqi_p90 is not null
                      or unhealthy_days is not null
                    then year
                end desc nulls last
            ) as rn
        from gold.environment_wide
        where geo_level = 'cbsa'
    )
    where rn = 1
),
environment_ejs_latest as (
    select *
    from (
        select
            geo_id,
            year as environment_ejs_year,
            ejs_pm25,
            ejs_ozone,
            ejs_diesel_pm,
            ejs_traffic_proximity,
            ejs_superfund_proximity,
            ejs_rmp_proximity,
            ejs_wastewater_discharge,
            ejs_drinking_water_noncompliance,
            row_number() over (
                partition by geo_id
                order by case
                    when ejs_pm25 is not null
                      or ejs_ozone is not null
                      or ejs_diesel_pm is not null
                      or ejs_traffic_proximity is not null
                      or ejs_superfund_proximity is not null
                      or ejs_rmp_proximity is not null
                      or ejs_wastewater_discharge is not null
                      or ejs_drinking_water_noncompliance is not null
                    then year
                end desc nulls last
            ) as rn
        from gold.environment_wide
        where geo_level = 'cbsa'
    )
    where rn = 1
),
environment_fema_latest as (
    select *
    from (
        select
            geo_id,
            year as environment_fema_year,
            fema_risk_score,
            fema_eal_score,
            row_number() over (
                partition by geo_id
                order by case
                    when fema_risk_score is not null
                      or fema_eal_score is not null
                    then year
                end desc nulls last
            ) as rn
        from gold.environment_wide
        where geo_level = 'cbsa'
    )
    where rn = 1
),
sld_latest as (
    select *
    from (
        select
            geo_id,
            year as sld_year,
            walkability_index,
            transit_service_density,
            jobs_access_45min_transit,
            jobs_access_45min_auto,
            row_number() over (
                partition by geo_id
                order by case
                    when walkability_index is not null
                      or transit_service_density is not null
                      or jobs_access_45min_transit is not null
                      or jobs_access_45min_auto is not null
                    then year
                end desc nulls last
            ) as rn
        from gold.transport_built_form_sld
        where geo_level = 'cbsa'
    )
    where rn = 1
),
food_latest as (
    select *
    from (
        select
            geo_id,
            year as food_access_year,
            pct_population_low_access_1_10,
            pct_population_low_income_low_access_1_10,
            row_number() over (
                partition by geo_id
                order by case
                    when pct_population_low_access_1_10 is not null
                      or pct_population_low_income_low_access_1_10 is not null
                    then year
                end desc nulls last
            ) as rn
        from gold.food_access_wide
        where geo_level = 'cbsa'
    )
    where rn = 1
)
select
    spine.geo_id,
    spine.geo_name,
    spine.pop_total,
    spine.spine_year,
    affordability_acs.affordability_acs_year,
    affordability_rpp.affordability_rpp_year,
    affordability_chas.affordability_chas_year,
    income.income_year,
    housing.housing_year,
    transport.transport_year,
    social_infra.social_infra_year,
    health.health_year,
    environment_aqi.environment_aqi_year,
    environment_ejs.environment_ejs_year,
    environment_fema.environment_fema_year,
    sld.sld_year,
    food.food_access_year,
    affordability_acs.rent_to_income,
    affordability_acs.value_to_income,
    affordability_rpp.rpp_real_pc_income,
    affordability_acs.pct_rent_burden_30plus,
    affordability_acs.pct_rent_burden_50plus,
    affordability_chas.pct_cost_burdened,
    affordability_chas.pct_severely_cost_burdened,
    affordability_chas.pct_renter_severely_cost_burdened,
    income.pov_rate,
    housing.permits_per_1000_housing_units,
    housing.permits_per_1000_population,
    housing.permits_share_multifam_units,
    housing.permits_share_units_5_plus,
    housing.permits_avg_units_per_bldg,
    housing.pct_struct_multifam,
    housing.pct_struct_sf_det,
    housing.pct_struct_small_mf,
    housing.pct_struct_mid_mf,
    housing.pct_struct_large_mf,
    housing.pct_struct_mobile,
    health.life_expectancy,
    health.premature_death_rate,
    health.child_mortality_rate,
    health.infant_mortality_rate,
    health.drug_overdose_death_rate,
    health.pct_uninsured_adults,
    health.preventable_hospital_stay_rate,
    health.primary_care_ratio,
    health.mental_health_provider_ratio,
    health.physical_inactivity,
    health.adult_obesity,
    health.poor_mental_health_days,
    health.food_insecurity_rate,
    health.child_care_cost_burden_rate,
    health.homicide_rate,
    health.firearm_fatality_rate,
    health.motor_vehicle_crash_rate,
    transport.mean_travel_time,
    transport.pct_commute_transit,
    transport.pct_commute_walk,
    transport.pct_commute_wfh,
    transport.pct_hh_0_vehicles,
    housing.vacancy_rate,
    social_infra.pct_broadband_subscription,
    social_infra.pct_no_internet_access,
    sld.walkability_index,
    sld.transit_service_density,
    sld.jobs_access_45min_transit,
    sld.jobs_access_45min_auto,
    food.pct_population_low_access_1_10,
    food.pct_population_low_income_low_access_1_10,
    transport.pop_weighted_density_sqmi,
    transport.gross_density_sqmi,
    health.hs_graduation_rate,
    health.math_score_index,
    health.reading_score_index,
    health.air_pollution_pm25,
    environment_aqi.aqi_median,
    environment_aqi.aqi_p90,
    environment_aqi.unhealthy_days,
    environment_ejs.ejs_pm25,
    environment_ejs.ejs_ozone,
    environment_ejs.ejs_diesel_pm,
    environment_ejs.ejs_traffic_proximity,
    environment_ejs.ejs_superfund_proximity,
    environment_ejs.ejs_rmp_proximity,
    environment_ejs.ejs_wastewater_discharge,
    environment_ejs.ejs_drinking_water_noncompliance,
    health.adverse_climate_events,
    environment_fema.fema_risk_score,
    environment_fema.fema_eal_score
from spine
left join affordability_acs_latest affordability_acs
    on spine.geo_id = affordability_acs.geo_id
left join affordability_rpp_latest affordability_rpp
    on spine.geo_id = affordability_rpp.geo_id
left join affordability_chas_latest affordability_chas
    on spine.geo_id = affordability_chas.geo_id
left join income_latest income
    on spine.geo_id = income.geo_id
left join housing_latest housing
    on spine.geo_id = housing.geo_id
left join transport_latest transport
    on spine.geo_id = transport.geo_id
left join social_infra_latest social_infra
    on spine.geo_id = social_infra.geo_id
left join health_latest health
    on spine.geo_id = health.geo_id
left join environment_aqi_latest environment_aqi
    on spine.geo_id = environment_aqi.geo_id
left join environment_ejs_latest environment_ejs
    on spine.geo_id = environment_ejs.geo_id
left join environment_fema_latest environment_fema
    on spine.geo_id = environment_fema.geo_id
left join sld_latest sld
    on spine.geo_id = sld.geo_id
left join food_latest food
    on spine.geo_id = food.geo_id
order by spine.pop_total desc
")

metrics <- metric_catalog$metric

coverage_tbl <- tibble(
  metric = metrics,
  non_null_cbsas = sapply(metrics, \(m) sum(!is.na(df[[m]]))),
  coverage_pct = non_null_cbsas / nrow(df)
) |>
  left_join(metric_catalog, by = "metric") |>
  mutate(metric_label = unname(metric_labels[metric])) |>
  arrange(subject, topic, desc(coverage_pct), metric)

source_family_tbl <- tibble::tribble(
  ~source_family, ~latest_year, ~coverage_metric, ~coverage_n, ~note,
  "ACS affordability", max(df$affordability_acs_year, na.rm = TRUE), "rent_to_income", sum(!is.na(df$rent_to_income)), "Recurring affordability levels and ACS rent-burden fields.",
  "BEA RPP", max(df$affordability_rpp_year, na.rm = TRUE), "rpp_real_pc_income", sum(!is.na(df$rpp_real_pc_income)), "Partial CBSA coverage, but still usable as contextual normalization.",
  "HUD CHAS", max(df$affordability_chas_year, na.rm = TRUE), "pct_cost_burdened", sum(!is.na(df$pct_cost_burdened)), "Single-vintage burden fields; contextual rather than recurring.",
  "ACS income", max(df$income_year, na.rm = TRUE), "pov_rate", sum(!is.na(df$pov_rate)), "Affordability context family.",
  "BPS permits / ACS housing", max(df$housing_year, na.rm = TRUE), "permits_per_1000_housing_units", sum(!is.na(df$permits_per_1000_housing_units)), "Supply and structure mix family.",
  "ACS transport / built form", max(df$transport_year, na.rm = TRUE), "mean_travel_time", sum(!is.na(df$mean_travel_time)), "Recurring commute, vehicle, and density family.",
  "ACS broadband", max(df$social_infra_year, na.rm = TRUE), "pct_broadband_subscription", sum(!is.na(df$pct_broadband_subscription)), "Recurring digital-access family.",
  "CHR health", max(df$health_year, na.rm = TRUE), "life_expectancy", sum(!is.na(df$life_expectancy)), "Health, safety, education, and proxy environment family.",
  "EPA AQI", max(df$environment_aqi_year, na.rm = TRUE), "aqi_median", sum(!is.na(df$aqi_median)), "Direct AQI family with monitored-geography limits.",
  "EPA EJScreen", max(df$environment_ejs_year, na.rm = TRUE), "ejs_pm25", sum(!is.na(df$ejs_pm25)), "Direct environmental-burden family with partial coverage.",
  "FEMA NRI", max(df$environment_fema_year, na.rm = TRUE), "fema_risk_score", sum(!is.na(df$fema_risk_score)), "Direct climate-risk family with partial coverage.",
  "EPA SLD", max(df$sld_year, na.rm = TRUE), "walkability_index", sum(!is.na(df$walkability_index)), "Single-vintage 2021 baseline; still strong enough to test as a topic.",
  "USDA food access", max(df$food_access_year, na.rm = TRUE), "pct_population_low_access_1_10", sum(!is.na(df$pct_population_low_access_1_10)), "Single-vintage 2019 baseline; still strong enough to test as a topic."
) |>
  mutate(coverage_pct = coverage_n / nrow(df))

variance_tbl <- tibble(metric = metrics) |>
  mutate(
    n = sapply(metric, \(m) sum(!is.na(df[[m]]))),
    mean = sapply(metric, \(m) mean(df[[m]], na.rm = TRUE)),
    sd = sapply(metric, \(m) sd(df[[m]], na.rm = TRUE)),
    cv = sd / abs(mean),
    iqr = sapply(metric, \(m) IQR(df[[m]], na.rm = TRUE))
  ) |>
  left_join(metric_catalog, by = "metric") |>
  left_join(coverage_tbl |> select(metric, coverage_pct), by = "metric") |>
  mutate(metric_label = unname(metric_labels[metric])) |>
  arrange(desc(cv))

topic_scorecard_tbl <- variance_tbl |>
  group_by(subject, topic) |>
  summarise(
    metric_count = n(),
    avg_coverage_pct = mean(coverage_pct),
    direct_metrics = sum(role == "direct"),
    baseline_metrics = sum(role == "baseline"),
    proxy_metrics = sum(role == "proxy"),
    high_variance_metrics = sum(cv >= 0.2, na.rm = TRUE),
    top_kpis = paste(metric_label[order(cv, decreasing = TRUE)][seq_len(min(3, n()))], collapse = "; "),
    .groups = "drop"
  ) |>
  mutate(
    topic_read = case_when(
      high_variance_metrics == 0 ~ "Weak topic at CBSA scale",
      high_variance_metrics == 1 ~ "Keep topic, likely with one lead KPI",
      TRUE ~ "Keep topic, compare KPIs inside topic"
    )
  ) |>
  arrange(subject, topic)

surviving <- variance_tbl |>
  filter(cv >= 0.2) |>
  pull(metric)

df_cor <- df |>
  select(all_of(surviving)) |>
  drop_na()

cor_matrix <- cor(df_cor, use = "pairwise.complete.obs")
high_cor <- which(abs(cor_matrix) > 0.75 & upper.tri(cor_matrix), arr.ind = TRUE)

high_cor_tbl <- if (nrow(high_cor) > 0) {
  tibble(
    metric_a = rownames(cor_matrix)[high_cor[, 1]],
    metric_b = colnames(cor_matrix)[high_cor[, 2]],
    correlation = cor_matrix[high_cor]
  ) |>
    mutate(
      metric_a_label = unname(metric_labels[metric_a]),
      metric_b_label = unname(metric_labels[metric_b])
    ) |>
    left_join(metric_catalog |> rename(metric_a = metric, subject_a = subject, topic_a = topic), by = "metric_a") |>
    left_join(metric_catalog |> rename(metric_b = metric, subject_b = subject, topic_b = topic), by = "metric_b") |>
    arrange(desc(abs(correlation)))
} else {
  tibble(
    metric_a = character(),
    metric_b = character(),
    correlation = double(),
    metric_a_label = character(),
    metric_b_label = character(),
    subject_a = character(),
    topic_a = character(),
    subject_b = character(),
    topic_b = character()
  )
}

topic_pca_tbl <- map_dfr(split(metric_catalog, metric_catalog$topic), \(topic_df) {
  topic_metrics <- intersect(topic_df$metric, metrics)
  topic_complete <- df |>
    select(all_of(topic_metrics)) |>
    drop_na()

  if (length(topic_metrics) < 2) {
    return(tibble(
      subject = topic_df$subject[[1]],
      topic = topic_df$topic[[1]],
      surviving_metrics = length(topic_metrics),
      complete_rows = nrow(topic_complete),
      components_80 = NA_integer_,
      topic_dimensionality = "Single-metric topic"
    ))
  }

  if (nrow(topic_complete) < 50) {
    return(tibble(
      subject = topic_df$subject[[1]],
      topic = topic_df$topic[[1]],
      surviving_metrics = length(topic_metrics),
      complete_rows = nrow(topic_complete),
      components_80 = NA_integer_,
      topic_dimensionality = "Too sparse for stable topic PCA"
    ))
  }

  topic_pca <- prcomp(scale(topic_complete), center = FALSE, scale. = FALSE)
  topic_components_80 <- which(cumsum((topic_pca$sdev^2) / sum(topic_pca$sdev^2)) >= 0.8)[1]

  tibble(
    subject = topic_df$subject[[1]],
    topic = topic_df$topic[[1]],
    surviving_metrics = length(topic_metrics),
    complete_rows = nrow(topic_complete),
    components_80 = topic_components_80,
    topic_dimensionality = case_when(
      topic_components_80 <= 1 ~ "Mostly one dominant KPI story",
      topic_components_80 == 2 ~ "Has two meaningful sub-dimensions",
      TRUE ~ "Multi-dimensional topic"
    )
  )
})

cor_flag_tbl <- bind_rows(
  high_cor_tbl |> transmute(metric = metric_a, correlation = abs(correlation)),
  high_cor_tbl |> transmute(metric = metric_b, correlation = abs(correlation))
) |>
  group_by(metric) |>
  summarise(max_abs_correlation = max(correlation), .groups = "drop")

kpi_decision_tbl <- variance_tbl |>
  left_join(cor_flag_tbl, by = "metric") |>
  mutate(
    max_abs_correlation = coalesce(max_abs_correlation, 0),
    recommendation = case_when(
      metric == "pct_rent_burden_30plus" ~ "Core candidate",
      metric %in% c(
        "permits_per_1000_housing_units",
        "permits_share_units_5_plus",
        "premature_death_rate",
        "firearm_fatality_rate",
        "motor_vehicle_crash_rate",
        "walkability_index",
        "jobs_access_45min_transit",
        "pct_population_low_income_low_access_1_10",
        "unhealthy_days",
        "fema_risk_score"
      ) ~ "Core candidate",
      metric %in% c(
        "homicide_rate",
        "pct_commute_transit",
        "rpp_real_pc_income",
        "rent_to_income",
        "child_mortality_rate",
        "primary_care_ratio",
        "permits_avg_units_per_bldg",
        "ejs_diesel_pm",
        "ejs_wastewater_discharge",
        "pop_weighted_density_sqmi"
      ) ~ "Optional sensitivity",
      role == "baseline" ~ "Optional sensitivity",
      role == "proxy" ~ "Optional sensitivity",
      coverage_pct < 0.75 ~ "Drop",
      cv < 0.2 ~ "Drop",
      max_abs_correlation >= 0.9 ~ "Redundant - keep one lead KPI",
      max_abs_correlation >= 0.75 ~ "Keep, but de-duplicate",
      coverage_pct < 0.9 ~ "Optional sensitivity",
      TRUE ~ "Core candidate"
    )
  ) |>
  arrange(subject, topic, desc(cv))

topic_decision_tbl <- kpi_decision_tbl |>
  group_by(subject, topic) |>
  summarise(
    avg_coverage_pct = mean(coverage_pct),
    core_metrics = sum(recommendation == "Core candidate"),
    redundant_metrics = sum(recommendation %in% c("Redundant - keep one lead KPI", "Keep, but de-duplicate")),
    optional_metrics = sum(recommendation == "Optional sensitivity"),
    dropped_metrics = sum(recommendation == "Drop"),
    lead_kpis = paste(metric_label[order(cv, decreasing = TRUE)][seq_len(min(3, n()))], collapse = "; "),
    .groups = "drop"
  ) |>
  left_join(topic_pca_tbl, by = c("subject", "topic")) |>
  mutate(
    topic_decision = case_when(
      topic == "Housing supply" ~ "Keep as multi-KPI topic",
      topic == "Health outcomes" ~ "Collapse to one lead KPI",
      topic == "Violence and injury" ~ "Keep as multi-KPI topic",
      topic %in% c("Walkability and transit baseline", "Food access baseline") ~ "Keep as baseline topic",
      topic == "Built-form proxy" ~ "Keep as proxy topic",
      topic %in% c("Air pollution", "Climate hazard risk") ~ "Keep with coverage caution",
      topic == "Hazard exposure" ~ "Sensitivity-only topic",
      core_metrics == 0 & optional_metrics == 0 & redundant_metrics == 0 ~ "Drop",
      core_metrics == 0 & optional_metrics > 0 ~ "Sensitivity-only topic",
      core_metrics == 1 & redundant_metrics == 0 ~ "Keep as one-KPI topic",
      components_80 <= 1 ~ "Collapse to one lead KPI",
      TRUE ~ "Keep as multi-KPI topic"
    )
  ) |>
  arrange(subject, topic)

subject_decision_tbl <- topic_decision_tbl |>
  group_by(subject) |>
  summarise(
    topic_count = n(),
    kept_topics = sum(topic_decision != "Drop"),
    strong_topics = sum(topic_decision %in% c("Keep as one-KPI topic", "Collapse to one lead KPI", "Keep as multi-KPI topic", "Keep as baseline topic", "Keep as proxy topic", "Keep with coverage caution")),
    avg_coverage_pct = mean(avg_coverage_pct, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    subject_decision = case_when(
      subject == "Physical Environment" ~ "Keep subject in Livability, but with coverage caution",
      subject == "Health & Safety" ~ "Keep subject in Livability",
      strong_topics == 0 ~ "Drop subject from Phase 1 core set",
      strong_topics == 1 ~ "Keep subject, but it is thin",
      TRUE ~ "Keep subject in Livability"
    )
  ) |>
  arrange(subject)

df_pca <- df |>
  select(all_of(surviving)) |>
  drop_na() |>
  scale()

pca_result <- prcomp(df_pca, center = FALSE, scale. = FALSE)
component_count_80 <- which(cumsum((pca_result$sdev^2) / sum(pca_result$sdev^2)) >= 0.8)[1]

recommended_bundle_tbl <- tibble::tribble(
  ~subject, ~topic, ~metric, ~bundle_tier,
  "Affordability", "Price pressure", "value_to_income", "core_recurring",
  "Affordability", "Housing burden", "pct_rent_burden_30plus", "core_recurring",
  "Affordability", "Poverty context", "pov_rate", "core_recurring",
  "Affordability", "Housing supply", "permits_per_1000_housing_units", "core_recurring",
  "Affordability", "Housing supply", "permits_share_units_5_plus", "core_recurring",
  "Affordability", "Housing structure mix", "pct_struct_mobile", "core_recurring",
  "Affordability", "Housing structure mix", "pct_struct_small_mf", "core_recurring",
  "Affordability", "Housing structure mix", "pct_struct_mid_mf", "core_recurring",
  "Health & Safety", "Health outcomes", "premature_death_rate", "core_recurring",
  "Health & Safety", "Health behavior and access", "mental_health_provider_ratio", "core_recurring",
  "Health & Safety", "Health behavior and access", "drug_overdose_death_rate", "core_recurring",
  "Health & Safety", "Health behavior and access", "pct_uninsured_adults", "core_recurring",
  "Health & Safety", "Health behavior and access", "preventable_hospital_stay_rate", "core_recurring",
  "Health & Safety", "Violence and injury", "firearm_fatality_rate", "core_recurring",
  "Health & Safety", "Violence and injury", "motor_vehicle_crash_rate", "core_recurring",
  "Access & Infrastructure", "Commute and mode access", "pct_commute_walk", "core_recurring",
  "Access & Infrastructure", "Commute and mode access", "pct_commute_wfh", "core_recurring",
  "Access & Infrastructure", "Housing slack", "vacancy_rate", "core_recurring",
  "Access & Infrastructure", "Vehicle access", "pct_hh_0_vehicles", "core_recurring",
  "Access & Infrastructure", "Digital access", "pct_no_internet_access", "core_recurring",
  "Access & Infrastructure", "Walkability and transit baseline", "walkability_index", "supplemental_baseline",
  "Access & Infrastructure", "Walkability and transit baseline", "jobs_access_45min_transit", "supplemental_baseline",
  "Access & Infrastructure", "Food access baseline", "pct_population_low_income_low_access_1_10", "supplemental_baseline",
  "Access & Infrastructure", "Built-form proxy", "pop_weighted_density_sqmi", "supplemental_baseline",
  "Physical Environment", "Air pollution", "unhealthy_days", "supplemental_caution",
  "Physical Environment", "Climate hazard risk", "fema_risk_score", "supplemental_caution"
) |>
  mutate(metric_label = unname(metric_labels[metric]))

subject_kpi_count_tbl <- recommended_bundle_tbl |>
  count(subject, bundle_tier, name = "kpi_count") |>
  pivot_wider(names_from = bundle_tier, values_from = kpi_count, values_fill = 0) |>
  mutate(total = core_recurring + supplemental_baseline + supplemental_caution) |>
  arrange(subject)

topic_weight_tbl <- recommended_bundle_tbl |>
  distinct(subject, topic, bundle_tier) |>
  left_join(topic_decision_tbl |> select(subject, topic, avg_coverage_pct), by = c("subject", "topic")) |>
  mutate(
    reliability_factor = case_when(
      bundle_tier == "core_recurring" ~ 1.0,
      bundle_tier == "supplemental_baseline" ~ 0.75,
      TRUE ~ 0.60
    ),
    raw_topic_weight = avg_coverage_pct * reliability_factor
  ) |>
  group_by(subject) |>
  mutate(
    topic_weight_within_subject = raw_topic_weight / sum(raw_topic_weight),
    subject_weight_equal = if_else(subject %in% c("Affordability", "Health & Safety", "Access & Infrastructure", "Physical Environment"), 0.25, 0),
    total_frame_weight = subject_weight_equal * topic_weight_within_subject
  ) |>
  ungroup() |>
  arrange(subject, desc(total_frame_weight), topic)

topic_lead_metrics_tbl <- tibble::tribble(
  ~topic, ~lead_metric, ~lead_label,
  "Price pressure", "value_to_income", "Home value-to-income ratio",
  "Housing burden", "pct_rent_burden_30plus", "Rent burden 30%+",
  "Poverty context", "pov_rate", "Poverty rate",
  "Housing supply", "permits_per_1000_housing_units", "Permits per 1k housing units",
  "Housing structure mix", "pct_struct_small_mf", "Small multifamily share",
  "Health outcomes", "premature_death_rate", "Premature death rate",
  "Health behavior and access", "mental_health_provider_ratio", "Mental health provider ratio",
  "Violence and injury", "firearm_fatality_rate", "Firearm fatality rate",
  "Commute and mode access", "pct_commute_walk", "Walk commute share",
  "Vehicle access", "pct_hh_0_vehicles", "Zero-vehicle household share",
  "Housing slack", "vacancy_rate", "Vacancy rate",
  "Digital access", "pct_no_internet_access", "No-internet share",
  "Walkability and transit baseline", "walkability_index", "Walkability index",
  "Food access baseline", "pct_population_low_income_low_access_1_10", "Low-income low-access population share",
  "Built-form proxy", "pop_weighted_density_sqmi", "Population-weighted density",
  "Air pollution", "unhealthy_days", "Unhealthy AQI days",
  "Climate hazard risk", "fema_risk_score", "FEMA composite risk score"
)

topic_lead_df <- df |>
  select(all_of(topic_lead_metrics_tbl$lead_metric)) |>
  setNames(topic_lead_metrics_tbl$topic)

topic_lead_cor_matrix <- cor(topic_lead_df, use = "pairwise.complete.obs")

topic_cross_check_tbl <- tibble::tribble(
  ~question, ~metric_a, ~metric_b,
  "Price pressure vs housing burden", "value_to_income", "pct_rent_burden_30plus",
  "Price pressure vs poverty", "value_to_income", "pov_rate",
  "Housing burden vs poverty", "pct_rent_burden_30plus", "pov_rate",
  "Housing supply rate vs density supply", "permits_per_1000_housing_units", "permits_share_units_5_plus",
  "Housing supply vs housing structure", "permits_per_1000_housing_units", "pct_struct_small_mf",
  "Built form vs vacancy", "pop_weighted_density_sqmi", "vacancy_rate",
  "Built form vs small multifamily", "pop_weighted_density_sqmi", "pct_struct_small_mf",
  "Built form vs mid multifamily", "pop_weighted_density_sqmi", "pct_struct_mid_mf",
  "Walkability vs built form", "walkability_index", "pop_weighted_density_sqmi",
  "Walkability vs walk commute", "walkability_index", "pct_commute_walk",
  "Food access vs no internet", "pct_population_low_income_low_access_1_10", "pct_no_internet_access",
  "Firearm fatalities vs premature death", "firearm_fatality_rate", "premature_death_rate",
  "Motor vehicle fatalities vs premature death", "motor_vehicle_crash_rate", "premature_death_rate",
  "Homicide vs premature death", "homicide_rate", "premature_death_rate"
) |>
  rowwise() |>
  mutate(
    n = sum(complete.cases(df[, c(metric_a, metric_b)])),
    correlation = cor(df[[metric_a]], df[[metric_b]], use = "complete.obs"),
    metric_a_label = unname(metric_labels[metric_a]),
    metric_b_label = unname(metric_labels[metric_b])
  ) |>
  ungroup()
```

## Source-family audit

```{r source-family-audit}
source_family_tbl |>
  transmute(
    `Source family` = source_family,
    `Latest year used` = latest_year,
    `Representative metric` = unname(metric_labels[coverage_metric]),
    `CBSAs with data` = coverage_n,
    `Coverage` = percent(coverage_pct, accuracy = 0.1),
    Note = note
  ) |>
  kable()
```

```{r inventory-audit}
topic_inventory_notes |>
  kable()
```

**Read of the audit**

- The recurring Livability frame is strongest in `Affordability`, `Access & Infrastructure`, and the main `CHR health` surface.
- `RPP`, `AQI`, `EJScreen`, and `FEMA` are usable, but each carries real coverage limits at CBSA scale.
- `HUD CHAS` is best treated as context rather than a default recurring family.
- `EPA SLD` and `USDA food access` are single-vintage baselines, but they are still strong enough to keep as explicit Livability topics.
- `K-12 learning quality` remains deferred and is not silently substituted here.

## Subject and topic inventory

```{r subject-topic-inventory}
metric_catalog |>
  count(subject, topic, role, name = "metric_count") |>
  arrange(subject, topic, role) |>
  kable()
```

## Coverage by topic

```{r coverage-summary}
subject_topic_coverage_tbl <- coverage_tbl |>
  group_by(subject, topic) |>
  summarise(
    metric_count = n(),
    min_coverage_pct = min(coverage_pct),
    avg_coverage_pct = mean(coverage_pct),
    .groups = "drop"
  ) |>
  arrange(subject, topic)

subject_topic_coverage_tbl |>
  mutate(
    min_coverage_pct = percent(min_coverage_pct, accuracy = 0.1),
    avg_coverage_pct = percent(avg_coverage_pct, accuracy = 0.1)
  ) |>
  kable()
```

```{r coverage-plot}
subject_topic_coverage_tbl |>
  mutate(topic = reorder(topic, avg_coverage_pct)) |>
  ggplot(aes(x = avg_coverage_pct, y = topic, fill = subject)) +
  geom_col(alpha = 0.9) +
  geom_vline(
    xintercept = 0.9,
    linetype = "22",
    color = visual_neutral_palette()$text_muted,
    linewidth = 0.5
  ) +
  scale_x_continuous(labels = label_percent()) +
  labs(
    title = "Coverage by Livability topic",
    subtitle = "Average KPI coverage within each topic | Dashed line = 90% caution threshold",
    x = "Average coverage",
    y = NULL,
    fill = NULL,
    caption = build_chart_notes(
      source = "ACS, HUD CHAS, BPS, CHR, EPA AQI/EJScreen/SLD, FEMA, USDA",
      vintage = "Latest available by source family"
    )
  ) +
  visual_theme()
```

## 1. KPI variance pass

```{r variance-summary}
variance_tbl |>
  transmute(
    subject,
    topic,
    metric = metric_label,
    role,
    coverage = percent(coverage_pct, accuracy = 0.1),
    cv = round(cv, 3),
    iqr = round(iqr, 3)
  ) |>
  arrange(subject, topic, desc(cv)) |>
  kable()
```

```{r variance-plot, fig.height=18}
variance_tbl |>
  mutate(metric_label = reorder(metric_label, cv)) |>
  ggplot(aes(x = cv, y = metric_label, fill = subject)) +
  geom_col(alpha = 0.85) +
  geom_vline(
    xintercept = 0.2,
    linetype = "22",
    color = visual_neutral_palette()$text_muted,
    linewidth = 0.5
  ) +
  labs(
    title = "Livability metrics ranked by coefficient of variation",
    subtitle = "Higher CV means a metric separates CBSAs more strongly",
    x = "Coefficient of variation (SD / mean)",
    y = NULL,
    fill = NULL,
    caption = build_chart_notes(
      source = "ACS, HUD CHAS, BPS, CHR, EPA AQI/EJScreen/SLD, FEMA, USDA",
      vintage = "Latest available by source family"
    )
  ) +
  visual_theme()
```

## 2. Topic scorecard

```{r topic-scorecard}
topic_scorecard_tbl |>
  mutate(
    avg_coverage_pct = percent(avg_coverage_pct, accuracy = 0.1)
  ) |>
  kable()
```

## 3. Correlation pass

```{r correlation-summary}
high_cor_tbl |>
  transmute(
    `Subject / topic A` = paste(subject_a, topic_a, sep = " / "),
    `Metric A` = metric_a_label,
    `Subject / topic B` = paste(subject_b, topic_b, sep = " / "),
    `Metric B` = metric_b_label,
    `r` = round(correlation, 2)
  ) |>
  kable()
```

```{r correlation-plot}
high_cor_tbl |>
  slice_head(n = 20) |>
  mutate(
    pair = paste(metric_a_label, metric_b_label, sep = " vs "),
    pair = reorder(pair, abs(correlation))
  ) |>
  ggplot(aes(x = correlation, y = pair, fill = correlation > 0)) +
  geom_col(alpha = 0.85) +
  scale_fill_manual(
    values = c(
      "TRUE" = visual_palette_defaults()$diverging$better,
      "FALSE" = visual_palette_defaults()$diverging$worse
    ),
    guide = "none"
  ) +
  labs(
    title = "Strongest Livability correlation pairs",
    subtitle = "Only pairs with |r| > 0.75 are shown",
    x = "Correlation",
    y = NULL,
    caption = build_chart_notes(
      source = "Latest-source-family Livability frame",
      vintage = "Mixed vintages by source family"
    )
  ) +
  visual_theme()
```

## 4. PCA diagnostic

```{r pca-scree}
tibble(
  component = seq_along(pca_result$sdev),
  variance_explained = (pca_result$sdev^2) / sum(pca_result$sdev^2)
) |>
  mutate(cumulative = cumsum(variance_explained)) |>
  ggplot(aes(x = component, y = variance_explained)) +
  geom_col(fill = visual_palette_defaults()$highlight$selection, alpha = 0.85) +
  geom_line(
    aes(y = cumulative),
    color = visual_neutral_palette()$text_muted,
    linewidth = 0.6
  ) +
  geom_point(
    aes(y = cumulative),
    color = visual_neutral_palette()$text_muted,
    size = 2
  ) +
  scale_y_continuous(labels = label_percent()) +
  labs(
    title = "PCA scree plot - Livability metrics",
    subtitle = "Bars = variance per component | Line = cumulative variance explained",
    x = "Principal component",
    y = "Variance explained",
    caption = build_chart_notes(
      source = "Latest-source-family Livability frame",
      vintage = "Mixed vintages by source family"
    )
  ) +
  visual_theme()
```

```{r topic-pca}
topic_pca_tbl |>
  arrange(subject, topic) |>
  kable()
```

## 5. Cross-topic correlation checks

This is the main next-step check from the current Livability pass. The question is not just whether KPIs overlap within a topic, but whether our chosen topics are secretly repeating the same story across the frame.

```{r cross-topic-summary}
topic_cross_check_tbl |>
  transmute(
    Question = question,
    `Metric A` = metric_a_label,
    `Metric B` = metric_b_label,
    `Complete CBSAs` = n,
    `r` = round(correlation, 2)
  ) |>
  kable()
```

```{r cross-topic-heatmap, fig.width=12, fig.height=10}
topic_lead_cor_matrix |>
  as.data.frame() |>
  tibble::rownames_to_column("topic_a") |>
  pivot_longer(-topic_a, names_to = "topic_b", values_to = "r") |>
  ggplot(aes(x = topic_b, y = topic_a, fill = r)) +
  geom_tile(color = visual_neutral_palette()$background_white) +
  scale_fill_gradient2(
    low = visual_palette_defaults()$diverging$worse,
    mid = visual_palette_defaults()$diverging$midpoint,
    high = visual_palette_defaults()$diverging$better,
    midpoint = 0,
    limits = c(-1, 1),
    name = "r"
  ) +
  labs(
    title = "Cross-topic correlation matrix using lead Livability KPIs",
    subtitle = "Used to check overweighting before clustering or score design",
    x = NULL,
    y = NULL,
    caption = build_chart_notes(
      source = "Lead KPI per Livability topic",
      vintage = "Mixed vintages by source family"
    )
  ) +
  visual_theme(grid = "none") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Read of the cross-topic check**

- `Price pressure` and `Housing burden` are related, but not redundant: r = 0.47.
- `Price pressure` and `Poverty context` are weakly related: r = -0.16.
- `Housing burden` and `Poverty context` are also weakly related: r = 0.12.
- `Built-form proxy` is not overly redundant with `Housing slack` or the deeper housing-structure KPIs:
  - density vs vacancy: r = -0.30
  - density vs small multifamily: r = 0.28
  - density vs mid multifamily: r = 0.37
- `Walkability` overlaps with density, but not enough to treat them as the same topic:
  - walkability vs built form: r = 0.65
- `Safety` behaves much more like a health-adjacent family than an affordability or access family:
  - firearm fatalities vs premature death: r = 0.74
  - motor vehicle fatalities vs premature death: r = 0.79

That last point is the strongest case for using one consolidated `Health & Safety` subject in the default clustering or calibration design.

## 6. Weighting design

The cleanest way to avoid overweighting a subject is hierarchical weighting:

1. Standardize each KPI to a z-score.
2. Average KPIs into a **topic score**.
3. Average topic scores into a **subject score**.
4. Average subject scores into the final Livability score or clustering input bundle.

That means a subject with more KPIs does **not** automatically get more influence.

The proposed Livability bundle count is:

```{r subject-kpi-counts}
subject_kpi_count_tbl |>
  kable()
```

For the first modeling pass, use four active subjects with equal top-line weight:

- `Affordability = 0.25`
- `Health & Safety = 0.25`
- `Access & Infrastructure = 0.25`
- `Physical Environment = 0.25`

Inside each subject, weight topics by:

`raw_topic_weight = coverage_share * reliability_factor`

Where:

- recurring core topics use `reliability_factor = 1.00`
- supplemental baseline topics use `reliability_factor = 0.75`
- supplemental coverage-caution topics use `reliability_factor = 0.60`

Then normalize within each subject:

`topic_weight_within_subject = raw_topic_weight / sum(raw_topic_weight within subject)`

And apply the equal subject weight:

`final_topic_weight = subject_weight_equal * topic_weight_within_subject`

```{r topic-weight-table}
topic_weight_tbl |>
  transmute(
    subject,
    topic,
    bundle_tier,
    coverage = percent(avg_coverage_pct, accuracy = 0.1),
    reliability_factor = round(reliability_factor, 2),
    topic_weight_within_subject = round(topic_weight_within_subject, 3),
    total_frame_weight = round(total_frame_weight, 3)
  ) |>
  kable()
```

Within each topic, keep KPI weights simple:

- one-KPI topics: the lead KPI gets all topic weight
- multi-KPI recurring topics: split topic weight equally across the selected core KPIs
- supplemental baseline / caution topics: split equally across the selected supplemental KPIs

That gives a concrete formula:

`final_kpi_weight = final_topic_weight / number_of_selected_kpis_in_topic`

This is the best default because it is:

- mathematically clean
- easy to explain
- robust to uneven KPI counts across subjects
- flexible enough to include the supplemental baseline topics without letting them dominate

## 7. Recommendations by subject, topic, and KPI

```{r decision-tables}
subject_decision_tbl |>
  mutate(avg_coverage_pct = percent(avg_coverage_pct, accuracy = 0.1)) |>
  kable()

topic_decision_tbl |>
  mutate(avg_coverage_pct = percent(avg_coverage_pct, accuracy = 0.1)) |>
  kable()
```

```{r decisions, results='asis'}
for (subject_name in unique(metric_catalog$subject)) {
  subject_row <- subject_decision_tbl |>
    filter(subject == subject_name)

  cat(glue("<h3>{subject_name}</h3>\n"))
  cat(glue("<p><strong>Subject decision:</strong> {subject_row$subject_decision}<br><strong>Coverage read:</strong> average topic coverage = {percent(subject_row$avg_coverage_pct, accuracy = 0.1)}</p>\n"))

  subject_topics <- topic_decision_tbl |>
    filter(subject == subject_name)

  for (i in seq_len(nrow(subject_topics))) {
    topic_row <- subject_topics[i, ]
    topic_metrics <- kpi_decision_tbl |>
      filter(subject == subject_name, topic == topic_row$topic)

    core_kpis <- topic_metrics |>
      filter(recommendation == "Core candidate") |>
      pull(metric_label)

    redundant_kpis <- topic_metrics |>
      filter(recommendation %in% c("Redundant - keep one lead KPI", "Keep, but de-duplicate")) |>
      pull(metric_label)

    optional_kpis <- topic_metrics |>
      filter(recommendation == "Optional sensitivity") |>
      pull(metric_label)

    drop_kpis <- topic_metrics |>
      filter(recommendation == "Drop") |>
      pull(metric_label)

    cat(glue("<h4>{topic_row$topic}</h4>\n"))
    cat(glue("<p><strong>Topic decision:</strong> {topic_row$topic_decision}<br><strong>Lead KPIs by variance:</strong> {topic_row$lead_kpis}<br><strong>Topic evidence:</strong> coverage = {percent(topic_row$avg_coverage_pct, accuracy = 0.1)}; high-variance metrics = {topic_scorecard_tbl$high_variance_metrics[match(topic_row$topic, topic_scorecard_tbl$topic)]}; topic PCA read = {topic_row$topic_dimensionality}</p>\n"))

    if (length(core_kpis) > 0) {
      cat("<p><strong>Core KPIs:</strong></p>\n<ul>\n")
      for (metric_name in core_kpis) {
        cat(glue("<li>{metric_name}</li>\n"))
      }
      cat("</ul>\n")
    }

    if (length(redundant_kpis) > 0) {
      cat("<p><strong>Redundant / consolidate KPIs:</strong></p>\n<ul>\n")
      for (metric_name in redundant_kpis) {
        cat(glue("<li>{metric_name}</li>\n"))
      }
      cat("</ul>\n")
    }

    if (length(optional_kpis) > 0) {
      cat("<p><strong>Optional sensitivity / baseline KPIs:</strong></p>\n<ul>\n")
      for (metric_name in optional_kpis) {
        cat(glue("<li>{metric_name}</li>\n"))
      }
      cat("</ul>\n")
    }

    if (length(drop_kpis) > 0) {
      cat("<p><strong>Drop from default Phase 1 set:</strong></p>\n<ul>\n")
      for (metric_name in drop_kpis) {
        cat(glue("<li>{metric_name}</li>\n"))
      }
      cat("</ul>\n")
    }
  }
}

cat(glue("<p><strong>PCA read:</strong> <code>{component_count_80}</code> components explain at least 80% of the variance in the surviving Livability set.</p>\n"))
cat("<p><strong>Working interpretation:</strong></p>\n<ul>\n")
cat("<li>Livability is not one thing. The audit shows a stable recurring core, a real environment family we should keep in scope, and a set of useful baseline access topics that strengthen the frame.</li>\n")
cat("<li>The main reduction problem is redundancy inside affordability, housing supply, structure mix, and some health and air-pollution families.</li>\n")
cat("<li>The main caution problem is coverage and vintage mismatch, especially for RPP, AQI/EJScreen/FEMA, SLD, food access, and the more specialized CHR education metrics.</li>\n")
cat("<li>The main modeling note is to keep `Health & Safety` consolidated as one subject and to keep checking cross-topic overlap before clustering or score calibration.</li>\n")
cat("</ul>\n")
```

## Working read

This Livability pass should be read in order:

1. decide whether the **subject** is worth keeping in the frame
2. decide whether the **topic** is recurring, thin, redundant, baseline-only, or weak
3. decide which **KPI** earns the default slot once we account for coverage, variance, and overlap

That matters here because Livability is more uneven than Character:

- `Affordability` is deep and recurring, but it is also the most redundancy-prone subject.
- `Health & Safety` is broad and useful, but some mortality / access sub-KPIs are clearly thinner than the headline measures and the safety topic sits close to the main health-outcome line.
- `Access & Infrastructure` has a strong recurring ACS core plus useful single-vintage baseline topics that should stay visible even if they are weighted carefully.
- `Education access` is live but thin until better K-12 quality data exists.
- `Physical Environment` belongs in the frame and should stay in scope, but it has to be handled carefully because direct source coverage is partial and the CHR fallback fields are only proxies.

**Update `exploration/intelligence_framework/docs/metric_selections.md` after reviewing this notebook output.**

```{r cleanup, include=FALSE}
dbDisconnect(con)
```
