---
title: "Variable Selection - Opportunity"
subtitle: "Subject, topic, and KPI diagnostic for Opportunity frame inputs"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 3
    code-fold: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)

options(
  dplyr.summarise.inform = FALSE,
  knitr.kable.NA = ""
)

library(here)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(glue)
library(purrr)
library(stringr)
library(knitr)
library(ggrepel)
library(tibble)

source(here("exploration/intelligence_framework/R/utils.R"))
con <- db_connect()
```

```{r catalog, include=FALSE}
industry_family_labels <- c(
  ag_mining = "Ag / mining",
  construction = "Construction",
  manufacturing = "Manufacturing",
  wholesale = "Wholesale",
  retail = "Retail",
  transport_util = "Transport / utilities",
  information = "Information",
  finance_real = "Finance / real estate",
  professional = "Professional services",
  educ_health = "Education / health",
  arts_accomm_food = "Arts / accommodations / food",
  other_services = "Other services"
)

gdp_family_labels <- c(
  natural_resources = "Natural resources",
  manufacturing = "Manufacturing",
  construction = "Construction",
  trade = "Trade",
  transportation = "Transportation",
  information = "Information",
  fire = "Finance / insurance / real estate",
  professional = "Professional services",
  edu_health = "Education / health",
  leisure = "Leisure / hospitality",
  gov = "Government"
)

metric_labels <- c(
  income_pc_growth_1yr = "Income per-capita growth (1yr)",
  income_pc_growth_5yr = "Income per-capita growth (5yr)",
  income_pc_cagr_5yr = "Income per-capita CAGR (5yr)",
  qcew_total_covered_avg_wkly_wage = "Covered weekly wage",
  qcew_private_avg_wkly_wage = "Private weekly wage",
  pct_unemployment_rate = "Unemployment rate",
  unemployment_rate_change_1yr = "Unemployment-rate change (1yr)",
  lfpr = "Labor-force participation rate",
  lfpr_growth_5yr = "LFPR growth (5yr)",
  pov_rate = "Poverty rate",
  pov_rate_change_1yr = "Poverty-rate change (1yr)",
  pov_rate_change_5yr = "Poverty-rate change (5yr)",
  gini_index = "Gini index",
  economic_connectedness = "Economic connectedness",
  hpi_yoy_pct = "HPI YoY growth",
  hpi_5yr_pct = "HPI growth (5yr)",
  hpi_10yr_pct = "HPI growth (10yr)",
  zori_annual_avg = "ZORI annual average",
  zori_annual_avg_yoy_pct = "ZORI YoY growth",
  zori_december = "ZORI December level",
  zori_december_yoy_pct = "ZORI December YoY growth",
  pop_growth_1yr = "Population growth (1yr)",
  pop_growth_5yr = "Population growth (5yr)",
  pop_cagr_5yr = "Population CAGR (5yr)",
  irs_net_migration_rate = "IRS net migration rate",
  irs_net_agi = "IRS net AGI",
  irs_inflow_agi = "IRS inflow AGI",
  irs_outflow_agi = "IRS outflow AGI",
  permits_per_1000_housing_units = "Permits per 1k housing units",
  permits_per_1000_population = "Permits per 1k population",
  permits_share_multifam_units = "Multifamily permit share",
  permits_share_units_5_plus = "5+ unit permit share",
  permits_avg_units_per_bldg = "Average units per building",
  real_gdp_growth_5yr = "Real GDP growth (5yr)",
  real_gdp_cagr_5yr = "Real GDP CAGR (5yr)",
  real_gdp_pc_growth_5yr = "Real GDP per-capita growth (5yr)",
  productivity_growth_5yr = "Productivity growth (5yr)",
  industry_concentration_hhi = "Industry concentration HHI",
  bfs_business_applications = "Business applications",
  bfs_business_applications_yoy_pct = "Business applications YoY",
  bfs_business_application_rate_per_1000_establishments = "Business applications per 1k establishments",
  bfs_business_applications_per_1000_residents = "Business applications per 1k residents",
  cbp_total_estabs = "Total establishments",
  cbp_estabs_per_1000_residents = "Establishments per 1k residents",
  pct_ba_plus = "BA+ share",
  pct_ba_plus_change_5yr = "BA+ share change (5yr)"
)

metric_labels <- c(
  metric_labels,
  setNames(
    paste0(unname(industry_family_labels), " employment share"),
    paste0("pct_qcew_private_emp_", names(industry_family_labels))
  ),
  setNames(
    paste0(unname(gdp_family_labels), " GDP share"),
    paste0("pct_real_gdp_", names(gdp_family_labels))
  ),
  setNames(
    paste0(unname(industry_family_labels), " weekly wage"),
    paste0("qcew_private_avg_wkly_wage_", names(industry_family_labels))
  ),
  setNames(
    paste0(unname(industry_family_labels), " establishment share"),
    paste0("pct_cbp_estabs_", names(industry_family_labels))
  ),
  setNames(
    paste0(unname(industry_family_labels), " location quotient"),
    paste0("lq_", names(industry_family_labels))
  )
)

resident_topics <- tribble(
  ~metric, ~subject, ~topic, ~source_family, ~role, ~note,
  "income_pc_growth_1yr", "Resident Opportunity", "Income growth", "BEA CAINC growth", "direct", "Short-run resident momentum.",
  "income_pc_growth_5yr", "Resident Opportunity", "Income growth", "BEA CAINC growth", "direct", "Structural resident momentum.",
  "income_pc_cagr_5yr", "Resident Opportunity", "Income growth", "BEA CAINC growth", "direct", "Alternative scaling of 5-year income growth.",
  "qcew_total_covered_avg_wkly_wage", "Resident Opportunity", "Wage levels", "QCEW wages", "direct", "Broad employer-side wage level.",
  "qcew_private_avg_wkly_wage", "Resident Opportunity", "Wage levels", "QCEW wages", "direct", "Private-sector wage level.",
  "pct_unemployment_rate", "Resident Opportunity", "Labor market tightness", "ACS labor", "direct", "Headline unemployment rate.",
  "unemployment_rate_change_1yr", "Resident Opportunity", "Labor market tightness", "ACS labor", "direct", "Short-run labor cooling / tightening.",
  "lfpr", "Resident Opportunity", "Labor market tightness", "ACS labor", "direct", "Labor-force attachment level.",
  "lfpr_growth_5yr", "Resident Opportunity", "Labor market tightness", "ACS labor", "direct", "Longer-run labor-force momentum.",
  "pov_rate", "Resident Opportunity", "Poverty and inclusion", "ACS income distribution", "direct", "Current resident distress level.",
  "pov_rate_change_1yr", "Resident Opportunity", "Poverty and inclusion", "ACS income distribution", "direct", "Short-run poverty shift.",
  "pov_rate_change_5yr", "Resident Opportunity", "Poverty and inclusion", "ACS income distribution", "direct", "Long-run poverty shift.",
  "gini_index", "Resident Opportunity", "Poverty and inclusion", "ACS income distribution", "direct", "Inclusive-vs-concentrated growth check.",
  "economic_connectedness", "Resident Opportunity", "Intergenerational mobility proxy", "Social Capital Atlas", "proxy", "Opportunity Atlas is still deferred; use as the live mobility proxy."
)

market_topics <- tribble(
  ~metric, ~subject, ~topic, ~source_family, ~role, ~note,
  "hpi_yoy_pct", "Market / Investor Opportunity", "Home price appreciation", "FHFA HPI", "direct", "Short-run home price momentum.",
  "hpi_5yr_pct", "Market / Investor Opportunity", "Home price appreciation", "FHFA HPI", "direct", "Structural housing-market momentum.",
  "hpi_10yr_pct", "Market / Investor Opportunity", "Home price appreciation", "FHFA HPI", "optional", "Longer-run housing-cycle context.",
  "zori_annual_avg", "Market / Investor Opportunity", "Rent growth", "Zillow ZORI", "baseline", "Rent level context; not a pure momentum KPI.",
  "zori_annual_avg_yoy_pct", "Market / Investor Opportunity", "Rent growth", "Zillow ZORI", "direct", "Recurring annual rent momentum.",
  "zori_december", "Market / Investor Opportunity", "Rent growth", "Zillow ZORI", "baseline", "Year-end rent level context.",
  "zori_december_yoy_pct", "Market / Investor Opportunity", "Rent growth", "Zillow ZORI", "optional", "Year-end momentum variant.",
  "pop_growth_1yr", "Market / Investor Opportunity", "Population growth", "ACS population", "direct", "Short-run population momentum.",
  "pop_growth_5yr", "Market / Investor Opportunity", "Population growth", "ACS population", "direct", "Structural growth signal.",
  "pop_cagr_5yr", "Market / Investor Opportunity", "Population growth", "ACS population", "direct", "Alternative scaling of 5-year population growth.",
  "irs_net_migration_rate", "Market / Investor Opportunity", "Migration and wealth flows", "IRS migration", "direct", "Demand-side migration pressure.",
  "irs_net_agi", "Market / Investor Opportunity", "Migration and wealth flows", "IRS migration", "direct", "Wealth-flow signal.",
  "irs_inflow_agi", "Market / Investor Opportunity", "Migration and wealth flows", "IRS migration", "direct", "Gross in-migration wealth.",
  "irs_outflow_agi", "Market / Investor Opportunity", "Migration and wealth flows", "IRS migration", "direct", "Gross out-migration wealth.",
  "permits_per_1000_housing_units", "Market / Investor Opportunity", "Permit activity", "BPS permits", "direct", "Forward supply intensity.",
  "permits_per_1000_population", "Market / Investor Opportunity", "Permit activity", "BPS permits", "direct", "Alternative permit denominator.",
  "permits_share_multifam_units", "Market / Investor Opportunity", "Permit activity", "BPS permits", "direct", "Supply composition toward multifamily.",
  "permits_share_units_5_plus", "Market / Investor Opportunity", "Permit activity", "BPS permits", "direct", "Higher-density supply composition.",
  "permits_avg_units_per_bldg", "Market / Investor Opportunity", "Permit activity", "BPS permits", "direct", "Scale of new building supply."
)

business_core_topics <- tribble(
  ~metric, ~subject, ~topic, ~source_family, ~role, ~note,
  "real_gdp_growth_5yr", "Business & Industry Opportunity", "GDP growth", "BEA GDP growth", "direct", "Headline structural output growth.",
  "real_gdp_cagr_5yr", "Business & Industry Opportunity", "GDP growth", "BEA GDP growth", "direct", "Alternative scaling of GDP growth.",
  "real_gdp_pc_growth_5yr", "Business & Industry Opportunity", "GDP growth", "BEA GDP growth", "direct", "Per-capita GDP growth.",
  "productivity_growth_5yr", "Business & Industry Opportunity", "GDP growth", "BEA GDP growth", "direct", "Productivity-focused growth lens.",
  "industry_concentration_hhi", "Business & Industry Opportunity", "Industry concentration", "BEA industry structure", "direct", "Diversification / concentration diagnostic.",
  "bfs_business_applications", "Business & Industry Opportunity", "Business formation", "BFS applications", "baseline", "Entrepreneurial flow level.",
  "bfs_business_applications_yoy_pct", "Business & Industry Opportunity", "Business formation", "BFS applications", "direct", "Short-run entrepreneurial momentum.",
  "bfs_business_application_rate_per_1000_establishments", "Business & Industry Opportunity", "Business formation", "BFS applications", "direct", "Business formation intensity vs current base.",
  "bfs_business_applications_per_1000_residents", "Business & Industry Opportunity", "Business formation", "BFS applications", "direct", "Population-scaled entrepreneurial intensity.",
  "cbp_total_estabs", "Business & Industry Opportunity", "Establishment density", "CBP establishments", "baseline", "Level context; strongly size-related.",
  "cbp_estabs_per_1000_residents", "Business & Industry Opportunity", "Establishment density", "CBP establishments", "direct", "Population-scaled establishment density.",
  "pct_ba_plus", "Business & Industry Opportunity", "Human capital momentum", "ACS education", "baseline", "Current human-capital level; primary home is Character.",
  "pct_ba_plus_change_5yr", "Business & Industry Opportunity", "Human capital momentum", "ACS education", "direct", "Human-capital accumulation signal."
)

metric_catalog <- bind_rows(
  resident_topics,
  market_topics,
  business_core_topics,
  tibble(
    metric = paste0("pct_qcew_private_emp_", names(industry_family_labels)),
    subject = "Business & Industry Opportunity",
    topic = "Sector employment mix",
    source_family = "QCEW sector employment shares",
    role = "direct",
    note = "Private-sector employment composition by broad industry family."
  ),
  tibble(
    metric = paste0("pct_real_gdp_", names(gdp_family_labels)),
    subject = "Business & Industry Opportunity",
    topic = "Sector GDP mix",
    source_family = "BEA sector GDP shares",
    role = "direct",
    note = "Output composition by broad industry family."
  ),
  tibble(
    metric = paste0("qcew_private_avg_wkly_wage_", names(industry_family_labels)),
    subject = "Business & Industry Opportunity",
    topic = "Sector wage mix",
    source_family = "QCEW sector wages",
    role = "direct",
    note = "Sector wage-quality diagnostic."
  ),
  tibble(
    metric = paste0("pct_cbp_estabs_", names(industry_family_labels)),
    subject = "Business & Industry Opportunity",
    topic = "Establishment mix",
    source_family = "CBP establishment shares",
    role = "direct",
    note = "Business-base composition by broad industry family."
  ),
  tibble(
    metric = paste0("lq_", names(industry_family_labels)),
    subject = "Business & Industry Opportunity",
    topic = "Location quotient specialization",
    source_family = "QCEW location quotients",
    role = "direct",
    note = "Industry specialization vs national benchmark."
  )
)

topic_inventory_notes <- tribble(
  ~subject, ~topic, ~status, ~note,
  "Resident Opportunity", "Intergenerational mobility proxy", "proxy", "Opportunity Atlas remains deferred. Phase 1 uses `economic_connectedness` as the live proxy and treats it as a sensitivity-aware topic rather than a perfect substitute.",
  "Market / Investor Opportunity", "Rent growth", "mixed direct + baseline", "ZORI gives live rent momentum, but rent levels are context fields rather than true trajectory KPIs and coverage is materially narrower than HPI.",
  "Business & Industry Opportunity", "Business formation", "live but vintage-limited", "BFS business-application levels continue into 2024, but the CBP-backed rate fields stop at 2023. Use the rate as the default KPI and name the vintage constraint explicitly.",
  "Business & Industry Opportunity", "Establishment density", "mixed direct + baseline", "Overall establishment density is usable, but raw establishment counts are size-driven and belong only as context.",
  "Business & Industry Opportunity", "Human capital momentum", "mixed direct + baseline", "The level field (`pct_ba_plus`) lives conceptually in Character; the 5-year change is the Opportunity-leading indicator."
)
```

## Goal

Build a complete Phase 1 Opportunity selection pass that is explicit at three levels:

1. **Subject**: should the subject stay in Opportunity at all?
2. **Topic**: should the topic stay, collapse to one KPI, remain a sensitivity-only topic, or drop?
3. **KPI**: which metrics are core, redundant, optional sensitivity tests, baseline-only context, proxies, or drops?

This notebook uses the same CBSA universe as the Character and Livability passes:

- `geo_level = 'cbsa'`
- `pop_total >= 100000`
- latest available usable row by **source family or source sub-family**

## Decision rules

These rules are used throughout the notebook and should be read as screens, not rigid auto-decisions:

- **Coverage**: below `90%` coverage is a caution flag; below `75%` is a strong warning.
- **Variance**: below `CV = 0.2` is weak differentiation at CBSA scale.
- **Correlation**: above `|r| = 0.75` means we should not carry every KPI at full weight inside the same topic.
- **PCA**: diagnostic only. It tells us whether a topic is really multi-dimensional after the basic screens.

## Phase 1 inventory

Opportunity is being treated as three main subjects plus one explicit proxy topic:

- `Resident Opportunity`
- `Market / Investor Opportunity`
- `Business & Industry Opportunity`

The audit is intentionally complete, so this notebook distinguishes:

- recurring direct KPIs
- baseline-only or level-context KPIs
- proxy KPIs
- deferred topics that still need to be named in the audit

```{r data}
df <- dbGetQuery(con, "
with spine as (
    select
        geo_id,
        geo_name,
        pop_total,
        year as spine_year
    from gold.population_demographics
    where geo_level = 'cbsa'
      and year = 2024
      and pop_total >= 100000
),
income_growth_latest as (
    select *
    from (
        select
            geo_id,
            year as income_growth_year,
            income_pc_growth_1yr,
            income_pc_growth_5yr,
            income_pc_cagr_5yr,
            row_number() over (
                partition by geo_id
                order by case
                    when income_pc_growth_1yr is not null
                      or income_pc_growth_5yr is not null
                      or income_pc_cagr_5yr is not null
                    then year
                end desc nulls last
            ) as rn
        from gold.economics_income_wide
        where geo_level = 'cbsa'
    )
    where rn = 1
),
income_distribution_latest as (
    select *
    from (
        select
            geo_id,
            year as income_distribution_year,
            pov_rate,
            pov_rate_change_1yr,
            pov_rate_change_5yr,
            gini_index,
            row_number() over (
                partition by geo_id
                order by case
                    when pov_rate is not null
                      or pov_rate_change_1yr is not null
                      or pov_rate_change_5yr is not null
                      or gini_index is not null
                    then year
                end desc nulls last
            ) as rn
        from gold.economics_income_wide
        where geo_level = 'cbsa'
    )
    where rn = 1
),
labor_latest as (
    select *
    from (
        select
            geo_id,
            year as labor_year,
            pct_unemployment_rate,
            unemployment_rate_change_1yr,
            lfpr,
            lfpr_growth_5yr,
            row_number() over (
                partition by geo_id
                order by case
                    when pct_unemployment_rate is not null
                      or unemployment_rate_change_1yr is not null
                      or lfpr is not null
                      or lfpr_growth_5yr is not null
                    then year
                end desc nulls last
            ) as rn
        from gold.economics_labor_wide
        where geo_level = 'cbsa'
    )
    where rn = 1
),
population_latest as (
    select *
    from (
        select
            geo_id,
            year as population_year,
            pop_growth_1yr,
            pop_growth_5yr,
            pop_cagr_5yr,
            pct_ba_plus,
            pct_ba_plus_change_5yr,
            row_number() over (
                partition by geo_id
                order by year desc
            ) as rn
        from gold.population_demographics
        where geo_level = 'cbsa'
    )
    where rn = 1
),
migration_latest as (
    select *
    from (
        select
            geo_id,
            year as migration_year,
            irs_net_migration_rate,
            irs_net_agi,
            irs_inflow_agi,
            irs_outflow_agi,
            row_number() over (
                partition by geo_id
                order by case
                    when irs_net_migration_rate is not null
                      or irs_net_agi is not null
                      or irs_inflow_agi is not null
                      or irs_outflow_agi is not null
                    then year
                end desc nulls last
            ) as rn
        from gold.migration_wide
        where geo_level = 'cbsa'
    )
    where rn = 1
),
housing_hpi_latest as (
    select *
    from (
        select
            geo_id,
            year as housing_hpi_year,
            hpi_yoy_pct,
            hpi_5yr_pct,
            hpi_10yr_pct,
            row_number() over (
                partition by geo_id
                order by case
                    when hpi_yoy_pct is not null
                      or hpi_5yr_pct is not null
                      or hpi_10yr_pct is not null
                    then year
                end desc nulls last
            ) as rn
        from gold.housing_market_wide
        where geo_level = 'cbsa'
    )
    where rn = 1
),
housing_zori_latest as (
    select *
    from (
        select
            geo_id,
            year as housing_zori_year,
            zori_annual_avg,
            zori_annual_avg_yoy_pct,
            zori_december,
            zori_december_yoy_pct,
            row_number() over (
                partition by geo_id
                order by case
                    when zori_annual_avg is not null
                      or zori_annual_avg_yoy_pct is not null
                      or zori_december is not null
                      or zori_december_yoy_pct is not null
                    then year
                end desc nulls last
            ) as rn
        from gold.housing_market_wide
        where geo_level = 'cbsa'
    )
    where rn = 1
),
housing_supply_latest as (
    select *
    from (
        select
            geo_id,
            year as housing_supply_year,
            permits_per_1000_housing_units,
            permits_per_1000_population,
            permits_share_multifam_units,
            permits_share_units_5_plus,
            permits_avg_units_per_bldg,
            row_number() over (
                partition by geo_id
                order by case
                    when permits_per_1000_housing_units is not null
                      or permits_per_1000_population is not null
                      or permits_share_multifam_units is not null
                      or permits_share_units_5_plus is not null
                      or permits_avg_units_per_bldg is not null
                    then year
                end desc nulls last
            ) as rn
        from gold.housing_core_wide
        where geo_level = 'cbsa'
    )
    where rn = 1
),
gdp_growth_latest as (
    select *
    from (
        select
            geo_id,
            year as gdp_growth_year,
            real_gdp_growth_5yr,
            real_gdp_cagr_5yr,
            real_gdp_pc_growth_5yr,
            productivity_growth_5yr,
            row_number() over (
                partition by geo_id
                order by case
                    when real_gdp_growth_5yr is not null
                      or real_gdp_cagr_5yr is not null
                      or real_gdp_pc_growth_5yr is not null
                      or productivity_growth_5yr is not null
                    then year
                end desc nulls last
            ) as rn
        from gold.economics_gdp_wide
        where geo_level = 'cbsa'
    )
    where rn = 1
),
industry_bea_latest as (
    select *
    from (
        select
            geo_id,
            year as industry_bea_year,
            industry_concentration_hhi,
            pct_real_gdp_natural_resources,
            pct_real_gdp_manufacturing,
            pct_real_gdp_construction,
            pct_real_gdp_trade,
            pct_real_gdp_transportation,
            pct_real_gdp_information,
            pct_real_gdp_fire,
            pct_real_gdp_professional,
            pct_real_gdp_edu_health,
            pct_real_gdp_leisure,
            pct_real_gdp_gov,
            row_number() over (
                partition by geo_id
                order by case
                    when industry_concentration_hhi is not null
                      or pct_real_gdp_professional is not null
                      or pct_real_gdp_manufacturing is not null
                    then year
                end desc nulls last
            ) as rn
        from gold.economics_industry_wide
        where geo_level = 'cbsa'
    )
    where rn = 1
),
industry_qcew_latest as (
    select *
    from (
        select
            geo_id,
            year as industry_qcew_year,
            qcew_total_covered_avg_wkly_wage,
            qcew_private_avg_wkly_wage,
            qcew_private_avg_wkly_wage_ag_mining,
            qcew_private_avg_wkly_wage_construction,
            qcew_private_avg_wkly_wage_manufacturing,
            qcew_private_avg_wkly_wage_wholesale,
            qcew_private_avg_wkly_wage_retail,
            qcew_private_avg_wkly_wage_transport_util,
            qcew_private_avg_wkly_wage_information,
            qcew_private_avg_wkly_wage_finance_real,
            qcew_private_avg_wkly_wage_professional,
            qcew_private_avg_wkly_wage_educ_health,
            qcew_private_avg_wkly_wage_arts_accomm_food,
            qcew_private_avg_wkly_wage_other_services,
            pct_qcew_private_emp_ag_mining,
            pct_qcew_private_emp_construction,
            pct_qcew_private_emp_manufacturing,
            pct_qcew_private_emp_wholesale,
            pct_qcew_private_emp_retail,
            pct_qcew_private_emp_transport_util,
            pct_qcew_private_emp_information,
            pct_qcew_private_emp_finance_real,
            pct_qcew_private_emp_professional,
            pct_qcew_private_emp_educ_health,
            pct_qcew_private_emp_arts_accomm_food,
            pct_qcew_private_emp_other_services,
            lq_ag_mining,
            lq_construction,
            lq_manufacturing,
            lq_wholesale,
            lq_retail,
            lq_transport_util,
            lq_information,
            lq_finance_real,
            lq_professional,
            lq_educ_health,
            lq_arts_accomm_food,
            lq_other_services,
            row_number() over (
                partition by geo_id
                order by case
                    when pct_qcew_private_emp_professional is not null
                      or qcew_private_avg_wkly_wage_professional is not null
                      or lq_professional is not null
                    then year
                end desc nulls last
            ) as rn
        from gold.economics_industry_wide
        where geo_level = 'cbsa'
    )
    where rn = 1
),
industry_bfs_flow_latest as (
    select *
    from (
        select
            geo_id,
            year as industry_bfs_flow_year,
            bfs_business_applications,
            bfs_business_applications_yoy_pct,
            bfs_business_applications_per_1000_residents,
            row_number() over (
                partition by geo_id
                order by case
                    when bfs_business_applications is not null
                      or bfs_business_applications_yoy_pct is not null
                      or bfs_business_applications_per_1000_residents is not null
                    then year
                end desc nulls last
            ) as rn
        from gold.economics_industry_wide
        where geo_level = 'cbsa'
    )
    where rn = 1
),
industry_bfs_rate_latest as (
    select *
    from (
        select
            geo_id,
            year as industry_bfs_rate_year,
            bfs_business_application_rate_per_1000_establishments,
            row_number() over (
                partition by geo_id
                order by case
                    when bfs_business_application_rate_per_1000_establishments is not null
                    then year
                end desc nulls last
            ) as rn
        from gold.economics_industry_wide
        where geo_level = 'cbsa'
    )
    where rn = 1
),
industry_cbp_latest as (
    select *
    from (
        select
            geo_id,
            year as industry_cbp_year,
            cbp_total_estabs,
            cbp_estabs_per_1000_residents,
            pct_cbp_estabs_ag_mining,
            pct_cbp_estabs_construction,
            pct_cbp_estabs_manufacturing,
            pct_cbp_estabs_wholesale,
            pct_cbp_estabs_retail,
            pct_cbp_estabs_transport_util,
            pct_cbp_estabs_information,
            pct_cbp_estabs_finance_real,
            pct_cbp_estabs_professional,
            pct_cbp_estabs_educ_health,
            pct_cbp_estabs_arts_accomm_food,
            pct_cbp_estabs_other_services,
            row_number() over (
                partition by geo_id
                order by case
                    when cbp_estabs_per_1000_residents is not null
                      or pct_cbp_estabs_professional is not null
                      or cbp_total_estabs is not null
                    then year
                end desc nulls last
            ) as rn
        from gold.economics_industry_wide
        where geo_level = 'cbsa'
    )
    where rn = 1
),
social_latest as (
    select
        geo_id,
        geo_name as social_geo_name,
        economic_connectedness
    from gold.social_fabric_wide
    where geo_level = 'cbsa'
)
select
    spine.*,
    income_growth.income_growth_year,
    income_growth.income_pc_growth_1yr,
    income_growth.income_pc_growth_5yr,
    income_growth.income_pc_cagr_5yr,
    income_distribution.income_distribution_year,
    income_distribution.pov_rate,
    income_distribution.pov_rate_change_1yr,
    income_distribution.pov_rate_change_5yr,
    income_distribution.gini_index,
    labor.labor_year,
    labor.pct_unemployment_rate,
    labor.unemployment_rate_change_1yr,
    labor.lfpr,
    labor.lfpr_growth_5yr,
    population.population_year,
    population.pop_growth_1yr,
    population.pop_growth_5yr,
    population.pop_cagr_5yr,
    population.pct_ba_plus,
    population.pct_ba_plus_change_5yr,
    migration.migration_year,
    migration.irs_net_migration_rate,
    migration.irs_net_agi,
    migration.irs_inflow_agi,
    migration.irs_outflow_agi,
    housing_hpi.housing_hpi_year,
    housing_hpi.hpi_yoy_pct,
    housing_hpi.hpi_5yr_pct,
    housing_hpi.hpi_10yr_pct,
    housing_zori.housing_zori_year,
    housing_zori.zori_annual_avg,
    housing_zori.zori_annual_avg_yoy_pct,
    housing_zori.zori_december,
    housing_zori.zori_december_yoy_pct,
    housing_supply.housing_supply_year,
    housing_supply.permits_per_1000_housing_units,
    housing_supply.permits_per_1000_population,
    housing_supply.permits_share_multifam_units,
    housing_supply.permits_share_units_5_plus,
    housing_supply.permits_avg_units_per_bldg,
    gdp_growth.gdp_growth_year,
    gdp_growth.real_gdp_growth_5yr,
    gdp_growth.real_gdp_cagr_5yr,
    gdp_growth.real_gdp_pc_growth_5yr,
    gdp_growth.productivity_growth_5yr,
    industry_bea.industry_bea_year,
    industry_bea.industry_concentration_hhi,
    industry_bea.pct_real_gdp_natural_resources,
    industry_bea.pct_real_gdp_manufacturing,
    industry_bea.pct_real_gdp_construction,
    industry_bea.pct_real_gdp_trade,
    industry_bea.pct_real_gdp_transportation,
    industry_bea.pct_real_gdp_information,
    industry_bea.pct_real_gdp_fire,
    industry_bea.pct_real_gdp_professional,
    industry_bea.pct_real_gdp_edu_health,
    industry_bea.pct_real_gdp_leisure,
    industry_bea.pct_real_gdp_gov,
    industry_qcew.industry_qcew_year,
    industry_qcew.qcew_total_covered_avg_wkly_wage,
    industry_qcew.qcew_private_avg_wkly_wage,
    industry_qcew.qcew_private_avg_wkly_wage_ag_mining,
    industry_qcew.qcew_private_avg_wkly_wage_construction,
    industry_qcew.qcew_private_avg_wkly_wage_manufacturing,
    industry_qcew.qcew_private_avg_wkly_wage_wholesale,
    industry_qcew.qcew_private_avg_wkly_wage_retail,
    industry_qcew.qcew_private_avg_wkly_wage_transport_util,
    industry_qcew.qcew_private_avg_wkly_wage_information,
    industry_qcew.qcew_private_avg_wkly_wage_finance_real,
    industry_qcew.qcew_private_avg_wkly_wage_professional,
    industry_qcew.qcew_private_avg_wkly_wage_educ_health,
    industry_qcew.qcew_private_avg_wkly_wage_arts_accomm_food,
    industry_qcew.qcew_private_avg_wkly_wage_other_services,
    industry_qcew.pct_qcew_private_emp_ag_mining,
    industry_qcew.pct_qcew_private_emp_construction,
    industry_qcew.pct_qcew_private_emp_manufacturing,
    industry_qcew.pct_qcew_private_emp_wholesale,
    industry_qcew.pct_qcew_private_emp_retail,
    industry_qcew.pct_qcew_private_emp_transport_util,
    industry_qcew.pct_qcew_private_emp_information,
    industry_qcew.pct_qcew_private_emp_finance_real,
    industry_qcew.pct_qcew_private_emp_professional,
    industry_qcew.pct_qcew_private_emp_educ_health,
    industry_qcew.pct_qcew_private_emp_arts_accomm_food,
    industry_qcew.pct_qcew_private_emp_other_services,
    industry_qcew.lq_ag_mining,
    industry_qcew.lq_construction,
    industry_qcew.lq_manufacturing,
    industry_qcew.lq_wholesale,
    industry_qcew.lq_retail,
    industry_qcew.lq_transport_util,
    industry_qcew.lq_information,
    industry_qcew.lq_finance_real,
    industry_qcew.lq_professional,
    industry_qcew.lq_educ_health,
    industry_qcew.lq_arts_accomm_food,
    industry_qcew.lq_other_services,
    industry_bfs_flow.industry_bfs_flow_year,
    industry_bfs_flow.bfs_business_applications,
    industry_bfs_flow.bfs_business_applications_yoy_pct,
    industry_bfs_rate.industry_bfs_rate_year,
    industry_bfs_rate.bfs_business_application_rate_per_1000_establishments,
    industry_bfs_flow.bfs_business_applications_per_1000_residents,
    industry_cbp.industry_cbp_year,
    industry_cbp.cbp_total_estabs,
    industry_cbp.cbp_estabs_per_1000_residents,
    industry_cbp.pct_cbp_estabs_ag_mining,
    industry_cbp.pct_cbp_estabs_construction,
    industry_cbp.pct_cbp_estabs_manufacturing,
    industry_cbp.pct_cbp_estabs_wholesale,
    industry_cbp.pct_cbp_estabs_retail,
    industry_cbp.pct_cbp_estabs_transport_util,
    industry_cbp.pct_cbp_estabs_information,
    industry_cbp.pct_cbp_estabs_finance_real,
    industry_cbp.pct_cbp_estabs_professional,
    industry_cbp.pct_cbp_estabs_educ_health,
    industry_cbp.pct_cbp_estabs_arts_accomm_food,
    industry_cbp.pct_cbp_estabs_other_services,
    social.economic_connectedness
from spine
left join income_growth_latest income_growth
    on spine.geo_id = income_growth.geo_id
left join income_distribution_latest income_distribution
    on spine.geo_id = income_distribution.geo_id
left join labor_latest labor
    on spine.geo_id = labor.geo_id
left join population_latest population
    on spine.geo_id = population.geo_id
left join migration_latest migration
    on spine.geo_id = migration.geo_id
left join housing_hpi_latest housing_hpi
    on spine.geo_id = housing_hpi.geo_id
left join housing_zori_latest housing_zori
    on spine.geo_id = housing_zori.geo_id
left join housing_supply_latest housing_supply
    on spine.geo_id = housing_supply.geo_id
left join gdp_growth_latest gdp_growth
    on spine.geo_id = gdp_growth.geo_id
left join industry_bea_latest industry_bea
    on spine.geo_id = industry_bea.geo_id
left join industry_qcew_latest industry_qcew
    on spine.geo_id = industry_qcew.geo_id
left join industry_bfs_flow_latest industry_bfs_flow
    on spine.geo_id = industry_bfs_flow.geo_id
left join industry_bfs_rate_latest industry_bfs_rate
    on spine.geo_id = industry_bfs_rate.geo_id
left join industry_cbp_latest industry_cbp
    on spine.geo_id = industry_cbp.geo_id
left join social_latest social
    on spine.geo_id = social.geo_id
order by spine.pop_total desc
")

metrics <- metric_catalog$metric

coverage_tbl <- tibble(
  metric = metrics,
  non_null_cbsas = sapply(metrics, \(m) sum(!is.na(df[[m]]))),
  coverage_pct = non_null_cbsas / nrow(df)
) |>
  left_join(metric_catalog, by = "metric") |>
  mutate(metric_label = unname(metric_labels[metric])) |>
  arrange(subject, topic, desc(coverage_pct), metric)

source_family_tbl <- tibble::tribble(
  ~source_family, ~latest_year, ~coverage_metric, ~coverage_n, ~note,
  "BEA CAINC growth", max(df$income_growth_year, na.rm = TRUE), "income_pc_growth_5yr", sum(!is.na(df$income_pc_growth_5yr)), "Growth fields require the latest non-null growth row, which is 2023 in the current Gold surface.",
  "ACS income distribution", max(df$income_distribution_year, na.rm = TRUE), "gini_index", sum(!is.na(df$gini_index)), "Inclusive-growth companion fields remain available on the latest ACS row.",
  "ACS labor", max(df$labor_year, na.rm = TRUE), "pct_unemployment_rate", sum(!is.na(df$pct_unemployment_rate)), "Labor-market levels and short-run changes on the latest ACS labor row.",
  "ACS population", max(df$population_year, na.rm = TRUE), "pop_growth_5yr", sum(!is.na(df$pop_growth_5yr)), "Population momentum and BA+ change live on the current population surface.",
  "IRS migration", max(df$migration_year, na.rm = TRUE), "irs_net_migration_rate", sum(!is.na(df$irs_net_migration_rate)), "IRS migration and AGI flows remain slightly partial at CBSA scale.",
  "FHFA HPI", max(df$housing_hpi_year, na.rm = TRUE), "hpi_5yr_pct", sum(!is.na(df$hpi_5yr_pct)), "Home-price momentum now uses the latest non-null HPI row on the Zillow market surface.",
  "Zillow ZORI", max(df$housing_zori_year, na.rm = TRUE), "zori_annual_avg_yoy_pct", sum(!is.na(df$zori_annual_avg_yoy_pct)), "Rent momentum is live but materially sparser than HPI coverage.",
  "BPS permits", max(df$housing_supply_year, na.rm = TRUE), "permits_per_1000_housing_units", sum(!is.na(df$permits_per_1000_housing_units)), "Forward supply metrics are fully live on the housing-core surface for this universe.",
  "BEA GDP growth", max(df$gdp_growth_year, na.rm = TRUE), "real_gdp_growth_5yr", sum(!is.na(df$real_gdp_growth_5yr)), "Growth fields require the latest non-null GDP-growth row, which is behind the 2024 spine.",
  "BEA sector GDP shares", max(df$industry_bea_year, na.rm = TRUE), "pct_real_gdp_professional", sum(!is.na(df$pct_real_gdp_professional)), "GDP composition and HHI come from the latest non-null BEA industry row.",
  "QCEW sector shares and wages", max(df$industry_qcew_year, na.rm = TRUE), "pct_qcew_private_emp_professional", sum(!is.na(df$pct_qcew_private_emp_professional)), "Employment mix, location quotients, and sector wages are all live in 2024.",
  "BFS application flow", max(df$industry_bfs_flow_year, na.rm = TRUE), "bfs_business_applications_yoy_pct", sum(!is.na(df$bfs_business_applications_yoy_pct)), "Business-application levels and YoY momentum remain live into 2024.",
  "BFS application rate", max(df$industry_bfs_rate_year, na.rm = TRUE), "bfs_business_application_rate_per_1000_establishments", sum(!is.na(df$bfs_business_application_rate_per_1000_establishments)), "The CBP-backed rate field stops where CBP overlap exists, so the usable latest row is 2023.",
  "CBP establishments", max(df$industry_cbp_year, na.rm = TRUE), "cbp_estabs_per_1000_residents", sum(!is.na(df$cbp_estabs_per_1000_residents)), "Establishment density and mix also stop at the last CBP-supported year.",
  "Social Capital Atlas", NA_integer_, "economic_connectedness", sum(!is.na(df$economic_connectedness)), "Static baseline source; coverage is broad enough to test as a proxy topic."
) |>
  mutate(coverage_pct = coverage_n / nrow(df))

variance_tbl <- tibble(metric = metrics) |>
  mutate(
    n = sapply(metric, \(m) sum(!is.na(df[[m]]))),
    mean = sapply(metric, \(m) mean(df[[m]], na.rm = TRUE)),
    sd = sapply(metric, \(m) sd(df[[m]], na.rm = TRUE)),
    cv = sd / abs(mean),
    iqr = sapply(metric, \(m) IQR(df[[m]], na.rm = TRUE))
  ) |>
  left_join(metric_catalog, by = "metric") |>
  left_join(coverage_tbl |> select(metric, coverage_pct), by = "metric") |>
  mutate(metric_label = unname(metric_labels[metric])) |>
  arrange(desc(cv))

topic_scorecard_tbl <- variance_tbl |>
  group_by(subject, topic) |>
  summarise(
    metric_count = n(),
    avg_coverage_pct = mean(coverage_pct),
    direct_metrics = sum(role == "direct"),
    baseline_metrics = sum(role == "baseline"),
    proxy_metrics = sum(role == "proxy"),
    optional_metrics = sum(role == "optional"),
    high_variance_metrics = sum(cv >= 0.2, na.rm = TRUE),
    top_kpis = paste(metric_label[order(cv, decreasing = TRUE)][seq_len(min(3, n()))], collapse = "; "),
    .groups = "drop"
  ) |>
  mutate(
    topic_read = case_when(
      high_variance_metrics == 0 ~ "Weak topic at CBSA scale",
      high_variance_metrics == 1 ~ "Keep topic, likely with one lead KPI",
      TRUE ~ "Keep topic, compare KPIs inside topic"
    )
  ) |>
  arrange(subject, topic)

surviving <- variance_tbl |>
  filter(cv >= 0.2, coverage_pct >= 0.75) |>
  pull(metric)

df_cor <- df |>
  select(all_of(surviving)) |>
  drop_na()

cor_matrix <- cor(df_cor, use = "pairwise.complete.obs")
high_cor <- which(abs(cor_matrix) > 0.75 & upper.tri(cor_matrix), arr.ind = TRUE)

high_cor_tbl <- if (nrow(high_cor) > 0) {
  tibble(
    metric_a = rownames(cor_matrix)[high_cor[, 1]],
    metric_b = colnames(cor_matrix)[high_cor[, 2]],
    correlation = cor_matrix[high_cor]
  ) |>
    mutate(
      metric_a_label = unname(metric_labels[metric_a]),
      metric_b_label = unname(metric_labels[metric_b])
    ) |>
    left_join(metric_catalog |> rename(metric_a = metric, subject_a = subject, topic_a = topic), by = "metric_a") |>
    left_join(metric_catalog |> rename(metric_b = metric, subject_b = subject, topic_b = topic), by = "metric_b") |>
    arrange(desc(abs(correlation)))
} else {
  tibble(
    metric_a = character(),
    metric_b = character(),
    correlation = double(),
    metric_a_label = character(),
    metric_b_label = character(),
    subject_a = character(),
    topic_a = character(),
    subject_b = character(),
    topic_b = character()
  )
}

topic_pca_tbl <- map_dfr(split(metric_catalog, metric_catalog$topic), \(topic_df) {
  topic_metrics <- variance_tbl |>
    filter(
      topic == topic_df$topic[[1]],
      coverage_pct >= 0.75,
      cv >= 0.2
    ) |>
    pull(metric)

  if (length(topic_metrics) == 0) {
    return(tibble(
      subject = topic_df$subject[[1]],
      topic = topic_df$topic[[1]],
      surviving_metrics = 0L,
      complete_rows = NA_integer_,
      components_80 = NA_integer_,
      topic_dimensionality = "No screened metrics survive"
    ))
  }

  topic_complete <- df |>
    select(all_of(topic_metrics)) |>
    drop_na()

  if (length(topic_metrics) < 2) {
    return(tibble(
      subject = topic_df$subject[[1]],
      topic = topic_df$topic[[1]],
      surviving_metrics = length(topic_metrics),
      complete_rows = nrow(topic_complete),
      components_80 = NA_integer_,
      topic_dimensionality = "Single-metric topic"
    ))
  }

  if (nrow(topic_complete) < 50) {
    return(tibble(
      subject = topic_df$subject[[1]],
      topic = topic_df$topic[[1]],
      surviving_metrics = length(topic_metrics),
      complete_rows = nrow(topic_complete),
      components_80 = NA_integer_,
      topic_dimensionality = "Too sparse for stable topic PCA"
    ))
  }

  topic_pca <- prcomp(scale(topic_complete), center = FALSE, scale. = FALSE)
  topic_components_80 <- which(cumsum((topic_pca$sdev^2) / sum(topic_pca$sdev^2)) >= 0.8)[1]

  tibble(
    subject = topic_df$subject[[1]],
    topic = topic_df$topic[[1]],
    surviving_metrics = length(topic_metrics),
    complete_rows = nrow(topic_complete),
    components_80 = topic_components_80,
    topic_dimensionality = case_when(
      topic_components_80 <= 1 ~ "Mostly one dominant KPI story",
      topic_components_80 == 2 ~ "Has two meaningful sub-dimensions",
      TRUE ~ "Multi-dimensional topic"
    )
  )
})

cross_topic_lead_tbl <- tribble(
  ~metric, ~subject, ~topic,
  "income_pc_growth_5yr", "Resident Opportunity", "Income growth",
  "qcew_private_avg_wkly_wage", "Resident Opportunity", "Wage levels",
  "pct_unemployment_rate", "Resident Opportunity", "Labor market tightness",
  "lfpr", "Resident Opportunity", "Labor market tightness",
  "pov_rate_change_5yr", "Resident Opportunity", "Poverty and inclusion",
  "gini_index", "Resident Opportunity", "Poverty and inclusion",
  "economic_connectedness", "Resident Opportunity", "Intergenerational mobility proxy",
  "hpi_5yr_pct", "Market / Investor Opportunity", "Home price appreciation",
  "hpi_yoy_pct", "Market / Investor Opportunity", "Home price appreciation",
  "zori_annual_avg_yoy_pct", "Market / Investor Opportunity", "Rent growth",
  "pop_growth_5yr", "Market / Investor Opportunity", "Population growth",
  "irs_net_migration_rate", "Market / Investor Opportunity", "Migration and wealth flows",
  "irs_net_agi", "Market / Investor Opportunity", "Migration and wealth flows",
  "permits_per_1000_housing_units", "Market / Investor Opportunity", "Permit activity",
  "permits_share_units_5_plus", "Market / Investor Opportunity", "Permit activity",
  "productivity_growth_5yr", "Business & Industry Opportunity", "GDP growth",
  "industry_concentration_hhi", "Business & Industry Opportunity", "Industry concentration",
  "bfs_business_application_rate_per_1000_establishments", "Business & Industry Opportunity", "Business formation",
  "cbp_estabs_per_1000_residents", "Business & Industry Opportunity", "Establishment density",
  "pct_ba_plus_change_5yr", "Business & Industry Opportunity", "Human capital momentum",
  "lq_professional", "Business & Industry Opportunity", "Location quotient specialization",
  "lq_information", "Business & Industry Opportunity", "Location quotient specialization",
  "lq_manufacturing", "Business & Industry Opportunity", "Location quotient specialization",
  "pct_real_gdp_information", "Business & Industry Opportunity", "Sector GDP mix",
  "pct_qcew_private_emp_manufacturing", "Business & Industry Opportunity", "Sector employment mix",
  "qcew_private_avg_wkly_wage_professional", "Business & Industry Opportunity", "Sector wage mix",
  "pct_cbp_estabs_information", "Business & Industry Opportunity", "Establishment mix"
)

cross_topic_cor_tbl <- {
  lead_metrics <- cross_topic_lead_tbl$metric
  lead_cor_matrix <- cor(
    df |>
      select(all_of(lead_metrics)),
    use = "pairwise.complete.obs"
  )

  lead_pairs <- which(abs(lead_cor_matrix) >= 0.6 & upper.tri(lead_cor_matrix), arr.ind = TRUE)

  if (nrow(lead_pairs) > 0) {
    tibble(
      metric_a = rownames(lead_cor_matrix)[lead_pairs[, 1]],
      metric_b = colnames(lead_cor_matrix)[lead_pairs[, 2]],
      correlation = lead_cor_matrix[lead_pairs]
    ) |>
      left_join(cross_topic_lead_tbl |> rename(metric_a = metric, subject_a = subject, topic_a = topic), by = "metric_a") |>
      left_join(cross_topic_lead_tbl |> rename(metric_b = metric, subject_b = subject, topic_b = topic), by = "metric_b") |>
      filter(topic_a != topic_b) |>
      mutate(
        metric_a_label = unname(metric_labels[metric_a]),
        metric_b_label = unname(metric_labels[metric_b])
      ) |>
      arrange(desc(abs(correlation)))
  } else {
    tibble(
      metric_a = character(),
      metric_b = character(),
      correlation = double(),
      subject_a = character(),
      topic_a = character(),
      subject_b = character(),
      topic_b = character(),
      metric_a_label = character(),
      metric_b_label = character()
    )
  }
}

business_topic_competitor_tbl <- tribble(
  ~metric, ~topic_family, ~cluster_shortlist,
  "lq_professional", "Location quotient specialization", TRUE,
  "lq_information", "Location quotient specialization", TRUE,
  "lq_manufacturing", "Location quotient specialization", TRUE,
  "pct_real_gdp_information", "Sector GDP mix", TRUE,
  "pct_real_gdp_professional", "Sector GDP mix", FALSE,
  "pct_real_gdp_manufacturing", "Sector GDP mix", FALSE,
  "pct_qcew_private_emp_manufacturing", "Sector employment mix", TRUE,
  "pct_qcew_private_emp_information", "Sector employment mix", FALSE,
  "pct_qcew_private_emp_professional", "Sector employment mix", FALSE,
  "qcew_private_avg_wkly_wage_professional", "Sector wage mix", TRUE,
  "qcew_private_avg_wkly_wage_information", "Sector wage mix", FALSE,
  "qcew_private_avg_wkly_wage_manufacturing", "Sector wage mix", FALSE,
  "pct_cbp_estabs_information", "Establishment mix", TRUE,
  "pct_cbp_estabs_professional", "Establishment mix", FALSE,
  "pct_cbp_estabs_manufacturing", "Establishment mix", FALSE
) |>
  left_join(
    variance_tbl |>
      select(metric, coverage_pct, cv, iqr) |>
      distinct(),
    by = "metric"
  ) |>
  mutate(metric_label = unname(metric_labels[metric]))

business_topic_competitor_cor_tbl <- {
  business_metrics <- business_topic_competitor_tbl$metric
  business_cor_matrix <- cor(
    df |>
      select(all_of(business_metrics)),
    use = "pairwise.complete.obs"
  )

  business_pairs <- which(abs(business_cor_matrix) >= 0.6 & upper.tri(business_cor_matrix), arr.ind = TRUE)

  if (nrow(business_pairs) > 0) {
    tibble(
      metric_a = rownames(business_cor_matrix)[business_pairs[, 1]],
      metric_b = colnames(business_cor_matrix)[business_pairs[, 2]],
      correlation = business_cor_matrix[business_pairs]
    ) |>
      left_join(business_topic_competitor_tbl |> rename(metric_a = metric, family_a = topic_family), by = "metric_a") |>
      left_join(business_topic_competitor_tbl |> rename(metric_b = metric, family_b = topic_family), by = "metric_b") |>
      mutate(
        metric_a_label = unname(metric_labels[metric_a]),
        metric_b_label = unname(metric_labels[metric_b])
      ) |>
      arrange(desc(abs(correlation)))
  } else {
    tibble(
      metric_a = character(),
      metric_b = character(),
      correlation = double(),
      family_a = character(),
      family_b = character(),
      metric_a_label = character(),
      metric_b_label = character()
    )
  }
}

business_shortlist_metrics <- business_topic_competitor_tbl |>
  filter(cluster_shortlist) |>
  pull(metric)

business_shortlist_complete <- df |>
  select(all_of(business_shortlist_metrics)) |>
  drop_na()

business_shortlist_pca_tbl <- if (nrow(business_shortlist_complete) >= 50) {
  business_shortlist_pca <- prcomp(scale(business_shortlist_complete), center = FALSE, scale. = FALSE)

  tibble(
    metric = rownames(business_shortlist_pca$rotation),
    pc1 = business_shortlist_pca$rotation[, 1],
    pc2 = business_shortlist_pca$rotation[, 2],
    pc3 = business_shortlist_pca$rotation[, 3]
  ) |>
    left_join(business_topic_competitor_tbl |> select(metric, topic_family), by = "metric") |>
    mutate(metric_label = unname(metric_labels[metric]))
} else {
  tibble(
    metric = character(),
    pc1 = double(),
    pc2 = double(),
    pc3 = double(),
    topic_family = character(),
    metric_label = character()
  )
}

cor_flag_tbl <- bind_rows(
  high_cor_tbl |> transmute(metric = metric_a, correlation = abs(correlation)),
  high_cor_tbl |> transmute(metric = metric_b, correlation = abs(correlation))
) |>
  group_by(metric) |>
  summarise(max_abs_correlation = max(correlation), .groups = "drop")

manual_recommendation_overrides <- tribble(
  ~metric, ~recommendation_manual,
  "income_pc_growth_5yr", "Core candidate",
  "income_pc_growth_1yr", "Optional sensitivity",
  "income_pc_cagr_5yr", "Drop",
  "economic_connectedness", "Optional sensitivity / proxy",
  "pct_unemployment_rate", "Core candidate",
  "lfpr", "Core candidate",
  "unemployment_rate_change_1yr", "Optional sensitivity",
  "lfpr_growth_5yr", "Optional sensitivity",
  "pov_rate_change_5yr", "Core candidate",
  "gini_index", "Optional sensitivity",
  "pov_rate", "Baseline context only",
  "pov_rate_change_1yr", "Optional sensitivity",
  "qcew_private_avg_wkly_wage", "Core candidate",
  "qcew_total_covered_avg_wkly_wage", "Redundant - keep one lead KPI",
  "hpi_5yr_pct", "Core candidate",
  "hpi_yoy_pct", "Core candidate",
  "hpi_10yr_pct", "Optional sensitivity",
  "zori_annual_avg_yoy_pct", "Core candidate",
  "zori_annual_avg", "Baseline context only",
  "zori_december", "Drop",
  "zori_december_yoy_pct", "Optional sensitivity",
  "pop_growth_5yr", "Core candidate",
  "pop_growth_1yr", "Optional sensitivity",
  "pop_cagr_5yr", "Redundant - keep one lead KPI",
  "irs_net_migration_rate", "Core candidate",
  "irs_net_agi", "Core candidate",
  "irs_inflow_agi", "Redundant - keep one lead KPI",
  "irs_outflow_agi", "Redundant - keep one lead KPI",
  "permits_per_1000_housing_units", "Core candidate",
  "permits_share_units_5_plus", "Core candidate",
  "permits_share_multifam_units", "Optional sensitivity",
  "permits_per_1000_population", "Redundant - keep one lead KPI",
  "permits_avg_units_per_bldg", "Optional sensitivity",
  "productivity_growth_5yr", "Core candidate",
  "real_gdp_growth_5yr", "Optional sensitivity",
  "real_gdp_pc_growth_5yr", "Redundant - keep one lead KPI",
  "real_gdp_cagr_5yr", "Drop",
  "industry_concentration_hhi", "Core candidate",
  "bfs_business_application_rate_per_1000_establishments", "Core candidate",
  "bfs_business_applications_yoy_pct", "Optional sensitivity",
  "bfs_business_applications_per_1000_residents", "Optional sensitivity",
  "bfs_business_applications", "Baseline context only",
  "cbp_estabs_per_1000_residents", "Core candidate",
  "cbp_total_estabs", "Baseline context only",
  "pct_ba_plus_change_5yr", "Core candidate",
  "pct_ba_plus", "Baseline context only",
  "pct_qcew_private_emp_professional", "Core candidate",
  "pct_qcew_private_emp_information", "Core candidate",
  "pct_qcew_private_emp_manufacturing", "Core candidate",
  "pct_qcew_private_emp_educ_health", "Optional sensitivity",
  "pct_real_gdp_professional", "Core candidate",
  "pct_real_gdp_information", "Core candidate",
  "pct_real_gdp_manufacturing", "Core candidate",
  "pct_real_gdp_edu_health", "Optional sensitivity",
  "qcew_private_avg_wkly_wage_professional", "Optional sensitivity",
  "qcew_private_avg_wkly_wage_information", "Core candidate",
  "qcew_private_avg_wkly_wage_manufacturing", "Optional sensitivity",
  "qcew_private_avg_wkly_wage_finance_real", "Optional sensitivity",
  "lq_professional", "Core candidate",
  "lq_information", "Core candidate",
  "lq_manufacturing", "Core candidate",
  "lq_educ_health", "Optional sensitivity",
  "pct_cbp_estabs_professional", "Optional sensitivity",
  "pct_cbp_estabs_information", "Optional sensitivity",
  "pct_cbp_estabs_manufacturing", "Optional sensitivity"
)

kpi_decision_tbl <- variance_tbl |>
  left_join(cor_flag_tbl, by = "metric") |>
  left_join(manual_recommendation_overrides, by = "metric") |>
  mutate(
    max_abs_correlation = coalesce(max_abs_correlation, 0),
    recommendation_auto = case_when(
      role == "proxy" & coverage_pct >= 0.75 & cv >= 0.2 ~ "Optional sensitivity / proxy",
      role == "proxy" ~ "Drop",
      role == "baseline" & coverage_pct >= 0.75 & cv >= 0.2 ~ "Baseline context only",
      role == "baseline" ~ "Drop",
      role == "optional" & coverage_pct >= 0.75 & cv >= 0.2 ~ "Optional sensitivity",
      role == "optional" ~ "Drop",
      coverage_pct < 0.75 ~ "Drop",
      cv < 0.2 ~ "Drop",
      max_abs_correlation >= 0.9 ~ "Redundant - keep one lead KPI",
      max_abs_correlation >= 0.75 ~ "Keep, but de-duplicate",
      coverage_pct < 0.9 ~ "Optional sensitivity",
      TRUE ~ "Core candidate"
    ),
    recommendation = coalesce(recommendation_manual, recommendation_auto),
    recommendation = case_when(
      topic %in% c("Establishment mix", "Sector GDP mix", "Sector employment mix", "Sector wage mix") &
        !metric %in% c("pct_real_gdp_information") &
        recommendation %in% c("Core candidate", "Redundant - keep one lead KPI", "Keep, but de-duplicate") ~ "Optional sensitivity",
      TRUE ~ recommendation
    )
  ) |>
  arrange(subject, topic, desc(cv))

topic_decision_tbl <- kpi_decision_tbl |>
  group_by(subject, topic) |>
  summarise(
    avg_coverage_pct = mean(coverage_pct),
    core_metrics = sum(recommendation == "Core candidate"),
    redundant_metrics = sum(recommendation %in% c("Redundant - keep one lead KPI", "Keep, but de-duplicate")),
    proxy_metrics = sum(recommendation == "Optional sensitivity / proxy"),
    baseline_metrics = sum(recommendation == "Baseline context only"),
    optional_metrics = sum(recommendation == "Optional sensitivity"),
    dropped_metrics = sum(recommendation == "Drop"),
    lead_kpis = paste(metric_label[order(cv, decreasing = TRUE)][seq_len(min(3, n()))], collapse = "; "),
    .groups = "drop"
  ) |>
  left_join(topic_pca_tbl, by = c("subject", "topic")) |>
  mutate(
    topic_decision = case_when(
      core_metrics == 0 & redundant_metrics == 0 & optional_metrics == 0 & proxy_metrics == 0 & baseline_metrics == 0 ~ "Drop",
      core_metrics == 0 & redundant_metrics == 0 & optional_metrics == 0 & proxy_metrics > 0 ~ "Sensitivity-only proxy topic",
      core_metrics == 0 & redundant_metrics == 0 & optional_metrics > 0 ~ "Sensitivity-only topic",
      core_metrics == 0 & redundant_metrics == 0 & baseline_metrics > 0 ~ "Baseline-only topic",
      core_metrics == 1 & redundant_metrics == 0 ~ "Keep as one-KPI topic",
      components_80 <= 1 ~ "Collapse to one lead KPI",
      TRUE ~ "Keep as multi-KPI topic"
    )
  ) |>
  arrange(subject, topic)

subject_decision_tbl <- topic_decision_tbl |>
  group_by(subject) |>
  summarise(
    topic_count = n(),
    kept_topics = sum(topic_decision != "Drop"),
    strong_topics = sum(topic_decision %in% c("Keep as one-KPI topic", "Collapse to one lead KPI", "Keep as multi-KPI topic")),
    avg_coverage_pct = mean(avg_coverage_pct, na.rm = TRUE),
    subject_decision = case_when(
      strong_topics == 0 ~ "Drop subject from Phase 1 core set",
      strong_topics == 1 ~ "Keep subject, but it is thin",
      TRUE ~ "Keep subject in Opportunity"
    ),
    .groups = "drop"
  ) |>
  arrange(subject)

df_pca <- df |>
  select(all_of(surviving)) |>
  drop_na() |>
  scale()

pca_result <- prcomp(df_pca, center = FALSE, scale. = FALSE)
component_count_80 <- which(cumsum((pca_result$sdev^2) / sum(pca_result$sdev^2)) >= 0.8)[1]

resident_divergence_tbl <- df |>
  filter(!is.na(income_pc_growth_1yr), !is.na(income_pc_growth_5yr)) |>
  mutate(
    divergence = income_pc_growth_1yr - income_pc_growth_5yr,
    label = if_else(abs(divergence) >= quantile(abs(divergence), 0.95, na.rm = TRUE), geo_name, NA_character_)
  )

market_divergence_tbl <- df |>
  filter(!is.na(hpi_yoy_pct), !is.na(hpi_5yr_pct)) |>
  mutate(
    divergence = hpi_yoy_pct - hpi_5yr_pct,
    label = if_else(abs(divergence) >= quantile(abs(divergence), 0.95, na.rm = TRUE), geo_name, NA_character_)
  )
```

## Source-family audit

```{r source-family-audit, echo=FALSE}
source_family_tbl |>
  transmute(
    `Source family` = source_family,
    `Latest year used` = latest_year,
    `Representative metric` = unname(metric_labels[coverage_metric]),
    `CBSAs with data` = coverage_n,
    `Coverage` = percent(coverage_pct, accuracy = 0.1),
    Note = note
  ) |>
  kable()
```

```{r inventory-audit, echo=FALSE}
topic_inventory_notes |>
  kable()
```

**Read of the audit**

- Opportunity is only coherent if we respect the different usable vintages across families.
- The biggest trap in this frame is assuming the latest row is always the latest usable row. In the current Gold surface, `2024` rows often blank out `5yr` growth metrics while still keeping level fields alive.
- `ZORI`, `IRS AGI flows`, `BFS rate fields`, and the static `economic_connectedness` proxy are all usable, but each needs an explicit caveat rather than being treated as clean recurring panels.
- `Intergenerational mobility` is still not a live Phase 1 direct topic. It is named here, but only through the `economic_connectedness` proxy.

## Subject and topic inventory

```{r subject-topic-inventory, echo=FALSE}
metric_catalog |>
  count(subject, topic, role, name = "metric_count") |>
  arrange(subject, topic, role) |>
  kable()
```

## Coverage by topic

```{r coverage-summary, echo=FALSE}
subject_topic_coverage_tbl <- coverage_tbl |>
  group_by(subject, topic) |>
  summarise(
    metric_count = n(),
    min_coverage_pct = min(coverage_pct),
    avg_coverage_pct = mean(coverage_pct),
    .groups = "drop"
  ) |>
  arrange(subject, topic)

subject_topic_coverage_tbl |>
  mutate(
    min_coverage_pct = percent(min_coverage_pct, accuracy = 0.1),
    avg_coverage_pct = percent(avg_coverage_pct, accuracy = 0.1)
  ) |>
  kable()
```

```{r coverage-plot, echo=FALSE}
subject_topic_coverage_tbl |>
  mutate(topic = reorder(topic, avg_coverage_pct)) |>
  ggplot(aes(x = avg_coverage_pct, y = topic, fill = subject)) +
  geom_col(alpha = 0.9) +
  geom_vline(
    xintercept = 0.9,
    linetype = "22",
    color = visual_neutral_palette()$text_muted,
    linewidth = 0.5
  ) +
  scale_x_continuous(labels = label_percent()) +
  labs(
    title = "Coverage by Opportunity topic",
    subtitle = "Average KPI coverage within each topic | Dashed line = 90% caution threshold",
    x = "Average coverage",
    y = NULL,
    fill = NULL,
    caption = build_chart_notes(
      source = "BEA, ACS, IRS, FHFA, Zillow, BPS, QCEW, BFS, CBP, Social Capital Atlas",
      vintage = "Latest usable row by source family"
    )
  ) +
  visual_theme()
```

## 1. KPI variance pass

```{r variance-summary, echo=FALSE}
variance_tbl |>
  transmute(
    subject,
    topic,
    metric = metric_label,
    role,
    coverage = percent(coverage_pct, accuracy = 0.1),
    cv = round(cv, 3),
    iqr = round(iqr, 3)
  ) |>
  arrange(subject, topic, desc(cv)) |>
  kable()
```

```{r variance-plot, echo=FALSE, fig.height=18}
variance_tbl |>
  mutate(metric_label = reorder(metric_label, cv)) |>
  ggplot(aes(x = cv, y = metric_label, fill = subject)) +
  geom_col(alpha = 0.85) +
  geom_vline(
    xintercept = 0.2,
    linetype = "22",
    color = visual_neutral_palette()$text_muted,
    linewidth = 0.5
  ) +
  labs(
    title = "Opportunity metrics ranked by coefficient of variation",
    subtitle = "Higher CV means a metric separates CBSAs more strongly",
    x = "Coefficient of variation (SD / mean)",
    y = NULL,
    fill = NULL,
    caption = build_chart_notes(
      source = "BEA, ACS, IRS, FHFA, Zillow, BPS, QCEW, BFS, CBP, Social Capital Atlas",
      vintage = "Latest usable row by source family"
    )
  ) +
  visual_theme()
```

## 2. Topic scorecard

```{r topic-scorecard, echo=FALSE}
topic_scorecard_tbl |>
  mutate(
    avg_coverage_pct = percent(avg_coverage_pct, accuracy = 0.1)
  ) |>
  kable()
```

## 3. Time-horizon diagnostics

```{r resident-time-horizon, echo=FALSE}
resident_divergence_tbl |>
  ggplot(aes(x = income_pc_growth_5yr, y = income_pc_growth_1yr)) +
  geom_point(alpha = 0.7, color = visual_palette_defaults()$highlight$opportunity) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "22",
    color = visual_neutral_palette()$text_muted,
    linewidth = 0.5
  ) +
  geom_text_repel(aes(label = label), size = 2.6, max.overlaps = 20) +
  labs(
    title = "Resident momentum: 1-year vs 5-year income growth",
    subtitle = "Above the line = accelerating lately | Below the line = cooling vs structural trend",
    x = "Income growth (5yr)",
    y = "Income growth (1yr)",
    caption = build_chart_notes(
      source = "BEA CAINC",
      vintage = paste0("Latest usable growth row: ", max(df$income_growth_year, na.rm = TRUE))
    )
  ) +
  visual_theme()
```

```{r market-time-horizon, echo=FALSE}
market_divergence_tbl |>
  ggplot(aes(x = hpi_5yr_pct, y = hpi_yoy_pct)) +
  geom_point(alpha = 0.7, color = visual_palette_defaults()$highlight$selection) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "22",
    color = visual_neutral_palette()$text_muted,
    linewidth = 0.5
  ) +
  geom_text_repel(aes(label = label), size = 2.6, max.overlaps = 20) +
  labs(
    title = "Market momentum: YoY vs 5-year HPI growth",
    subtitle = "Above the line = recent overheating / acceleration | Below the line = cooling vs structural trend",
    x = "HPI growth (5yr)",
    y = "HPI YoY growth",
    caption = build_chart_notes(
      source = "FHFA HPI",
      vintage = paste0("Latest usable HPI row: ", max(df$housing_hpi_year, na.rm = TRUE))
    )
  ) +
  visual_theme()
```

## 4. Correlation pass

```{r correlation-summary, echo=FALSE}
high_cor_tbl |>
  slice_head(n = 60) |>
  transmute(
    `Subject / topic A` = paste(subject_a, topic_a, sep = " / "),
    `Metric A` = metric_a_label,
    `Subject / topic B` = paste(subject_b, topic_b, sep = " / "),
    `Metric B` = metric_b_label,
    `r` = round(correlation, 2)
  ) |>
  kable()
```

```{r correlation-plot, echo=FALSE}
high_cor_tbl |>
  slice_head(n = 25) |>
  mutate(
    pair = paste(metric_a_label, metric_b_label, sep = " vs "),
    pair = reorder(pair, abs(correlation))
  ) |>
  ggplot(aes(x = correlation, y = pair, fill = correlation > 0)) +
  geom_col(alpha = 0.85) +
  scale_fill_manual(
    values = c(
      "TRUE" = visual_palette_defaults()$diverging$better,
      "FALSE" = visual_palette_defaults()$diverging$worse
    ),
    guide = "none"
  ) +
  labs(
    title = "Strongest Opportunity correlation pairs",
    subtitle = "Only pairs with |r| > 0.75 are shown",
    x = "Correlation",
    y = NULL,
    caption = build_chart_notes(
      source = "Latest-usable Opportunity frame",
      vintage = "Mixed vintages by source family"
    )
  ) +
  visual_theme()
```

## 5. Cross-subject and cross-topic overlap

```{r cross-topic-correlation, echo=FALSE}
cross_topic_cor_tbl |>
  slice_head(n = 40) |>
  transmute(
    `Subject / topic A` = paste(subject_a, topic_a, sep = " / "),
    `Metric A` = metric_a_label,
    `Subject / topic B` = paste(subject_b, topic_b, sep = " / "),
    `Metric B` = metric_b_label,
    `r` = round(correlation, 2)
  ) |>
  kable()
```

**Read of the overlap check**

- This is the guardrail against carrying the same story twice across subjects or topics.
- Metrics that look redundant for the cluster model can still remain valuable as descriptive deep-dive fields.
- The main place to watch is the handoff from `Resident Opportunity` into `Market / Investor Opportunity`, and then again inside the dense `Business & Industry Opportunity` structure families.

## 6. Business-topic competition

```{r business-topic-competitors, echo=FALSE}
business_topic_competitor_tbl |>
  mutate(
    coverage_pct = percent(coverage_pct, accuracy = 0.1),
    cv = round(cv, 3)
  ) |>
  transmute(
    `Topic family` = topic_family,
    `Metric` = metric_label,
    `Cluster shortlist` = if_else(cluster_shortlist, "Yes", "No"),
    `Coverage` = coverage_pct,
    `CV` = cv
  ) |>
  kable()
```

```{r business-topic-correlations, echo=FALSE}
business_topic_competitor_cor_tbl |>
  slice_head(n = 35) |>
  transmute(
    `Topic family A` = family_a,
    `Metric A` = metric_a_label,
    `Topic family B` = family_b,
    `Metric B` = metric_b_label,
    `r` = round(correlation, 2)
  ) |>
  kable()
```

```{r business-topic-pca, echo=FALSE}
business_shortlist_pca_tbl |>
  mutate(
    pc1 = round(pc1, 3),
    pc2 = round(pc2, 3),
    pc3 = round(pc3, 3)
  ) |>
  transmute(
    `Topic family` = topic_family,
    `Metric` = metric_label,
    `PC1 loading` = pc1,
    `PC2 loading` = pc2,
    `PC3 loading` = pc3
  ) |>
  kable()
```

**Read of the business-topic competition**

- The point of this section is not to let `Location quotient specialization` win by default just because it is a convenient topic.
- Instead, we test a cross-topic shortlist drawn from `Location quotient specialization`, `Sector GDP mix`, `Sector employment mix`, `Sector wage mix`, and `Establishment mix`.
- The notebook’s recommendation should be read as:
  - keep a small business-structure cluster bundle
  - let the rest of the business-topic KPIs stay alive as descriptive deep-dive or sensitivity fields
  - avoid carrying every parallel industry family into the default model just because the data exists

## 7. PCA diagnostic

```{r pca-scree, echo=FALSE}
tibble(
  component = seq_along(pca_result$sdev),
  variance_explained = (pca_result$sdev^2) / sum(pca_result$sdev^2)
) |>
  mutate(cumulative = cumsum(variance_explained)) |>
  ggplot(aes(x = component, y = variance_explained)) +
  geom_col(fill = visual_palette_defaults()$highlight$strength, alpha = 0.85) +
  geom_line(
    aes(y = cumulative),
    color = visual_neutral_palette()$text_muted,
    linewidth = 0.6
  ) +
  geom_point(
    aes(y = cumulative),
    color = visual_neutral_palette()$text_muted,
    size = 2
  ) +
  scale_y_continuous(labels = label_percent()) +
  labs(
    title = "PCA scree plot - Opportunity metrics",
    subtitle = "Bars = variance per component | Line = cumulative variance explained",
    x = "Principal component",
    y = "Variance explained",
    caption = build_chart_notes(
      source = "Latest-usable Opportunity frame",
      vintage = "Mixed vintages by source family"
    )
  ) +
  visual_theme()
```

```{r topic-pca, echo=FALSE}
topic_pca_tbl |>
  arrange(subject, topic) |>
  kable()
```

## 8. Recommendations by subject, topic, and KPI

```{r decision-tables, echo=FALSE}
subject_decision_tbl |>
  mutate(avg_coverage_pct = percent(avg_coverage_pct, accuracy = 0.1)) |>
  kable()

topic_decision_tbl |>
  mutate(avg_coverage_pct = percent(avg_coverage_pct, accuracy = 0.1)) |>
  kable()
```

```{r decisions, echo=FALSE, results='asis'}
for (subject_name in unique(metric_catalog$subject)) {
  subject_row <- subject_decision_tbl |>
    filter(subject == subject_name)

  cat(glue("<h3>{subject_name}</h3>\n"))
  cat(glue("<p><strong>Subject decision:</strong> {subject_row$subject_decision}<br><strong>Coverage read:</strong> average topic coverage = {percent(subject_row$avg_coverage_pct, accuracy = 0.1)}</p>\n"))

  subject_topics <- topic_decision_tbl |>
    filter(subject == subject_name)

  for (i in seq_len(nrow(subject_topics))) {
    topic_row <- subject_topics[i, ]
    topic_metrics <- kpi_decision_tbl |>
      filter(subject == subject_name, topic == topic_row$topic)

    core_kpis <- topic_metrics |>
      filter(recommendation == "Core candidate") |>
      pull(metric_label)

    redundant_kpis <- topic_metrics |>
      filter(recommendation %in% c("Redundant - keep one lead KPI", "Keep, but de-duplicate")) |>
      pull(metric_label)

    optional_kpis <- topic_metrics |>
      filter(recommendation == "Optional sensitivity") |>
      pull(metric_label)

    proxy_kpis <- topic_metrics |>
      filter(recommendation == "Optional sensitivity / proxy") |>
      pull(metric_label)

    baseline_kpis <- topic_metrics |>
      filter(recommendation == "Baseline context only") |>
      pull(metric_label)

    drop_kpis <- topic_metrics |>
      filter(recommendation == "Drop") |>
      pull(metric_label)

    cat(glue("<h4>{topic_row$topic}</h4>\n"))
    cat(glue("<p><strong>Topic decision:</strong> {topic_row$topic_decision}<br><strong>Lead KPIs by variance:</strong> {topic_row$lead_kpis}<br><strong>Topic evidence:</strong> coverage = {percent(topic_row$avg_coverage_pct, accuracy = 0.1)}; high-variance metrics = {topic_scorecard_tbl$high_variance_metrics[match(topic_row$topic, topic_scorecard_tbl$topic)]}; topic PCA read = {topic_row$topic_dimensionality}</p>\n"))

    if (length(core_kpis) > 0) {
      cat("<p><strong>Core KPIs:</strong></p>\n<ul>\n")
      for (metric_name in core_kpis) {
        cat(glue("<li>{metric_name}</li>\n"))
      }
      cat("</ul>\n")
    }

    if (length(redundant_kpis) > 0) {
      cat("<p><strong>Redundant / consolidate KPIs:</strong></p>\n<ul>\n")
      for (metric_name in redundant_kpis) {
        cat(glue("<li>{metric_name}</li>\n"))
      }
      cat("</ul>\n")
    }

    if (length(optional_kpis) > 0) {
      cat("<p><strong>Optional sensitivity KPIs:</strong></p>\n<ul>\n")
      for (metric_name in optional_kpis) {
        cat(glue("<li>{metric_name}</li>\n"))
      }
      cat("</ul>\n")
    }

    if (length(proxy_kpis) > 0) {
      cat("<p><strong>Proxy KPIs:</strong></p>\n<ul>\n")
      for (metric_name in proxy_kpis) {
        cat(glue("<li>{metric_name}</li>\n"))
      }
      cat("</ul>\n")
    }

    if (length(baseline_kpis) > 0) {
      cat("<p><strong>Baseline / context KPIs:</strong></p>\n<ul>\n")
      for (metric_name in baseline_kpis) {
        cat(glue("<li>{metric_name}</li>\n"))
      }
      cat("</ul>\n")
    }

    if (length(drop_kpis) > 0) {
      cat("<p><strong>Drop from default Phase 1 set:</strong></p>\n<ul>\n")
      for (metric_name in drop_kpis) {
        cat(glue("<li>{metric_name}</li>\n"))
      }
      cat("</ul>\n")
    }
  }
}

cat(glue("<p><strong>PCA read:</strong> <code>{component_count_80}</code> components explain at least 80% of the variance in the surviving Opportunity set.</p>\n"))
cat("<p><strong>Working interpretation:</strong></p>\n<ul>\n")
cat("<li>Opportunity is broad enough to justify three separate subjects, but the largest redundancy problem sits inside the business-and-industry families where sector-share, wage, and specialization metrics overlap heavily.</li>\n")
cat("<li>The biggest source-quality caution is vintage mismatch: growth, HHI, BFS-rate, and CBP fields are usable only when we deliberately pull their latest non-null rows rather than blindly taking the 2024 record.</li>\n")
cat("<li>Resident and market time-horizon diagnostics should be treated as momentum checks, not as a reason to automatically carry both the short-run and long-run version of every KPI.</li>\n")
cat("<li>For the cluster model, redundancy is a trimming rule rather than a deletion rule: many non-core KPIs still stay alive as descriptive or sensitivity fields for Deep Dive interpretation.</li>\n")
cat("<li>The business-structure decision should be read as a hybrid outcome, not an LQ-only shortcut: the notebook explicitly checks overlap across LQ, GDP mix, employment mix, wage mix, and establishment mix before naming the default bundle.</li>\n")
cat("</ul>\n")
```

## Working read

This Opportunity pass should be read in order:

1. decide whether the **subject** belongs in Opportunity
2. decide whether the **topic** is recurring, thin, redundant, baseline-only, proxy-only, or weak
3. decide which **KPI** earns the default slot once we account for coverage, variance, overlap, and source vintage

That matters here because Opportunity is more source-misaligned than Character or Livability:

- `Resident Opportunity` mixes true growth fields with level and inclusion fields that live on different usable vintages.
- `Market / Investor Opportunity` is relatively clean, but it is still heavily housing-weighted, rent momentum is coverage-constrained, and commercial real estate is still a live gap.
- `Business & Industry Opportunity` is analytically rich, but it is also the easiest place to over-count the same structure signal through employment shares, GDP shares, establishment shares, location quotients, and sector wage families.

**Update `exploration/intelligence_framework/docs/metric_selections.md` after reviewing this notebook output.**

```{r cleanup, include=FALSE}
dbDisconnect(con)
```
