---
title: "Phase 7 - Zone Methodology Review"
subtitle: "Review notebook for the modular Sprint 2 national tract model"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 2
    code-fold: true
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)

options(
  dplyr.summarise.inform = FALSE,
  knitr.kable.NA = ""
)

library(here)
library(dplyr)
library(readr)
library(arrow)
library(ggplot2)
library(knitr)

output_dir <- here("exploration/intelligence_framework/phase_7_zone_methodology", "outputs")

cluster_calibration <- read_csv(file.path(output_dir, "phase7_cluster_calibration.csv"), show_col_types = FALSE)
coverage_audit <- read_csv(file.path(output_dir, "phase7_coverage_audit.csv"), show_col_types = FALSE)
cluster_centroids <- read_csv(file.path(output_dir, "phase7_cluster_centroids.csv"), show_col_types = FALSE)
representative_tracts <- read_csv(file.path(output_dir, "phase7_representative_tracts.csv"), show_col_types = FALSE)
light_validation <- read_csv(file.path(output_dir, "phase7_light_validation_spot_checks.csv"), show_col_types = FALSE)
zone_scores <- read_parquet(file.path(output_dir, "zone_scores.parquet"))

selected_k <- zone_scores |>
  distinct(selected_k_for_run) |>
  pull(selected_k_for_run)

model_summary <- tibble(
  modeling_universe = nrow(zone_scores),
  cbsa_count = n_distinct(zone_scores$cbsa_code),
  selected_k_for_run = selected_k,
  share_in_oz = mean(zone_scores$is_opportunity_zone, na.rm = TRUE)
)
```

## Purpose

This notebook is the review layer for Sprint 2, not the build pipeline.

The canonical build now lives in:

- `R/run_phase7_zone_model.R`
- `R/phase7_*.R`

This notebook reads the saved artifacts in `outputs/` so we can:

- review the current candidate-`k` calibration
- inspect the provisional centroid structure
- review representative tracts for naming sanity checks
- look at light Jacksonville and Richmond spot checks before Sprint 3

The current Sprint 2 run uses sampled hierarchical calibration and skips GMM by default so the tract-scale pass stays practical to rerun.

## Model Snapshot

```{r model-summary}
model_summary |>
  mutate(share_in_oz = round(share_in_oz, 3)) |>
  kable()
```

## Coverage Audit

```{r coverage-audit}
coverage_audit |>
  select(metric_id, theme, completeness_pct, missing_tracts, missing_gt_20_pct) |>
  mutate(completeness_pct = round(completeness_pct, 3)) |>
  kable()
```

## Cluster Count Calibration

```{r cluster-calibration-plot, fig.height=4.5}
cluster_calibration |>
  ggplot(aes(x = k, y = kmeans_avg_silhouette)) +
  geom_line(color = "#355070", linewidth = 0.8) +
  geom_point(aes(color = k == selected_k), size = 2.8) +
  scale_color_manual(values = c("FALSE" = "#6c757d", "TRUE" = "#d62828"), guide = "none") +
  scale_x_continuous(breaks = cluster_calibration$k) +
  labs(
    title = "Phase 7 provisional cluster-count calibration",
    x = "Cluster count (k)",
    y = "Average k-means silhouette"
  ) +
  theme_minimal(base_size = 12)
```

```{r cluster-calibration-table}
cluster_calibration |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable()
```

## Provisional Cluster Profiles

```{r cluster-centroids}
cluster_centroids |>
  select(
    zone_kmeans_cluster,
    provisional_zone_type,
    tracts_in_cluster,
    subject_score_character,
    subject_score_livability,
    subject_score_opportunity,
    zone_type_name_status
  ) |>
  mutate(across(where(is.numeric), \(x) round(x, 2))) |>
  kable()
```

## Representative Tracts

```{r representative-tracts}
representative_tracts |>
  group_by(provisional_zone_type) |>
  slice_head(n = 5) |>
  ungroup() |>
  mutate(distance_to_center = round(distance_to_center, 3)) |>
  kable()
```

## Light Market Spot Checks

```{r light-validation}
light_validation |>
  mutate(cbsa_zone_share = round(cbsa_zone_share, 3)) |>
  kable()
```

## Final Read

- Sprint 2 now has a full modular Phase 7 package rather than a notebook-only workflow.
- The tract modeling universe is the current governed Gold tract backbone, not the legacy 396-CBSA scope.
- Candidate-`k` calibration is based on a `5,000`-tract sample because full tract pairwise distance matrices exceed local memory.
- `k` and cluster names are intentionally still provisional pending output review.
- GMM is currently disabled by default for tract-scale runs; the hard labels and benchmark outputs are the main Sprint 2 deliverables.
- Jacksonville and Richmond are included only as light distribution checks in this sprint; full spatial validation stays in Sprint 3.
