---
title: "Phase 3 - AQI Median Review"
subtitle: "Baseline vs revised Livability comparison for the AQI KPI swap"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 2
    code-fold: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)

options(
  dplyr.summarise.inform = FALSE,
  knitr.kable.NA = ""
)

library(here)
library(dplyr)
library(readr)
library(arrow)
library(ggplot2)
library(scales)
library(knitr)
library(tidyr)
library(stringr)

baseline_output_dir <- here(
  "exploration",
  "intelligence_framework",
  "phase_3_livability_calibration",
  "outputs"
)

revised_output_dir <- here(
  "exploration",
  "intelligence_framework",
  "phase_3_livability_calibration",
  "outputs_aqi_median_review"
)

read_environment_axis <- function(output_dir) {
  axis_df <- readr::read_csv(
    file.path(output_dir, "livability_phase3_environment_axis_scatter.csv"),
    show_col_types = FALSE
  )

  if (!"aqi_metric_id" %in% names(axis_df)) {
    axis_df <- axis_df |>
      mutate(
        aqi_metric_id = "aqi_unhealthy_days",
        aqi_value = aqi_unhealthy_days
      )
  }

  axis_df |>
    mutate(
      aqi_label = case_when(
        aqi_metric_id == "aqi_median" ~ "AQI median",
        aqi_metric_id == "aqi_unhealthy_days" ~ "AQI unhealthy days",
        TRUE ~ aqi_metric_id
      )
    )
}

read_phase3_bundle <- function(output_dir, scenario_label) {
  cluster_calibration <- readr::read_csv(
    file.path(output_dir, "livability_phase3_cluster_count_calibration.csv"),
    show_col_types = FALSE
  )
  cluster_sizes <- readr::read_csv(
    file.path(output_dir, "livability_phase3_cluster_count_sizes.csv"),
    show_col_types = FALSE
  )
  cluster_centroids <- readr::read_csv(
    file.path(output_dir, "livability_phase3_cluster_centroids.csv"),
    show_col_types = FALSE
  )
  cluster_representatives <- readr::read_csv(
    file.path(output_dir, "livability_phase3_cluster_representatives.csv"),
    show_col_types = FALSE
  )
  metric_completeness <- readr::read_csv(
    file.path(output_dir, "livability_phase3_metric_completeness.csv"),
    show_col_types = FALSE
  )
  pca_variance <- readr::read_csv(
    file.path(output_dir, "livability_phase3_pca_variance.csv"),
    show_col_types = FALSE
  )
  pca_loadings <- readr::read_csv(
    file.path(output_dir, "livability_phase3_pca_loadings.csv"),
    show_col_types = FALSE
  )
  livability_scores <- arrow::read_parquet(file.path(output_dir, "livability_scores.parquet"))
  environment_axis <- read_environment_axis(output_dir)

  list(
    scenario_label = scenario_label,
    cluster_calibration = cluster_calibration,
    cluster_sizes = cluster_sizes,
    cluster_centroids = cluster_centroids,
    cluster_representatives = cluster_representatives,
    metric_completeness = metric_completeness,
    pca_variance = pca_variance,
    pca_loadings = pca_loadings,
    livability_scores = livability_scores,
    environment_axis = environment_axis
  )
}

baseline <- read_phase3_bundle(baseline_output_dir, "baseline_aqi_unhealthy_days")
revised <- read_phase3_bundle(revised_output_dir, "review_aqi_median")

comparison_summary <- tibble::tribble(
  ~scenario, ~aqi_metric, ~rows, ~best_k_by_silhouette, ~silhouette_at_k6, ~aqi_fema_correlation,
  "Baseline", "aqi_unhealthy_days",
  nrow(baseline$livability_scores),
  baseline$cluster_calibration$k[which.max(baseline$cluster_calibration$avg_silhouette)],
  baseline$cluster_calibration |> filter(k == 6) |> pull(avg_silhouette),
  cor(baseline$environment_axis$aqi_value, baseline$environment_axis$fema_risk_score),
  "Revised", "aqi_median",
  nrow(revised$livability_scores),
  revised$cluster_calibration$k[which.max(revised$cluster_calibration$avg_silhouette)],
  revised$cluster_calibration |> filter(k == 6) |> pull(avg_silhouette),
  cor(revised$environment_axis$aqi_value, revised$environment_axis$fema_risk_score)
)

aqi_completeness_compare <- bind_rows(
  baseline$metric_completeness |> mutate(scenario = "Baseline"),
  revised$metric_completeness |> mutate(scenario = "Revised")
) |>
  filter(str_detect(metric_id, "^aqi_")) |>
  select(
    scenario,
    metric_id,
    non_null_cbsas,
    missing_cbsas,
    completeness_pct,
    median_value
  )

pca_variance_compare <- bind_rows(
  baseline$pca_variance |> mutate(scenario = "Baseline"),
  revised$pca_variance |> mutate(scenario = "Revised")
) |>
  filter(principal_component %in% paste0("PC", 1:8))

cluster_calibration_compare <- bind_rows(
  baseline$cluster_calibration |> mutate(scenario = "Baseline"),
  revised$cluster_calibration |> mutate(scenario = "Revised")
)

k6_cluster_size_compare <- bind_rows(
  baseline$cluster_sizes |> mutate(scenario = "Baseline"),
  revised$cluster_sizes |> mutate(scenario = "Revised")
) |>
  filter(k == 6) |>
  arrange(scenario, cluster)

percentile_shift_compare <- baseline$livability_scores |>
  select(
    cbsa_code,
    cbsa_name,
    baseline_percentile = livability_percentile,
    baseline_cluster_name = livability_cluster_name
  ) |>
  left_join(
    revised$livability_scores |>
      select(
        cbsa_code,
        revised_percentile = livability_percentile,
        revised_cluster_name = livability_cluster_name
      ),
    by = "cbsa_code"
  ) |>
  mutate(
    percentile_shift = revised_percentile - baseline_percentile,
    cluster_name_changed = baseline_cluster_name != revised_cluster_name
  ) |>
  arrange(desc(abs(percentile_shift)))
```

## Purpose

This review notebook compares the current published Phase 3 Livability outputs against the side-by-side `aqi_median` revision run.

It is meant to answer four questions before we move on to Phase 5:

- does the AQI swap materially change coverage or imputation behavior
- does the PCA story change in a meaningful way
- is `k = 6` still defensible
- do the revised clusters still look close enough to the published names to reuse them

## Snapshot

```{r snapshot}
comparison_summary |>
  mutate(
    across(where(is.numeric), \(x) round(x, 3))
  ) |>
  kable()
```

## AQI Completeness

```{r aqi-completeness}
aqi_completeness_compare |>
  mutate(
    completeness_pct = percent(completeness_pct, accuracy = 0.1),
    median_value = round(median_value, 2)
  ) |>
  kable()
```

## PCA Variance Comparison

```{r pca-variance-plot, fig.height=4.5}
pca_variance_compare |>
  ggplot(aes(x = principal_component, y = cumulative_variance, color = scenario, group = scenario)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Cumulative PCA variance by scenario",
    x = "Principal component",
    y = "Cumulative variance explained",
    color = "Scenario"
  ) +
  theme_minimal(base_size = 12)
```

```{r pca-variance-table}
pca_variance_compare |>
  select(scenario, principal_component, variance_explained, cumulative_variance) |>
  mutate(
    variance_explained = round(variance_explained, 3),
    cumulative_variance = round(cumulative_variance, 3)
  ) |>
  kable()
```

## Cluster Calibration Comparison

```{r cluster-calibration-plot, fig.height=4.5}
cluster_calibration_compare |>
  ggplot(aes(x = k, y = avg_silhouette, color = scenario)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = sort(unique(cluster_calibration_compare$k))) +
  labs(
    title = "Average silhouette by cluster count",
    x = "Cluster count (k)",
    y = "Average silhouette",
    color = "Scenario"
  ) +
  theme_minimal(base_size = 12)
```

```{r cluster-calibration-table}
cluster_calibration_compare |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable()
```

## Published `k = 6` Cluster Sizes

```{r cluster-size-compare}
k6_cluster_size_compare |>
  rename(livability_kmeans_cluster = cluster, metros_in_cluster = cluster_size) |>
  kable()
```

## Biggest Percentile Shifts

```{r percentile-shifts}
percentile_shift_compare |>
  slice_head(n = 20) |>
  mutate(
    baseline_percentile = round(baseline_percentile, 1),
    revised_percentile = round(revised_percentile, 1),
    percentile_shift = round(percentile_shift, 1)
  ) |>
  kable()
```

## Revised Representative Metros

```{r revised-representatives}
revised$cluster_representatives |>
  group_by(livability_cluster_name) |>
  slice_head(n = 6) |>
  ungroup() |>
  mutate(livability_percentile = round(livability_percentile, 1)) |>
  kable()
```

## AQI Vs. FEMA Comparison

```{r environment-axis-plot, fig.height=5}
bind_rows(
  baseline$environment_axis |> mutate(scenario = "Baseline"),
  revised$environment_axis |> mutate(scenario = "Revised")
) |>
  ggplot(aes(x = aqi_value, y = fema_risk_score, color = livability_cluster_name)) +
  geom_point(alpha = 0.7, size = 1.8) +
  facet_wrap(~scenario, scales = "free_x") +
  labs(
    title = "AQI and FEMA relationship under the baseline and revised KPI choice",
    x = "AQI value",
    y = "FEMA risk score",
    color = "Cluster"
  ) +
  theme_minimal(base_size = 12)
```

## Review Questions

- Does the revised run still make `k = 6` look like the best practical choice, even if another `k` edges it out numerically?
- Are the biggest percentile movers concentrated in a few AQI-sensitive metros, or does the whole map drift?
- Do the revised cluster profiles still match the current published labels closely enough to keep them?
- Is the revised AQI behavior easier to explain than the baseline rare-event count in the environmental hypothesis views?
