---
title: "Phase 3 - Livability Frame Review"
subtitle: "Review notebook for the modular Phase 3 Livability build"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 2
    code-fold: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)

options(
  dplyr.summarise.inform = FALSE,
  knitr.kable.NA = ""
)

library(here)
library(dplyr)
library(readr)
library(arrow)
library(ggplot2)
library(scales)
library(knitr)

output_dir <- here("exploration/intelligence_framework/phase_3_livability_calibration", "outputs")

cluster_calibration <- readr::read_csv(file.path(output_dir, "livability_phase3_cluster_count_calibration.csv"), show_col_types = FALSE)
cluster_sizes <- readr::read_csv(file.path(output_dir, "livability_phase3_cluster_count_sizes.csv"), show_col_types = FALSE)
cluster_centroids <- readr::read_csv(file.path(output_dir, "livability_phase3_cluster_centroids.csv"), show_col_types = FALSE)
cluster_representatives <- readr::read_csv(file.path(output_dir, "livability_phase3_cluster_representatives.csv"), show_col_types = FALSE)
gmm_summary <- readr::read_csv(file.path(output_dir, "livability_phase3_gmm_summary.csv"), show_col_types = FALSE)
gmm_hybrids <- readr::read_csv(file.path(output_dir, "livability_phase3_gmm_hybrids.csv"), show_col_types = FALSE)
k5_k6_changes <- readr::read_csv(file.path(output_dir, "livability_phase3_k5_to_k6_changed_metros.csv"), show_col_types = FALSE)
k5_k6_splits <- readr::read_csv(file.path(output_dir, "livability_phase3_k5_to_k6_split_summary.csv"), show_col_types = FALSE)
similarity_top10 <- readr::read_csv(file.path(output_dir, "livability_phase3_similarity_top10.csv"), show_col_types = FALSE)
health_affordability <- readr::read_csv(file.path(output_dir, "livability_phase3_health_affordability_scatter.csv"), show_col_types = FALSE)
environment_axis <- readr::read_csv(file.path(output_dir, "livability_phase3_environment_axis_scatter.csv"), show_col_types = FALSE)
livability_scores <- arrow::read_parquet(file.path(output_dir, "livability_scores.parquet"))

selected_k <- 6
selected_cluster_sizes <- cluster_sizes |>
  filter(k == selected_k) |>
  arrange(cluster)

model_summary <- tibble::tibble(
  modeling_universe = nrow(livability_scores),
  selected_k = selected_k,
  silhouette_at_k6 = cluster_calibration |>
    filter(k == selected_k) |>
    pull(avg_silhouette),
  health_affordability_correlation = health_affordability |>
    summarise(correlation = cor(affordability_score, health_score)) |>
    pull(correlation),
  environment_axis_correlation = environment_axis |>
    summarise(correlation = cor(aqi_unhealthy_days, fema_risk_score)) |>
    pull(correlation)
)
```

## Purpose

This notebook is the review layer for Phase 3, not the build pipeline.

The canonical build now lives in:

- `R/run_phase3_livability.R`
- `R/phase3_*.R`

This notebook reads the saved artifacts in `outputs/` so we can:

- review the chosen `k = 6` calibration
- inspect the published cluster structure
- look at soft-membership ambiguity
- sanity-check similarity outputs
- review the two main Livability hypothesis tests

## Model Snapshot

```{r model-summary}
model_summary |>
  mutate(
    silhouette_at_k6 = round(silhouette_at_k6, 3),
    health_affordability_correlation = round(health_affordability_correlation, 3),
    environment_axis_correlation = round(environment_axis_correlation, 3)
  ) |>
  kable()
```

## Cluster Count Calibration

```{r cluster-calibration-plot, fig.height=4.5}
cluster_calibration |>
  ggplot(aes(x = k, y = avg_silhouette)) +
  geom_line(color = "#355070", linewidth = 0.8) +
  geom_point(aes(color = k == selected_k), size = 2.8) +
  scale_color_manual(values = c("FALSE" = "#6c757d", "TRUE" = "#d62828"), guide = "none") +
  scale_x_continuous(breaks = cluster_calibration$k) +
  labs(
    title = "Cluster-count calibration",
    x = "Cluster count (k)",
    y = "Average silhouette"
  ) +
  theme_minimal(base_size = 12)
```

```{r cluster-calibration-table}
cluster_calibration |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable()
```

## Published Cluster Sizes

```{r selected-cluster-sizes}
selected_cluster_sizes |>
  rename(livability_kmeans_cluster = cluster, metros_in_cluster = cluster_size) |>
  left_join(
    cluster_centroids |>
      distinct(livability_kmeans_cluster, livability_cluster_name),
    by = "livability_kmeans_cluster"
  ) |>
  select(livability_kmeans_cluster, livability_cluster_name, metros_in_cluster) |>
  kable()
```

## Why `k = 6` Instead Of `k = 5`

The current published choice keeps the full 26-KPI set and uses `k = 6`. The extra cluster is not a complete remap; it is a refinement of the broader `k = 5` structure.

```{r k5-k6-split-summary}
k5_k6_splits |>
  arrange(desc(metros_in_split)) |>
  slice_head(n = 10) |>
  kable()
```

```{r k5-k6-changed-metros}
k5_k6_changes |>
  filter(changed_under_k6) |>
  arrange(desc(livability_percentile)) |>
  slice_head(n = 15) |>
  mutate(livability_percentile = round(livability_percentile, 1)) |>
  kable()
```

## Representative Metros By Cluster

```{r representatives}
cluster_representatives |>
  group_by(livability_cluster_name) |>
  slice_head(n = 6) |>
  ungroup() |>
  mutate(livability_percentile = round(livability_percentile, 1)) |>
  kable()
```

## Soft Membership Review

```{r gmm-summary}
gmm_summary |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable()
```

```{r gmm-hybrids}
gmm_hybrids |>
  arrange(membership_gap) |>
  slice_head(n = 15) |>
  mutate(across(contains("probability"), \(x) round(x, 3))) |>
  kable()
```

## Similarity Sanity Check

```{r similarity-preview}
similarity_top10 |>
  filter(cbsa_code %in% c(40060, 12420, 18140)) |>
  slice_head(n = 15) |>
  mutate(cosine_similarity = round(cosine_similarity, 3)) |>
  kable()
```

## Hypothesis Review

### Affordability Vs. Health

```{r health-affordability-plot, fig.height=5}
health_affordability |>
  ggplot(aes(
    x = affordability_score,
    y = health_score,
    color = livability_cluster_name
  )) +
  geom_hline(yintercept = 0, color = "#adb5bd", linewidth = 0.4) +
  geom_vline(xintercept = 0, color = "#adb5bd", linewidth = 0.4) +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "Affordability and health move together only moderately",
    x = "Affordability subject score",
    y = "Health and safety subject score",
    color = "Cluster"
  ) +
  theme_minimal(base_size = 12)
```

### AQI Vs. FEMA

```{r environment-axis-plot, fig.height=5}
environment_axis |>
  ggplot(aes(
    x = aqi_unhealthy_days,
    y = fema_risk_score,
    color = livability_cluster_name
  )) +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "FEMA adds signal beyond AQI alone",
    x = "AQI unhealthy days",
    y = "FEMA risk score",
    color = "Cluster"
  ) +
  theme_minimal(base_size = 12)
```

## Final Read

- The canonical Phase 3 build now lives in `R/run_phase3_livability.R`.
- The chosen published Livability typology keeps all `26` KPIs and uses `k = 6`.
- The `k = 6` choice is a moderate refinement over `k = 5`, not a wholesale rewrite of the metro map.
- GMM soft memberships are available for ambiguity-sensitive analysis.
- The main remaining dependency outside Phase 3 is the Livability/Opportunity scatter, which waits on Phase 4 scores.
