---
title: "Phase 4 - Opportunity Frame Review"
subtitle: "Review notebook for the modular Phase 4 Opportunity build"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 2
    code-fold: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)

options(
  dplyr.summarise.inform = FALSE,
  knitr.kable.NA = ""
)

library(here)
library(dplyr)
library(readr)
library(arrow)
library(ggplot2)
library(scales)
library(knitr)

output_dir <- here("exploration/intelligence_framework/phase_4_opportunity_calibration", "outputs")

comparison_summary <- readr::read_csv(file.path(output_dir, "opportunity_phase4_k5_k6_comparison_summary.csv"), show_col_types = FALSE)
comparison_sizes <- readr::read_csv(file.path(output_dir, "opportunity_phase4_k5_k6_cluster_sizes.csv"), show_col_types = FALSE)
cluster_centroids <- readr::read_csv(file.path(output_dir, "opportunity_phase4_cluster_centroids.csv"), show_col_types = FALSE)
cluster_representatives <- readr::read_csv(file.path(output_dir, "opportunity_phase4_cluster_representatives.csv"), show_col_types = FALSE)
cluster_metric_extremes <- readr::read_csv(file.path(output_dir, "opportunity_phase4_cluster_metric_extremes.csv"), show_col_types = FALSE)
gmm_summary <- readr::read_csv(file.path(output_dir, "opportunity_phase4_gmm_summary.csv"), show_col_types = FALSE)
gmm_hybrids <- readr::read_csv(file.path(output_dir, "opportunity_phase4_gmm_hybrids.csv"), show_col_types = FALSE)
similarity_top10 <- readr::read_csv(file.path(output_dir, "opportunity_phase4_similarity_top10.csv"), show_col_types = FALSE)
industry_leading_indicator_summary <- readr::read_csv(file.path(output_dir, "opportunity_phase4_industry_leading_indicator_summary.csv"), show_col_types = FALSE)
industry_leading_indicator_scatter <- readr::read_csv(file.path(output_dir, "opportunity_phase4_industry_leading_indicator_scatter.csv"), show_col_types = FALSE)
social_capital_summary <- readr::read_csv(file.path(output_dir, "opportunity_phase4_social_capital_summary.csv"), show_col_types = FALSE)
social_capital_scatter <- readr::read_csv(file.path(output_dir, "opportunity_phase4_social_capital_scatter.csv"), show_col_types = FALSE)
signal_divergence_summary <- readr::read_csv(file.path(output_dir, "opportunity_phase4_signal_divergence_summary.csv"), show_col_types = FALSE)
signal_divergence_scatter <- readr::read_csv(file.path(output_dir, "opportunity_phase4_signal_divergence_scatter.csv"), show_col_types = FALSE)
livability_opportunity_summary <- readr::read_csv(file.path(output_dir, "opportunity_phase4_livability_opportunity_summary.csv"), show_col_types = FALSE)
livability_opportunity_scatter <- readr::read_csv(file.path(output_dir, "opportunity_phase4_livability_opportunity_scatter.csv"), show_col_types = FALSE)
oz_high_opportunity_context <- readr::read_csv(file.path(output_dir, "opportunity_phase4_oz_high_opportunity_context.csv"), show_col_types = FALSE)
opportunity_scores <- arrow::read_parquet(file.path(output_dir, "opportunity_scores.parquet"))

selected_k <- 6
selected_cluster_sizes <- comparison_sizes |>
  filter(method == "kmeans", k == selected_k) |>
  arrange(cluster)

model_summary <- tibble::tibble(
  modeling_universe = nrow(opportunity_scores),
  selected_k = selected_k,
  kmeans_silhouette_at_k6 = comparison_summary |>
    filter(k == selected_k) |>
    pull(kmeans_avg_silhouette),
  hierarchical_silhouette_at_k6 = comparison_summary |>
    filter(k == selected_k) |>
    pull(hierarchical_avg_silhouette),
  social_capital_correlation = social_capital_summary$raw_correlation,
  industry_mix_model_r2 = industry_leading_indicator_summary$adjusted_r_squared,
  market_short_long_correlation = signal_divergence_summary$market_short_long_correlation,
  resident_short_long_correlation = signal_divergence_summary$resident_short_long_correlation
)
```

## Purpose

This notebook is the review layer for Phase 4, not the build pipeline.

The canonical build now lives in:

- `R/run_phase4_opportunity.R`
- `R/phase4_*.R`

This notebook reads the saved artifacts in `outputs/` so we can:

- review the locked reduced-set `k = 6` Opportunity model
- inspect the published cluster structure and names
- sanity-check soft membership and peer outputs
- review the first Opportunity hypothesis results
- inspect the Livability / Opportunity quadrant surface

## Model Snapshot

```{r model-summary}
model_summary |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable()
```

## Why `k = 6`

The published Opportunity typology uses the reduced clustering set and locks `k = 6`. This accepts a slightly weaker `k`-means silhouette than `k = 5` in exchange for a more useful subtype split on the low-opportunity side and cleaner hierarchical structure.

```{r k-comparison}
comparison_summary |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable()
```

```{r cluster-sizes}
selected_cluster_sizes |>
  rename(opportunity_kmeans_cluster = cluster, metros_in_cluster = cluster_size) |>
  left_join(
    cluster_centroids |>
      distinct(opportunity_kmeans_cluster, opportunity_cluster_name),
    by = "opportunity_kmeans_cluster"
  ) |>
  select(opportunity_kmeans_cluster, opportunity_cluster_name, metros_in_cluster) |>
  kable()
```

## Published Cluster Profiles

```{r centroids}
cluster_centroids |>
  transmute(
    opportunity_cluster_name,
    metros_in_cluster,
    opportunity_score = round(opportunity_score, 3),
    opportunity_percentile = round(opportunity_percentile, 1),
    business_score = round(subject_score_business_and_industry_opportunity, 3),
    market_score = round(subject_score_market_opportunity, 3),
    resident_score = round(subject_score_resident_opportunity, 3)
  ) |>
  kable()
```

```{r representatives}
cluster_representatives |>
  group_by(opportunity_cluster_name) |>
  slice_head(n = 6) |>
  ungroup() |>
  mutate(
    distance_to_center = round(distance_to_center, 2),
    opportunity_percentile = round(opportunity_percentile, 1)
  ) |>
  select(opportunity_cluster_name, cbsa_name, distance_to_center, opportunity_percentile) |>
  kable()
```

## Metric Signatures

These are the strongest positive and negative clustering signals within each published cluster.

```{r metric-extremes}
cluster_metric_extremes |>
  left_join(
    cluster_centroids |>
      distinct(opportunity_kmeans_cluster, opportunity_cluster_name),
    by = "opportunity_kmeans_cluster"
  ) |>
  mutate(avg_metric_score = round(avg_metric_score, 3)) |>
  arrange(opportunity_cluster_name, extreme_direction, desc(avg_metric_score)) |>
  select(opportunity_cluster_name, metric_id, avg_metric_score, extreme_direction) |>
  kable()
```

## Soft Membership Review

```{r gmm-summary}
gmm_summary |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable()
```

```{r gmm-hybrids}
gmm_hybrids |>
  arrange(membership_gap) |>
  slice_head(n = 15) |>
  mutate(across(contains("probability"), \(x) round(x, 3))) |>
  kable()
```

## Similarity Sanity Check

```{r peers}
similarity_top10 |>
  filter(cbsa_code %in% c(12580, 27260, 41700)) |>
  mutate(cosine_similarity = round(cosine_similarity, 3)) |>
  kable()
```

## Hypothesis Review

### Industry Mix As A Leading Indicator

```{r industry-leading-summary}
industry_leading_indicator_summary |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable()
```

```{r industry-leading-plot, fig.height=5}
industry_leading_indicator_scatter |>
  ggplot(aes(
    x = lq_professional,
    y = income_pc_growth_5yr,
    color = opportunity_cluster_name
  )) +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "2015 industry mix has measurable signal for later income growth",
    x = "2015 professional-services location quotient",
    y = "2022 income per-capita growth, 5-year",
    color = "Cluster"
  ) +
  theme_minimal(base_size = 12)
```

### Social Capital And Income Growth

```{r social-capital-summary}
social_capital_summary |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable()
```

```{r social-capital-plot, fig.height=5}
social_capital_scatter |>
  ggplot(aes(
    x = economic_connectedness,
    y = income_pc_growth_5yr,
    color = opportunity_cluster_name
  )) +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "Economic connectedness is not a simple positive Opportunity signal in the first pass",
    x = "Economic connectedness",
    y = "Income per-capita growth, 5-year",
    color = "Cluster"
  ) +
  theme_minimal(base_size = 12)
```

### Short-Run Vs. Long-Run Signal Divergence

```{r signal-summary}
signal_divergence_summary |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable()
```

```{r signal-plot, fig.height=5}
signal_divergence_scatter |>
  ggplot(aes(
    x = market_signal_gap,
    y = resident_signal_gap,
    color = opportunity_cluster_name
  )) +
  geom_hline(yintercept = 0, color = "#adb5bd", linewidth = 0.4) +
  geom_vline(xintercept = 0, color = "#adb5bd", linewidth = 0.4) +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "Short-run and long-run Opportunity signals often diverge",
    x = "Market short-run minus long-run signal",
    y = "Resident short-run minus long-run signal",
    color = "Cluster"
  ) +
  theme_minimal(base_size = 12)
```

## Livability / Opportunity Four Quadrants

```{r quadrant-summary}
livability_opportunity_summary |>
  kable()
```

```{r livability-opportunity-plot, fig.height=5}
livability_opportunity_scatter |>
  ggplot(aes(
    x = livability_percentile,
    y = opportunity_percentile,
    color = opportunity_cluster_name
  )) +
  geom_hline(yintercept = 50, color = "#adb5bd", linewidth = 0.4) +
  geom_vline(xintercept = 50, color = "#adb5bd", linewidth = 0.4) +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "Livability and Opportunity create a balanced four-quadrant split",
    x = "Livability percentile",
    y = "Opportunity percentile",
    color = "Opportunity cluster"
  ) +
  theme_minimal(base_size = 12)
```

## OZ Overlay Note

`OZ` refers to `Opportunity Zones`, the federal tax-incentive designation for selected lower-income census tracts. The CBSA rollup is now live, so we can use both tract-share and population-share OZ exposure as contextual overlays on the Opportunity map.

```{r oz-context}
oz_high_opportunity_context |>
  transmute(
    cbsa_name,
    opportunity_cluster_name,
    opportunity_percentile = round(opportunity_percentile, 1),
    pct_oz_tracts = scales::percent(pct_oz_tracts, accuracy = 0.1),
    pct_population_in_oz = scales::percent(pct_population_in_oz, accuracy = 0.1),
    oz_tract_count,
    total_tract_count
  ) |>
  slice_head(n = 15) |>
  kable()
```

## Final Read

- The canonical Phase 4 build now lives in `R/run_phase4_opportunity.R`.
- The published Opportunity typology uses the reduced clustering set and locks `k = 6`.
- The six-cluster structure is coherent, with the added value of a distinct distressed / thin-base subtype.
- Soft memberships and top-10 cosine peers are now available for ambiguity-sensitive review.
- The first social-capital pass is cautionary rather than confirmatory, while the 1-year vs. 5-year divergence test looks substantively useful.
- The Livability / Opportunity quadrant surface is now complete and ready for cross-frame follow-on work.
- The OZ overlay is now available as metro-level context through both `pct_oz_tracts` and `pct_population_in_oz`.
