---
title: "Phase 6 - Trajectory Overview"
subtitle: "Cross-frame trajectory review over built Phase 6 artifacts"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 2
    code-fold: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

options(
  dplyr.summarise.inform = FALSE,
  knitr.kable.NA = ""
)

library(here)
library(dplyr)
library(readr)
library(arrow)
library(ggplot2)
library(knitr)
library(tidyr)
library(scales)

output_dir <- here("exploration/intelligence_framework/phase_6_trajectory", "outputs")

trajectory_scores <- read_parquet(file.path(output_dir, "trajectory_scores.parquet")) |> as_tibble()
kpi_trajectory_long <- read_csv(file.path(output_dir, "phase6_kpi_trajectory_long.csv"), show_col_types = FALSE)
pattern_summary <- read_csv(file.path(output_dir, "phase6_pattern_summary.csv"), show_col_types = FALSE)
candidate_list <- read_csv(file.path(output_dir, "phase6_candidate_list.csv"), show_col_types = FALSE)

direction_counts <- bind_rows(
  trajectory_scores |> count(direction = character_direction, name = "metros") |> mutate(frame = "Character"),
  trajectory_scores |> count(direction = livability_direction, name = "metros") |> mutate(frame = "Livability"),
  trajectory_scores |> count(direction = opportunity_direction, name = "metros") |> mutate(frame = "Opportunity")
)

strength_long <- trajectory_scores |>
  transmute(
    cbsa_name,
    character = character_trajectory_strength,
    livability = livability_trajectory_strength,
    opportunity = opportunity_trajectory_strength
  ) |>
  pivot_longer(-cbsa_name, names_to = "frame", values_to = "trajectory_strength")
```

## Purpose

This notebook is the top-level review layer for Phase 6. It reads only the saved outputs from `phase_6_trajectory/outputs/`.

The current default analysis choices are:

- pattern thresholds use top-decile national ranks
- candidate scoring uses equal pattern weights with a smaller Phase 5 overlap bonus

If those defaults change, the outputs will become either more selective or more inclusive. The sensitivity notes written into the Phase 6 CSVs call that out explicitly.

## Model Snapshot

```{r}
tibble(
  cbsa_universe = nrow(trajectory_scores),
  trajectory_rows = nrow(kpi_trajectory_long),
  candidate_rows = nrow(candidate_list),
  patterns_with_any_members = sum(pattern_summary$cbsa_count > 0)
) |>
  kable()
```

## Direction Segmentation

```{r}
direction_counts |>
  kable()
```

```{r fig.height=5}
ggplot(direction_counts, aes(x = frame, y = metros, fill = direction)) +
  geom_col(position = "stack") +
  labs(
    title = "Each frame splits metros into the same four direction buckets",
    x = NULL,
    y = "CBSAs",
    fill = "Direction"
  ) +
  theme_minimal(base_size = 12)
```

## Strength Distribution

```{r fig.height=5}
ggplot(strength_long, aes(x = trajectory_strength, fill = frame)) +
  geom_histogram(alpha = 0.7, bins = 30, position = "identity") +
  facet_wrap(~frame, scales = "free_y") +
  labs(
    title = "Trajectory strength distributions by frame",
    x = "Trajectory strength",
    y = "CBSAs"
  ) +
  theme_minimal(base_size = 12) +
  guides(fill = "none")
```

## Top Trajectory Movers

```{r}
trajectory_scores |>
  transmute(
    cbsa_name,
    census_division,
    character_trajectory_strength,
    livability_trajectory_strength,
    opportunity_trajectory_strength,
    candidate_rank,
    candidate_score
  ) |>
  arrange(candidate_rank) |>
  slice_head(n = 20) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable()
```

## Pattern Summary

```{r}
pattern_summary |>
  kable()
```

## What Changed If We Tighten Or Loosen The Defaults

Look here for review:

- whether the top-decile default is producing too few members for a pattern, especially `bounce_back` and `environmental_risk_outlier`
- whether the candidate top ranks look dominated by one narrative even under equal pattern weights
- whether the high-ranked metros feel like real trajectory stories rather than just already-famous large metros
