---
title: "Phase 6 - Bounce Back Markets"
subtitle: "Converging-from-below Opportunity recovery review"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 2
    code-fold: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

options(
  dplyr.summarise.inform = FALSE,
  knitr.kable.NA = ""
)

library(here)
library(dplyr)
library(readr)
library(arrow)
library(ggplot2)
library(knitr)

output_dir <- here("exploration/intelligence_framework/phase_6_trajectory", "outputs")

trajectory_scores <- read_parquet(file.path(output_dir, "trajectory_scores.parquet")) |> as_tibble()
pattern_summary <- read_csv(file.path(output_dir, "phase6_pattern_summary.csv"), show_col_types = FALSE)

bounce_back <- trajectory_scores |> filter(is_bounce_back)
near_misses <- trajectory_scores |>
  filter(opportunity_direction == "converging-improving") |>
  arrange(desc(opportunity_trajectory_strength), opportunity_current_percentile, cbsa_name)
```

## Purpose

The bounce-back pattern looks for metros that are still relatively weak on current Opportunity standing but are recovering quickly.

The default filter is:

- Opportunity direction = `converging-improving`
- Opportunity trajectory strength in the top national decile
- current Opportunity percentile at or below `50`

## Current Pattern Count

```{r}
pattern_summary |>
  filter(pattern_flag == "is_bounce_back") |>
  kable()
```

## Matched Metros

```{r}
if (nrow(bounce_back) == 0) {
  knitr::kable(tibble(note = "No metros met the current bounce-back default."))
} else {
  bounce_back |>
    select(
      cbsa_name,
      census_division,
      opportunity_trajectory_strength,
      opportunity_trajectory_score,
      opportunity_direction,
      opportunity_current_percentile,
      candidate_rank
    ) |>
    mutate(across(where(is.numeric), \(x) round(x, 3))) |>
    kable()
}
```

## Near Misses

```{r}
near_misses |>
  slice_head(n = 25) |>
  select(
    cbsa_name,
    census_division,
    opportunity_trajectory_strength,
    opportunity_direction,
    opportunity_current_percentile,
    opp_turn_signal,
    candidate_rank
  ) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable()
```

## Review Lens

Look for:

- whether zero matches feels analytically correct or overly strict
- whether the near misses are the metros you would expect to anchor an Article 7 recovery story
- whether the current cutoff is screening out plausible rebound stories because their present percentile is slightly too high
