---
title: "Phase 6 - Hidden Livability Winners"
subtitle: "Improving Livability metros without top-tier cross-frame profile"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 2
    code-fold: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

options(
  dplyr.summarise.inform = FALSE,
  knitr.kable.NA = ""
)

library(here)
library(dplyr)
library(readr)
library(arrow)
library(ggplot2)
library(knitr)

output_dir <- here("exploration/intelligence_framework/phase_6_trajectory", "outputs")

trajectory_scores <- read_parquet(file.path(output_dir, "trajectory_scores.parquet")) |> as_tibble()
pattern_summary <- read_csv(file.path(output_dir, "phase6_pattern_summary.csv"), show_col_types = FALSE)

hidden_winners <- trajectory_scores |>
  filter(is_hidden_livability_winner) |>
  arrange(desc(livability_trajectory_strength), cross_frame_percentile, cbsa_name)
```

## Default Rule

The current default hidden-winner screen is:

- Livability direction = `diverging-improving`
- Livability trajectory strength in the top national decile
- Phase 5 cross-frame percentile between `35` and `70`

That middle-band cross-frame rule is meant to avoid metros that are already obvious national standouts.

## Matched Metros

```{r}
pattern_summary |>
  filter(pattern_flag == "is_hidden_livability_winner") |>
  kable()
```

```{r}
hidden_winners |>
  select(
    cbsa_name,
    census_division,
    livability_trajectory_strength,
    livability_trajectory_score,
    livability_current_percentile,
    cross_frame_percentile,
    candidate_rank
  ) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable()
```

## Position Versus Momentum

```{r fig.height=5}
ggplot(
  trajectory_scores,
  aes(
    x = cross_frame_percentile,
    y = livability_trajectory_strength,
    color = is_hidden_livability_winner
  )
) +
  geom_point(alpha = 0.7, size = 2) +
  geom_vline(xintercept = c(35, 70), linetype = "dashed") +
  geom_hline(yintercept = quantile(trajectory_scores$livability_trajectory_strength, 0.9, na.rm = TRUE), linetype = "dashed") +
  labs(
    title = "Hidden winners sit in the middle of the cross-frame pack but high on Livability movement",
    x = "Phase 5 cross-frame percentile",
    y = "Livability trajectory strength",
    color = "Hidden winner"
  ) +
  theme_minimal(base_size = 12)
```

## Review Lens

Look for:

- whether these metros really feel “under-discussed” rather than just expensive superstar markets
- whether the `35` to `70` cross-frame band is too wide or too narrow
- whether current Livability percentile should matter more than it does now
