---
title: "Phase 6 - Diverging From Themselves"
subtitle: "Cross-frame trajectory disagreement with Phase 5 overlap context"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 2
    code-fold: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

options(
  dplyr.summarise.inform = FALSE,
  knitr.kable.NA = ""
)

library(here)
library(dplyr)
library(readr)
library(arrow)
library(ggplot2)
library(knitr)

output_dir <- here("exploration/intelligence_framework/phase_6_trajectory", "outputs")

trajectory_scores <- read_parquet(file.path(output_dir, "trajectory_scores.parquet")) |> as_tibble()
candidate_list <- read_csv(file.path(output_dir, "phase6_candidate_list.csv"), show_col_types = FALSE)

diverging <- trajectory_scores |>
  filter(is_diverging_from_themselves) |>
  arrange(desc(trajectory_disagreement_sd), candidate_rank, cbsa_name)
```

## Why This Pattern Matters

This is the most direct Phase 6 continuation of the Phase 5 overlap logic.

The current default requires:

- at least one improving frame
- at least one declining frame
- trajectory disagreement in the top national decile

## Matched Metros

```{r}
diverging |>
  select(
    cbsa_name,
    census_division,
    character_direction,
    livability_direction,
    opportunity_direction,
    trajectory_disagreement_sd,
    phase5_overlap_rank,
    candidate_rank
  ) |>
  slice_head(n = 30) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable()
```

## Phase 5 Enrichment

```{r}
candidate_list |>
  filter(is_diverging_from_themselves) |>
  count(overlap_profile, half_alignment, sort = TRUE) |>
  kable()
```

## Trajectory Conflict Surface

```{r fig.height=5}
ggplot(
  trajectory_scores,
  aes(
    x = character_trajectory_score,
    y = opportunity_trajectory_score,
    color = is_diverging_from_themselves
  )
) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "Character versus Opportunity trajectory disagreement",
    x = "Character trajectory score",
    y = "Opportunity trajectory score",
    color = "Diverging"
  ) +
  theme_minimal(base_size = 12)
```

## Review Lens

Look for:

- whether the high-disagreement metros read like true multi-story markets
- whether Phase 5 overlap rank is reinforcing the right edge cases rather than simply boosting already famous metros
- whether Livability disagreement, rather than Character/Opportunity, is actually where the richest stories sit
