---
title: "Phase 6 - Fast Demographic Changers"
subtitle: "Character 5-year fast movers"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 2
    code-fold: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

options(
  dplyr.summarise.inform = FALSE,
  knitr.kable.NA = ""
)

library(here)
library(dplyr)
library(readr)
library(arrow)
library(ggplot2)
library(knitr)

output_dir <- here("exploration/intelligence_framework/phase_6_trajectory", "outputs")

trajectory_scores <- read_parquet(file.path(output_dir, "trajectory_scores.parquet")) |> as_tibble()
kpi_trajectory_long <- read_csv(file.path(output_dir, "phase6_kpi_trajectory_long.csv"), show_col_types = FALSE)

fast_changers <- trajectory_scores |>
  filter(is_fast_demographic_changer) |>
  arrange(desc(character_trajectory_strength), cbsa_name)

character_metric_leaders <- kpi_trajectory_long |>
  filter(frame_id == "character", window_years == 5) |>
  group_by(metric_id) |>
  summarise(
    mean_strength = mean(metric_trajectory_strength, na.rm = TRUE),
    top_examples = paste(head(cbsa_name[order(-metric_trajectory_strength)], 5), collapse = " | "),
    .groups = "drop"
  ) |>
  arrange(desc(mean_strength))
```

## Fast-Mover List

```{r}
fast_changers |>
  select(
    cbsa_name,
    census_division,
    character_trajectory_strength,
    character_trajectory_score,
    character_direction,
    candidate_rank
  ) |>
  slice_head(n = 30) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable()
```

## Which Character Metrics Are Carrying Movement

```{r}
character_metric_leaders |>
  mutate(mean_strength = round(mean_strength, 3)) |>
  kable()
```

## Character Trajectory Strength Distribution

```{r fig.height=5}
ggplot(
  trajectory_scores,
  aes(x = character_trajectory_strength, fill = is_fast_demographic_changer)
) +
  geom_histogram(bins = 30, alpha = 0.8) +
  labs(
    title = "Character fast movers are the top-decile tail of Character trajectory strength",
    x = "Character trajectory strength",
    y = "CBSAs",
    fill = "Fast changer"
  ) +
  theme_minimal(base_size = 12)
```

## Review Lens

Look for:

- whether the top markets feel like genuine demographic-change stories
- whether migration-heavy metrics are dominating too much now that 4 social-fabric metrics are context-only
- whether the candidate metros would still look compelling if civic/social-capital annual series were eventually backfilled
