---
title: "Phase 6 - Environmental Risk Outliers"
subtitle: "Dual-axis environmental worsening review"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 2
    code-fold: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

options(
  dplyr.summarise.inform = FALSE,
  knitr.kable.NA = ""
)

library(here)
library(dplyr)
library(readr)
library(arrow)
library(ggplot2)
library(knitr)

output_dir <- here("exploration/intelligence_framework/phase_6_trajectory", "outputs")

trajectory_scores <- read_parquet(file.path(output_dir, "trajectory_scores.parquet")) |> as_tibble()
pattern_summary <- read_csv(file.path(output_dir, "phase6_pattern_summary.csv"), show_col_types = FALSE)

environmental_near_misses <- trajectory_scores |>
  transmute(
    cbsa_name,
    census_division,
    livability_direction,
    worsening_rank_pct_aqi_unhealthy_days,
    worsening_rank_pct_fema_risk_score,
    candidate_rank
  ) |>
  arrange(desc(worsening_rank_pct_aqi_unhealthy_days + worsening_rank_pct_fema_risk_score), cbsa_name)
```

## Current Pattern Count

```{r}
pattern_summary |>
  filter(pattern_flag == "is_environmental_risk_outlier") |>
  kable()
```

## Near Misses

```{r}
environmental_near_misses |>
  slice_head(n = 25) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable()
```

## Dual-Axis Worsening Surface

```{r fig.height=5}
ggplot(
  trajectory_scores,
  aes(
    x = worsening_rank_pct_aqi_unhealthy_days,
    y = worsening_rank_pct_fema_risk_score,
    color = is_environmental_risk_outlier
  )
) +
  geom_vline(xintercept = 0.9, linetype = "dashed") +
  geom_hline(yintercept = 0.9, linetype = "dashed") +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "Current default requires both environmental worsening ranks to hit the top decile",
    x = "AQI worsening percentile",
    y = "FEMA risk worsening percentile",
    color = "Pattern match"
  ) +
  theme_minimal(base_size = 12)
```

## Review Lens

Look for:

- whether zero matches feels too strict for a first-pass environmental warning screen
- whether the near misses would become more useful under a slightly looser threshold
- whether the Livability direction requirement is screening out metros with genuine environmental deterioration but mixed non-environment Livability movement
