---
title: "Phase 6 - Candidate List Review"
subtitle: "Ranked Deep Dive candidate surface with Jacksonville and Richmond check"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 2
    code-fold: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

options(
  dplyr.summarise.inform = FALSE,
  knitr.kable.NA = ""
)

library(here)
library(dplyr)
library(readr)
library(ggplot2)
library(knitr)

output_dir <- here("exploration/intelligence_framework/phase_6_trajectory", "outputs")

candidate_list <- read_csv(file.path(output_dir, "phase6_candidate_list.csv"), show_col_types = FALSE)

jax_rva_check <- candidate_list |>
  filter(cbsa_name %in% c("Jacksonville, FL", "Richmond, VA"))
```

## Candidate Method

The current candidate score uses:

- equal weights across the five Phase 6 pattern flags
- a multiplier from overall trajectory strength
- a smaller Phase 5 overlap bonus

The sensitivity note in the CSV explains how changing those weights would move the rankings.

## Top 50 Candidates

```{r}
candidate_list |>
  slice_head(n = 50) |>
  select(
    candidate_rank,
    cbsa_name,
    census_division,
    pattern_count,
    pattern_signal_score,
    overall_trajectory_strength,
    phase5_overlap_rank,
    candidate_score
  ) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable()
```

## Jacksonville And Richmond Check

```{r}
jax_rva_check |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable()
```

## Rank Composition

```{r fig.height=5}
ggplot(
  candidate_list,
  aes(
    x = overall_trajectory_strength,
    y = candidate_score,
    color = factor(pattern_count)
  )
) +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "Candidate score rises with both pattern count and overall trajectory strength",
    x = "Overall trajectory strength",
    y = "Candidate score",
    color = "Pattern count"
  ) +
  theme_minimal(base_size = 12)
```

## Review Lens

Look for:

- whether the top 20 feel like true review-priority metros rather than arbitrary score artifacts
- whether Jacksonville and Richmond are positioned in ways that make sense for the first Deep Dive decision
- whether the overlap bonus is helping the right cases or over-rewarding Phase 5 contradiction alone
