---
title: "Phase 6 - Opportunity Turn Signals"
subtitle: "Short-run versus medium-term Opportunity divergence"
date: today
format:
  html:
    self-contained: true
    toc: true
    toc-depth: 2
    code-fold: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

options(
  dplyr.summarise.inform = FALSE,
  knitr.kable.NA = ""
)

library(here)
library(dplyr)
library(readr)
library(ggplot2)
library(knitr)

output_dir <- here("exploration/intelligence_framework/phase_6_trajectory", "outputs")

opp_turn_signals <- read_csv(file.path(output_dir, "phase6_opp_turn_signals.csv"), show_col_types = FALSE)
candidate_list <- read_csv(file.path(output_dir, "phase6_candidate_list.csv"), show_col_types = FALSE)
```

## Turn-Signal Summary

```{r}
tibble(
  cbsa_universe = nrow(opp_turn_signals),
  turn_signal_count = sum(opp_turn_signals$opp_turn_signal %in% TRUE, na.rm = TRUE)
) |>
  kable()
```

## Strongest Contradictions

```{r}
opp_turn_signals |>
  filter(opp_turn_signal %in% TRUE) |>
  mutate(
    momentum_gap = abs(frame_momentum_z_window_1yr - frame_momentum_z_window_5yr)
  ) |>
  arrange(desc(momentum_gap), cbsa_name) |>
  slice_head(n = 30) |>
  select(
    cbsa_name,
    census_division,
    frame_momentum_z_window_1yr,
    frame_momentum_z_window_5yr,
    frame_direction_window_1yr,
    frame_direction_window_5yr,
    opp_turn_signal_type
  ) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable()
```

## 1-Year Versus 5-Year Matrix

```{r fig.height=5}
ggplot(
  opp_turn_signals,
  aes(
    x = frame_momentum_z_window_5yr,
    y = frame_momentum_z_window_1yr,
    color = opp_turn_signal
  )
) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "Opportunity turn signals are metros where 1-year and 5-year momentum disagree",
    x = "5-year Opportunity momentum z-score",
    y = "1-year Opportunity momentum z-score",
    color = "Turn signal"
  ) +
  theme_minimal(base_size = 12)
```

## Candidate List Cross-Check

```{r}
candidate_list |>
  filter(opp_turn_signal %in% TRUE) |>
  slice_head(n = 25) |>
  select(
    candidate_rank,
    cbsa_name,
    opp_turn_signal_type,
    opportunity_trajectory_score,
    candidate_score
  ) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  kable()
```

## Review Lens

Look for:

- whether the turn-signal metros feel like real “market is turning” stories instead of noisy annual reversals
- whether CT metros with reduced 5-year metric coverage need to be interpreted more cautiously here
- whether the highest-ranked turn-signal metros belong in a Phase 6 story even when they do not score highly on the full candidate list
